\chapter{Shadow}

\noindent Sa-Kailan returned to Light with dust still hidden in his sandals.\par
He had cleaned them. He had even scraped the seams with a bone pin until one strap began to fray. It had not been enough. Each step toward the eastern practice halls gave back a faint dry rasp, and the sound irritated him because it seemed to report where he had been. The Oasis had followed him under stone. So had the sky.\par
So had the Book of Blood, and Va-Sheva’s voice at the marker line.\par
Two small wrongnesses at a threshold are not treated separately until Shield can afford the luxury.\par
He had slept poorly again. The basin beside his mat had been still, but his mind had not. It returned to the token under water, to the sign of Light shifting without moving, to the desert flash and the turned marker, to Va-Sheva not answering whether she had gone beyond the second ridge.\par
The outside had multiplied the number of things requiring answer.\par
Va-Raedin was already waiting at the first eastern passage.\par
He looked once at Sa-Kailan’s sandals, once at the hem of his robe, once at the bandage around his thumb.\par
“You went outside,” he said.\par
“Yes, Veil.”\par
“And returned.”\par
“Yes.”\par
“That is the usual aim of controlled prayer access.”\par
Sa-Kailan kept his face still.\par
Va-Raedin turned and began walking before asking the real question.\par
“At the Oasis, did the light behave?”\par
Sa-Kailan followed him.\par
“No.”\par
“Good. Then you saw something.”\par
“Outside light is too full.”\par
“Too full?”\par
“It comes from too many places. It strikes everything at once. Inside, light is disciplined. Outside, it hides by abundance.”\par
Va-Raedin glanced back, almost approving.\par
“And inside?”\par
“Inside it hides by scarcity.”\par
“Better.”\par
The word entered him too quickly. He tried not to look pleased and failed by enough for Va-Raedin to notice.\par
“Come,” the Veil said. “Today we begin again.”\par
They did not enter the first practice hall. Va-Raedin led him beyond the mirror rooms, past the chamber where the red point had trembled on the wall, and into a low service slit beside a lamp niche. The passage descended. The lamps were set near the floor, so that their light kept the feet visible while the upper body passed through darkness. It made Sa-Kailan feel divided: obedient below, unwitnessed above.\par
At the end stood a plain door of pale wood.\par
No sign of Light. No inscription. No aperture. No visible lock.\par
Va-Raedin stopped.\par
“What do you see?”\par
“A wooden door in Light territory.”\par
“Better than yesterday. Continue.”\par
“It does not belong. Light uses stone, metal, glass, mirror. Wood swells. Warps. Burns. Remembers damp.”\par
“Therefore?”\par
“Either it is unimportant, or its changing matters.”\par
Va-Raedin opened it.\par
Beyond was a room almost entirely dark. One lamp burned outside the room, behind them, high in the passage wall. Its light entered through a narrow slit in the door and cut one hard line across the floor.\par
Va-Raedin closed the door.\par
The line remained.\par
“Five shadows,” he said.\par
Sa-Kailan looked down.\par
“There is only one.”\par
“That is why you are Candidate.”\par
The answer irritated him exactly as intended.\par
Va-Raedin moved to the first covered object but did not uncover it.\par
“Yesterday you learned that light can embarrass you. Today you learn the principal forms by which it does so. Name the first.”\par
“Mirrors.”\par
“That is the object. Name the danger.”\par
“Reversal.”\par
“A child’s answer.”\par
“Vanity.”\par
“A moralist’s answer.”\par
Sa-Kailan looked at the covered mirror, the slit of light, the door behind him. A mirror did not merely reverse. It placed a sight somewhere other than where the body stood. It allowed the eye to travel without the man.\par
“Displacement,” he said. “Sight away from the body.”\par
Va-Raedin removed the cloth.\par
The mirror beneath was tall and narrow, angled toward the far wall. Sa-Kailan expected to see himself. Instead he saw the door behind him, the slit of light, the passage outside, and beyond that a figure standing in a place where no figure stood.\par
He turned before thinking.\par
The passage was empty.\par
When he looked back into the mirror, the figure remained.\par
“Who is it?” Va-Raedin asked.\par
“No one.”\par
“Then why did you turn?”\par
“Because I believed the sight before I questioned the path.”\par
“Good.”\par
Va-Raedin adjusted the mirror by less than a finger’s width. The figure vanished. In its place appeared the back of Sa-Kailan’s own head.\par
“A mirror can show what is behind the eye,” Va-Raedin said. “That is its use and its corruption. Most men think sight travels forward because their faces do. Light has no such loyalty.”\par
He placed a small bronze token behind Sa-Kailan, outside direct sight.\par
“Find it without turning.”\par
Sa-Kailan used the tall mirror first. The token appeared near the reflected door, but distance was uncertain. He shifted left; it vanished behind his reflected shoulder. He shifted right; it seemed to move farther away. A smaller polished plate near the water trough caught the floor in a second reflection, distorted but useful. He combined both views, stepped back, touched the ground with his heel, and heard the soft click of metal under sandal.\par
He bent and took the token without turning.\par
Va-Raedin accepted it back.\par
“First shadow?”\par
“Reflection displaces sight. It lets the eye stand where the body is not.”\par
“And the danger?”\par
“The sight may be true and still not be where I think it is.”\par
“That will stand.”\par
The word wounded less now. That made it more dangerous.\par
Va-Raedin covered the mirror.\par
“Second shadow.”\par
“Lenses.”\par
“Object.”\par
“False importance.”\par
This time Va-Raedin did not correct him immediately.\par
“Explain.”\par
“A lens makes one thing larger and another thing disappear. If a man does not know the lens is there, he mistakes focus for importance.”\par
“Better.”\par
The second station held a glass disk suspended from three bronze cords. Behind it, on the wall, were five dark marks no larger than seeds. A side lamp threw light through the disk so that one seed-mark swelled into a dark patch the size of Sa-Kailan’s palm. The others almost disappeared.\par
“Which mark matters?” Va-Raedin asked.\par
“Closest to what?”\par
“To judgment.”\par
“That is not a place.”\par
“It is always a place. You only learn its position too late.”\par
Sa-Kailan moved around the disk. The large mark blurred when he stepped aside. Another, nearly invisible from the first position, sharpened. Lowering his eye made the second mark swell while the first became a smear. Each mark could be made dominant if the glass, lamp, and observer agreed to make it so.\par
“None matters by itself,” he said. “The lens grants importance.”\par
“The hand that places the lens grants it. The lens obeys its shape.”\par
Va-Raedin tilted the disk. The enlarged mark stretched into a long black wound.\par
“What is it now?”\par
“The same mark.”\par
“Would a crowd know that?”\par
“No.”\par
“Would Faith?”\par
“Faith might make it speak.”\par
“Balance?”\par
“Measure it.”\par
“Shield?”\par
“Keep people from approaching until it knows whether the mark matters.”\par
“Justice?”\par
Sa-Kailan hesitated.\par
“Ask who placed the lens.”\par
“And Light?”\par
Sa-Kailan looked at the wound-shaped mark.\par
“Light must know how easily the seed became a wound.”\par
“Good.”\par
They worked with the lenses until Sa-Kailan’s eyes hurt. Glass disks. Water-filled bulbs. Oil jars. Thin plates that widened light, narrowed it, blurred it, or sharpened one edge while ruining another. Va-Raedin made him stand close, far, above, below, with one eye closed, then viewing by reflection first. Each success made the room less trustworthy.\par
By the end, size, clarity, and nearness had become suspect.\par
“Second shadow?”\par
“Focus creates false importance.”\par
“And the danger?”\par
“Seeing sharply can be another form of blindness.”\par
Va-Raedin’s face did not move.\par
“Careful. That one is almost useful.”\par
“You enjoy dangerous sentences.”\par
“I thought it was precise.”\par
“That is the usual excuse.”\par
He lowered one lamp and raised another.\par
The room changed.\par
Objects that had been ordinary became monstrous. The trough became a black wall across the floor. The lens cords became bars. The covered mirror cast a wedge that reached the far stone.\par
“Third shadow,” Va-Raedin said.\par
“Shadows.”\par
“Object.”\par
“Interruption.”\par
“Danger?”\par
“Assuming darkness is empty.”\par
“Find what is hidden.”\par
Sa-Kailan looked at the floor first and disliked himself for the predictability. The largest darkness drew his eye, so he denied it. He looked instead at the edges. Shadow was not merely absence; it was the record of something entering the path. A hidden object might cast a shadow. It might also hide inside another object’s shadow. Or reveal itself by preventing a shadow from falling where it should.\par
The rods cast five dark lines. The hanging lens cast an oval. The trough cast a heavy black band. He knelt beside it. Nothing. He stood, looked at the wall, then back to the floor. There — a faint brightness where darkness should have climbed.\par
Something low was blocking the floor-light but not the wall-light.\par
He reached into the trough’s shadow and found a black cord hanging from a hidden hook.\par
“What is it?” Va-Raedin asked.\par
“A cord.”\par
“What does it do?”\par
Sa-Kailan pulled it before judgment arrived.\par
A panel opened behind him with a hard click.\par
He froze.\par
The panel revealed only stone beyond it. A false consequence. But Va-Raedin’s silence made it clear that the real consequence had already happened.\par
The Veil crossed the room, closed the panel, and returned.\par
“Shadow hid the cord,” he said. “Pride pulled it.”\par
Sa-Kailan lowered his eyes.\par
“If that had opened a gate?”\par
“I should have waited.”\par
“If that had released water?”\par
“I should have waited.”\par
“If that had alarmed Shield?”\par
“I should have waited.”\par
“You returned from the Oasis with Shield in your mind. Keep it there. Seeing is not permission. Finding is not permission. A Candidate learns where light falls. He does not become master of everything light touches.”\par
Sa-Kailan thought of the old woman at the spring, the leaf in her hand, Va-Sheva saying that meaning no harm was not the same as remaining harmless.\par
“I understand.”\par
“No,” Va-Raedin said. “You agree. Understanding is not required.”\par
The phrase struck him at once. Shield at the cedar gate. Shield in Va-Raedin’s mouth. Either someone had told him, or the Mountain’s offices had been saying the same things to boys for centuries. Sa-Kailan no longer trusted such coincidences.\par
“Third shadow?”\par
“Darkness is evidence of interruption.”\par
“And the danger?”\par
“That discovery tempts the hand before judgment catches up.”\par
“Better.”\par
Va-Raedin permitted water then. A small cup, plain clay. Sa-Kailan drank too quickly, then found himself watching the rim instead of the room.\par
“You have been staring at water since yesterday,” Va-Raedin said.\par
“It does strange things.”\par
“Everything does strange things when attention arrives late.”\par
“I dropped the token in my basin.”\par
“I assumed you would.”\par
“You knew?”\par
“I knew you would either sleep, admire it, or misuse it. Sleep was unlikely. Admiration would bore you. Misuse teaches.”\par
Sa-Kailan felt exposed and robbed at once. The discovery had felt private. Light, apparently, had prepared even that humiliation in advance.\par
“I saw the sign shift through the water,” he said.\par
“The sign did not shift.”\par
“I know. The light did.”\par
Va-Raedin’s gaze sharpened.\par
Not praise. Not alarm. Attention.\par
“Fourth shadow,” he said.\par
The fourth room was bright, white-walled, and severe. A fixed lamp cast a beam downward through a bronze slit onto a table cut with narrow channels. At the far end of the table lay a small black mark beneath glass. The beam fell too far left to touch it.\par
“Strike the mark with light.”\par
Sa-Kailan inspected the apparatus. The lamp did not move. The slit did not move. No mirror, no plate, no hidden aperture presented itself. On the table were clear water, oil, crystal, a cloudy cup, a sealed clay pot, and several shallow recesses.\par
“May I move the vessels?”\par
“Yes.”\par
“May I open the pot?”\par
“No.”\par
He began with clear water. The beam passed through and turned, but not enough. Oil thickened it and blurred its edge. Crystal turned it more sharply, but broke the light where he needed one clean point. The cloudy liquid scattered too much at first, making the beam visible inside the cup and weak beyond it.\par
“Name the event.”\par
“The beam changes path when it crosses a substance.”\par
“Which substance?”\par
“Water, oil, glass, whatever the cloudy liquid is.”\par
“And air?”\par
He almost dismissed air, then remembered the desert trembling in heat, mist shining behind Faith’s rehearsal, the blue flame in the Hall bending toward Va-Elun.\par
“Air also,” he said.\par
“Usually less. Not never.”\par
Sa-Kailan returned to the table. No single vessel struck the mark cleanly. Water and crystal broke the beam. Oil before water made it wander. Water before oil came closer but ruined the edge. He wanted the clay pot.\par
Va-Raedin watched him want it.\par
“What is in the pot?” Sa-Kailan asked.\par
“That is what you are asking permission to learn.”\par
“Salt.”\par
Va-Raedin said nothing.\par
Balance green wax sealed the lid. A marked pot on a table of water and turning light. Salt changed water. He lifted the cloudy cup, smelled it, and found the mineral bitterness.\par
“Salted water.”\par
Still no confirmation.\par
He placed the cloudy cup beneath the beam. The light turned more than in clear water, less cleanly than through crystal. It struck near the mark. He adjusted the cup by a finger’s width. The beam touched the black point.\par
“Acceptable.”\par
“That was the answer.”\par
“That was the prepared answer. Again, without the salted cup.”\par
Sa-Kailan looked at the table, then at the sealed pot.\par
“May I open it now?”\par
“No.”\par
He looked again. Clear water. Oil. Crystal. Salted cup removed. Sealed pot. Fixed lamp. Black mark. Channels in the table. He had thought the channels were drainage. But drainage was a story one told before knowing what had been drained.\par
He poured clear water into a recess, then scraped a little white grain from the table’s edge with his fingernail.\par
“That is theft from Balance,” Va-Raedin said.\par
“It was already spilled.”\par
“It was already unmeasured.”\par
Sa-Kailan stopped.\par
Then Va-Raedin said, “Continue.”\par
He stirred the grains into the water. Nothing changed visibly. He placed the recess under the beam. The light turned slightly more. Not enough. Another scrape. More stirring. Another test. Each grain became morally expensive. At last the beam struck the mark cleanly.\par
“How much salt?” Va-Raedin asked.\par
“Enough.”\par
“A cook’s answer.”\par
“I do not know.”\par
“A dangerous answer, but honest.”\par
Sa-Kailan looked at the water.\par
“It looks the same.”\par
“Many changed things do.”\par
“And the light knows.”\par
“If one knows where to look.”\par
Water could be altered without announcing alteration. Light could reveal it. Or use it. Salt could change the path without moving lamp, slit, vessel, or mark.\par
“Fourth shadow?”\par
“Turning. Light changes path when it crosses substance.”\par
“And the danger?”\par
“The condition may change invisibly.”\par
Va-Raedin took the bowl away before Sa-Kailan could study it further.\par
“Fifth shadow.”\par
The last chamber held a long trough of clear water set into black stone. A narrow tube beneath the trough carried light from a hidden source into the water itself. At one end of the trough a small white mark had been painted below the waterline. At the other, a bronze plate stood dry outside the trough.\par
The beam travelled underwater toward the surface at an angle and vanished.\par
Or seemed to.\par
“Make the beam strike the bronze plate.”\par
Sa-Kailan leaned over the trough. The beam approached the surface from below, shallow against the water’s skin. If it escaped, it should strike the air above the trough and reach the plate. It did not.\par
He passed his hand above the surface. No light touched him.\par
He crouched lower.\par
At the surface, the beam turned downward and remained inside the water.\par
The water had become a mirror.\par
“It cannot leave,” he said.\par
“What cannot?”\par
“The light. No — the beam. When it reaches the surface at that angle, it does not pass into air. It returns into the water.”\par
“Like what?”\par
“A mirror.”\par
“Water is not a mirror.”\par
“It is here.”\par
“Why?”\par
Sa-Kailan looked at the angle, then at the surface. The water itself had not changed. The line had.\par
“If the beam rises steeply, it leaves. If it comes too shallow, the surface holds it.”\par
“Threshold return,” Va-Raedin said. “Every crossing has a condition. Some thresholds forgive. Some refuse. A door is not the only thing that may open or close.”\par
The phrase sounded old. Not invented for the room. Retrieved from deeper in Light.\par
“Change it,” Va-Raedin said.\par
“How?”\par
“Make it leave.”\par
Sa-Kailan searched for controls. The hidden source did not move. The tube was fixed. The trough was fixed. The beam’s angle was fixed. A lever raised the water level by the width of a fingernail, but the beam returned again. He lowered it. No use. He disturbed the surface with a rod. The beam scattered, then resumed.\par
“Do not bruise water because you dislike its answer.”\par
He stepped back.\par
Change the condition.\par
“Salt,” he said.\par
This time there was a small open dish of coarse crystals. He took a pinch and let it fall into the trough. The grains caught light as they descended. Nothing changed. Then the beam trembled. He stirred carefully.\par
The light shifted — but downward, held more strongly inside the water.\par
“Worse.”\par
“Worse for what?”\par
“For reaching the plate.”\par
“For leaving the water, yes.”\par
The distinction opened the problem. Salt had not made water better or worse. It had changed the threshold. More salt held the beam in. Less might release it.\par
He added clear water from a side vessel. The beam rose. At one mixture, a faint trace escaped the surface and fell short of the bronze plate. He lowered the water level, diluted again, and adjusted the trough’s side gate. The escaping trace strengthened and struck the lower edge of the plate.\par
A pale mark appeared on bronze.\par
“Acceptable.”\par
“No.”\par
The word came before obedience could stop it.\par
Va-Raedin turned.\par
Sa-Kailan felt the danger at once, but the brightness inside him was too quick.\par
“I mean — this is more than acceptable.”\par
“What is it?”\par
He tried to slow the words enough to make them lawful.\par
“It is a surface that becomes mirror without being made into one. A door that opens or closes because of angle and water. A way to change a path when the source cannot be moved.”\par
Va-Raedin’s face did not change, but the room seemed to tighten.\par
“Continue.”\par
“If light can be kept inside water, water can carry it where air cannot. If salt changes whether it leaves or stays, then the same trough can show, hide, or carry. A man looking down may see nothing. Or see himself. Or see through. Or fail to know that light is moving beneath him.”\par
He stopped because the thoughts were arriving too quickly.\par
Va-Raedin covered the salt.\par
“That is enough.”\par
“No.”\par
Again, too quick.\par
Sa-Kailan lowered his voice.\par
“Please. One more thought. If the source is fixed, and the mark is fixed, and the angle is fixed, the answer is still not fixed. You change what the light crosses. You change what the light is allowed to do.”\par
Va-Raedin regarded him in silence.\par
“You have learned too much for one morning.”\par
It pleased him. Then frightened him.\par
“Is that possible?”\par
“Yes.”\par
“For Light?”\par
“Especially for Light.”\par
Va-Raedin opened a hidden plug. Water drained. The beam broke, blurred, and vanished into useless glow. The trough became a trough. The bronze plate became bronze. The chamber lost its spell.\par
“You ended it.”\par
“I ended your belief that discovery entitles continuation.”\par
“I was not done.”\par
“That was visible.”\par
They returned to the first chamber of the Five Shadows. The mirror stood covered. The lenses hung still. The troughs and vessels waited. Sa-Kailan felt as if the room had rearranged itself, though nothing had moved. A room changed when the mind acquired a new way to be deceived by it.\par
Va-Raedin stood beside the covered mirror.\par
“Name them.”\par
Sa-Kailan gathered himself.\par
“Reflection. Focus. Shadow. Turning. Threshold return.”\par
“Not enough.”\par
He corrected himself.\par
“Sight away from the body. False importance by focus. Darkness as interruption and evidence. Light changing path through substance. Water becoming mirror at the threshold.”\par
Va-Raedin waited.\par
Sa-Kailan understood the silence only after it began to hurt.\par
“And each one is dangerous because each one can be arranged.”\par
“Now you are naming the danger.”\par
The sentence did not satisfy him. It opened hunger.\par
“Are these the five lessons?”\par
“They are five lessons.”\par
“Not the five?”\par
“Do not turn grammar into revelation.”\par
“But they are ordered.”\par
“Everything in Light is ordered.”\par
“Mirrors first because I must distrust where sight stands. Lenses because I must distrust importance. Shadows because I must distrust absence. Turning because I must distrust the path. Threshold return because I must distrust the surface.”\par
Va-Raedin’s eyes narrowed by almost nothing.\par
“Who taught you that?”\par
“You did.”\par
“No. I showed you rooms.”\par
“The rooms taught the order.”\par
“Rooms do not teach.”\par
“Then whoever arranged them did.”\par
Silence entered. Not Faith silence, full of prayer. Not Shield silence, full of readiness. Not Justice silence, waiting for sequence. Light silence: withheld angle.\par
At last Va-Raedin said, “A Candidate learns where light falls.”\par
Sa-Kailan knew the next sentence. A Veil learns who placed it. But he did not say it.\par
The door that changed when position changed. The lens that made seed into wound. The cord hidden inside shadow. The fixed beam moved by salted water. The surface that refused to let light leave. The lessons did not feel like lessons anymore. They felt like steps on a route whose destination had been removed from the map.\par
“A Candidate also learns where he has been made to stand,” he said.\par
Va-Raedin held his gaze.\par
Then, unexpectedly, he laughed once. It was a short, unwilling sound, gone before Sa-Kailan could decide whether it had happened.\par
“You are dangerous when tired.”\par
“Is that praise?”\par
“No.”\par
“Is it warning?”\par
“Yes.”\par
That was almost better.\par
Va-Raedin led him out by a side passage. They climbed a narrow stair, crossed above a dark service void, and entered a small gallery whose walls were carved with old Light markings. Sa-Kailan recognized some: aperture, lamp, eye, mirror, hand. Others he did not. A bowl beneath a star. A line entering water and returning. Five small marks arranged not in a circle, but in a descending sequence.\par
He slowed.\par
Va-Raedin stopped immediately.\par
“What do you see?”\par
“Old marks.”\par
“That is not what I asked.”\par
Sa-Kailan looked again.\par
Eye turned backward. Enlarged seed. Broken shadow. Bent line. Water surface with returning ray. Beneath them ran a narrow empty groove in the wall, leading toward a sealed panel no larger than his hand.\par
“They are not decoration.”\par
“Very little in the Mountain is decoration.”\par
“Records?”\par
“Of a kind.”\par
“Instructions?”\par
“Of a kind.”\par
“For Candidates?”\par
Va-Raedin did not answer.\par
That was enough to make Sa-Kailan’s mouth dry.\par
The five shadows were not only training. They were a language. Perhaps a map. Perhaps a gate waiting for the mind to become guilty enough to read it.\par
Va-Raedin touched the wall beside the marks.\par
“You will not study these yet.”\par
“Yet?”\par
“Do not be greedy over a mercy in grammar.”\par
“Why show me?”\par
“I did not. I brought you through a passage. You looked.”\par
“That distinction is convenient.”\par
“They usually are.”\par
He wanted to ask about the panel. He wanted to ask what a Veil learned after he stopped being Candidate. He wanted to ask whether the rooms had been built to test him, or whether every Candidate touched the same edge and was turned away.\par
He asked none of it.\par
Va-Raedin seemed almost pleased, which annoyed him.\par
They emerged near the eastern practice corridor. Beyond the walls, the ordinary Mountain continued: carts, children, water, gate chains, low voices. Outside Light, repetition sustained life. Inside Light, repetition concealed escalation. A test returned as a lesson. A lesson returned as a sign. A sign returned as a question.\par
At the candidate entrance, Va-Raedin turned.\par
“You will write nothing of today.”\par
“I was not going to.”\par
“You were.”\par
“I do not keep a record.”\par
“No. You keep arrangements in your head and mistake that for obedience.”\par
Sa-Kailan had no answer.\par
“You will not speak of salted water to another Candidate.”\par
“No, Veil.”\par
“You will not ask Balance for salt outside assignment.”\par
“No, Veil.”\par
“You will not attempt threshold return in basin, cup, channel, washing pool, jar, trough, or any other vessel whose proper use your curiosity believes it can improve.”\par
Sa-Kailan looked down.\par
“No, Veil.”\par
“Good.”\par
This time the word was not praise. It was a seal.\par
Va-Raedin began to turn away, then stopped.\par
“Sa-Kailan.”\par
The new name still caught inside him.\par
“Yes?”\par
“What did you learn?”\par
He could have named the five shadows again. He could have made them clean and pleasing. Reflection, focus, shadow, turning, threshold return. He wanted the praise that came when a structure settled correctly in another mind.\par
But the answer was not the list.\par
The answer was the panel. The gallery. The order. The fact that Va-Raedin had ended the exercise not when Sa-Kailan failed, but when he understood too quickly.\par
“I learned that a path can be hidden inside a lesson.”\par
Va-Raedin’s face changed by almost nothing. Almost.\par
“You learned,” he said, “that you enjoy dangerous sentences.”\par
“That too.”\par
“You will suffer for both.”\par
Then he left.\par
Sa-Kailan remained in the passage, alone but not unwatched. He could still feel the desert in him, though stone covered him on every side. He could still hear the Book of Blood accusing the open land. He could still see Va-Sheva’s hand cutting the air before explanation could reach it.\par
Pattern.\par
Reflection. Focus. Shadow. Turning. Threshold return.\par
Behind the eye. Through the lens. Inside the interruption. Across the substance. At the surface that would not let light leave.\par
The five lessons did not merely teach him how to see. 
They taught him that seeing could be staged, enlarged, denied, diverted, trapped — and that the room itself might be asking a question while pretending to correct an answer.\par
No ordinary Candidate was meant to reach the end of a path by asking whether the lessons were a path.\par
Sa-Kailan knew that.\par
He also knew, with a brightness that frightened him more than pride, that somewhere in Light there was an end to be reached.\par
\ornament
