\typesetchapter{Chapter 4}{Oasis}

\noindent The permission came on a strip of pale bark, folded twice and sealed with green wax.\par
For one foolish instant Sa-Kailan thought it belonged to Light. Then he saw the mark. Not silver-white, but Balance green: the cup and sealed jar pressed deep into the wax, crossed by a narrow black witness line and the small hard stamp of Shield.\par
He broke it with his thumb.\par
The old cut pulled under the bandage.\par
The script was plain, written for use rather than beauty.\par
Sa-Kailan, Candidate of Light, is granted one controlled prayer access before the Oasis at third afternoon shadow. Duration: one silence before and one silence after the water prayer. Route: east lower passage, cedar gate, second threshold, shaded descent. Witnessed by Balance. Held by Shield.\par
Below, in a different hand, someone had added:\par
Do not be late.\par
Prayer does not lengthen because boys do.\par
There was no signature.\par
Sa-Kailan read the strip once for permission, then again for hierarchy. Balance had granted. Justice had witnessed. Shield held. Faith would conduct the prayer. Light had no authority here except through him as a body to be admitted, washed, watched, counted, and returned.\par
It annoyed him.\par
It excited him more.\par
Outside.\par
Not free. Not unguarded. Not even fully outside, if his father’s corrections were to be believed. But outside enough for sky. Outside enough for the Oasis. Outside enough for the Mountain’s shadow to fall on his body instead of above it.\par
He dressed more carefully than the rite required. He tied the bandage on his thumb, untied it, and tied it again so the linen sat cleanly over the cut. When he flexed his hand, the skin tightened.\par
His mother watched from the wall recess where she had been folding cloth.\par
“You will keep your eyes lowered during the first prayer.”\par
“I know.”\par
“You will not ask Shield why a path is closed.”\par
“I know.”\par
“And if Balance tells you not to touch a channel, you will not lean over it to see whether the light bends.”\par
Sa-Kailan turned.\par
Her face was calm, but one corner of her mouth had moved.\par
“I do not do that.”\par
“You do not do it when watched.”\par
That was unfair because it had only recently become true.\par
His father looked up from the tablet he had been marking.\par
“Outside is not a reward.”\par
“No.”\par
“It is not a private revelation.”\par
“No.”\par
“It is a place where the Mountain becomes harder to protect.”\par
Sa-Kailan folded the permission bark and placed it inside his robe.\par
“Yes.”\par
His father studied him for a moment, perhaps deciding whether the word had entered or merely passed through.\par
“Stand in the shadow when instructed,” he said. “Do not look long at the desert. It teaches appetite.”\par
That sounded like something adults said because they had once looked too long themselves.\par
Sa-Kailan left before he answered.\par
The route to the cedar gate made familiar passages strange by consequence. He was not wandering. He was being moved outward by sequence. East lower passage. First checkpoint. Balance mark. Witness confirmation. Mouth washing. Cedar gate. Shield count. Second threshold. Shaded descent.\par
Each place had been there yesterday. Today each place asked a question.\par
At the first checkpoint, a Balance clerk checked the green wax against a slate. She examined the time mark, his name, his bandaged thumb, and the absence of anything hanging from his belt.\par
“Permission strip.”\par
He gave it to her.\par
She opened it, read it though she must already have known what it said, and pressed a clay counter into his palm.\par
“Return this counter at the second threshold on re-entry. If you lose it, the route remains open in the record and Shield searches until it is closed.”\par
“I will not lose it.”\par
“That is the usual intention.”\par
The counter was small, rough, and stamped with a number. He placed it in the inner fold of his robe, away from the Light token, as if the two objects might object to touching.\par
At the cedar gate, Faith took his hands and mouth.\par
Not the route. Faith did not command the route. But no one passed outward for prayer without being washed into the proper condition. Two attendants stood beside a shallow basin. Every leaving body was given water, not enough for comfort, only enough for rite.\par
There were seven in the access group: Sa-Kailan, three older women from the western family chambers, a boy with one cloudy eye led by his aunt, and two men who looked too tired to be anything but workers given prayer because grief had entered their household. One of the men carried no offering, which meant his request had been accepted under sorrow.\par
The Faith attendant poured water over Sa-Kailan’s hands. It was colder than he expected. It split around his knuckles and fell into the catch bowl. The old curiosity rose at once, bright and inappropriate.\par
The stream enlarged his fingers, thinned them, broke them, returned them.\par
He lowered his eyes before the attendant saw where they had gone.\par
“Speak only what is appointed before the Oasis,” she said.\par
“I will.”\par
“Take nothing from the ground.”\par
“I will not.”\par
“If the wind rises, turn with Shield, not with fear.”\par
He looked up.\par
The attendant held his eyes.\par
“Fear often faces the wrong direction.”\par
The boy with the cloudy eye began to cry when the water touched his mouth. His aunt bent close and whispered too softly for Sa-Kailan to hear. The two grieving men did not speak at all.\par
Beyond the cedar gate waited Shield.\par
The gate was not made entirely of cedar. Behind the wood were black iron bands and a locking bar thick enough to stop a cart. Two wardens stood before it. A third watched from a side niche, visible only because Sa-Kailan had begun to notice where watching could be hidden.\par
The senior warden read their marks one by one.\par
“Prayer access. Third afternoon shadow. Seven bodies out. Seven bodies due.”\par
He looked at Sa-Kailan last.\par
“Candidate of Light.”\par
“Yes.”\par
“Eyes where Faith places them. Feet where Balance marks them. If Shield speaks, you obey before understanding.”\par
“I understand.”\par
“No,” the warden said. “You agree. Understanding is not required.”\par
The iron bar was lifted by two hands and a lever set into the wall. A narrow bell sounded somewhere beyond the gate, answered by another farther down the stone.\par
The Mountain did not simply open. It informed itself that it had opened.\par
They passed through in single file.\par
The descent beyond the cedar gate was cut between natural rock and worked stone. At first it felt like any lower passage: dim, damp, controlled, with lamps set in protective recesses and water smell in the cracks.\par
Then the lamps stopped.\par
Light lay on the floor in a pale wedge.\par
No lamp-light. No measured reflection. No slit calculated by Light. It came from somewhere ahead, full of dust and height, careless in its angle, making each stair deeper and then shallower as Sa-Kailan moved.\par
The woman in front of him inhaled too sharply.\par
The Shield warden said, “Eyes down until the mark.”\par
They obeyed.\par
Sa-Kailan lowered his gaze to the steps. 

Outside light was not brighter only. It was undisciplined. His training recoiled from it. His body wanted it.\par

At the last turn, the Mountain opened.\par
Sky struck him first not as blue, though it was blue, but as height. The Mountain had ceilings. Even its highest halls taught the eye to expect an end. The sky did not end.\par
He kept his eyes low, but not low enough.\par
Above the dark shoulder of the Mountain, the open brightness continued without vault, beam, lamp, mirror, aperture, or permission.\par
Faith led them to the marked prayer stones.\par
Only there could they look.\par
The Oasis lay under the Mountain’s shadow.\par
Children’s gravures made it seem like a spring circled by trees. It was not that simple. It was a managed hollow: rock, shade, water, channel, fence, pen, reed, herb, work path, prayer stone, marker, outer line. Palms leaned inward over the shade. Hard-leaved trees grew where their roots could drink without stealing from the channel. Reeds marked slower water. White flowers clung to cracks near the spring. Moss darkened the stones closest to the source. Clay tubes for bees hung beneath reed awnings. Goat pens stood under woven shade. Damp trays for snails were stacked where the air stayed cool.\par
Beyond it all, the desert expanded.\par
Heat moved over the ground in pale shimmers. The Mountain’s shadow covered the Oasis like a hand. Outside that hand, the light was hard enough to erase mercy.\par
Sa-Kailan understood why children were told the Oasis was hidden from the eye above. The Mountain rose behind and over it, cutting the sky and holding the hollow in shade. From certain heights, perhaps, the water would vanish into shadow and stone. Light had been made to protect it, or the Mountain had grown in such a way that Faith could say God had placed it there and Light would have difficulty disagreeing.\par
The old women knelt.\par
Sa-Kailan knelt with them.\par
The Faith attendant unrolled a strip of pale bark and held it with both hands before the water.\par
This was not the children’s prayer.\par
It was the Book of Blood.\par
\begin{scripture}
And after the fifth day, God spoke unto Adam and unto Eve, and unto the five children whom He had set in the Oasis beneath the shadow of the Mountain; and God said, Let there be Keepers among you, that the Mountain be not without order, nor the Oasis without care, nor the people without peace. And God called the woman of water, and said unto her, Keep thou the Balance, and gather the fruits, and count the waters, and keep the resources of survival. And God called the woman of holy speech, and said unto her, Keep thou the Faith, and teach My Gospel unto the children, and pray beneath the Mountain; for I shall provide.
\end{scripture}
Balance workers moved in the distance as the words were spoken. One woman checked the level in a narrow cut along the main channel. A boy carried wet leaves toward a shaded rack. A clerk marked something on a hanging board and moved a shell counter from one hook to another.\par
Count the waters, the scripture said.\par
And there they were, counting.\par
\begin{scripture}
And God called the man of strength, and said unto him, Keep thou the Shield, and stand before the beasts of the Earth, and suffer not the children to be devoured. And God called the woman of brightness, and said unto her, Keep thou the Light, and illuminate the Mountain, that the people stumble not in darkness. And God called the man of peace, and said unto him, Keep thou Justice, and stand between brother and brother, that blood answer not blood without measure. And the Keepers received their charges, and each kept what had been given, and the people multiplied in the Oasis.
\end{scripture}
The Oasis made the five charges less like doctrine and more like tools laid out on stone. Balance kept water moving. Faith transmitted the reading. Shield watched the line where shade became exposure. Light revealed enough to move without stumbling. Justice was not visible, but the clay counter inside Sa-Kailan’s robe said Justice had already entered the route.\par
The text continued.\par
\begin{scripture}
And the Keepers had sons and daughters, and their children had children after them; and the Oasis was filled with voices, and the Mountain gave shadow, and the waters were counted, and the Gospel was taught, and the beasts were driven back, and the lamps were lit, and peace was kept among the people. And the Keeper of Shield had two sons; and the elder was great in strength, and broad in the shoulder, and mighty before the beasts; and the younger was small in his body, and cunning in his heart. And the younger looked upon his brother, and saw that the people loved his strength, and envy entered him as a serpent entereth the grass.
\end{scripture}
Sa-Kailan had heard the beginning many times. Children loved the elder brother because he was easy to love: strong, brave, loved by the people, broad before the beasts. They loved the younger too, before they understood they were not meant to. Smallness and cunning did not sound evil to a child who had neither size nor strength.\par
Now, kneeling under the open sky, the younger brother sounded less like a monster than a warning placed too close to his own ribs.\par
\begin{scripture}
And the younger spoke softly unto his brother, and smiled with his mouth while murder grew within him; and he rose against his brother in secret, and slew him where no beast had entered. And the first blood of man cried from the dust. And fear came upon the people, for Shield had defended the Oasis from the beasts without, but had not seen the beast within. And the younger fled from the Oasis, and took with him the wife of his brother, a daughter of Faith, and went out from the shadow of the Mountain into the open land.
\end{scripture}
The first blood of man cried from the dust.\par
Sa-Kailan looked toward Shield before he meant to. The perimeter wardens had not moved. Their stillness was not rest. It was arranged attention.\par
Shield had failed once in scripture because it had looked outward and not inward. Every warden since had inherited the shame of that line.\par
\begin{scripture}
And God saw that blood had entered the world of men, and that holy speech had been carried away by the murderer; and God cursed them, saying, Thou shalt wander under the eye of Heaven, and the Earth shall not hide thee. Thy children shall thirst for what they cannot keep, and conquer what they cannot love. Thou shalt remember the Mountain, but thou shalt not find it; and thou shalt hear the Gospel, but twist it into hunger. And from the murderer and the stolen daughter came the Outsiders.
\end{scripture}
The boy with the cloudy eye stopped crying.\par
His aunt held him so tightly that her knuckles whitened.\par
The Outsiders.\par
Not merely strangers. Kin deformed by blood, if the reading was true. Descendants of murder and stolen Faith. Children of a hunger that remembered the Mountain without finding it.\par
Sa-Kailan did not know whether he believed it in the way Faith wanted belief. But he believed enough. He believed brothers could envy brothers. He believed blood could change a people’s story. He believed holy words could be carried away and broken by those who needed them to justify desire.\par
That was the danger of the reading. It did not need to prove everything.\par
It only needed to be plausible where fear already lived.\par
The attendant lowered the bark.\par
For a while no one moved.\par
The silence after scripture was supposed to be empty. This one was not. It contained the spring, the goats, the bees, the old women, the boy with the damaged eye, the grieving men, the wardens, the desert, and the old story of a brother smiling while murder grew behind his mouth.\par
Then the practical world returned.\par
Balance workers resumed. A woman in green-brown checked the channel level against the cut mark. Another scraped fibres from soaked stalks and laid them over a stone rail. A child sprinkled water over snail trays with the seriousness of a priest.\par
Goats chewed under woven shade. Hides stretched on frames. Sinew dried in narrow strips. Bees moved in and out of clay tubes. Wax sheets lay under mesh, pale and faintly golden, protected from dust. Shells were crushed near the herb wall. Dung was gathered in a covered basket for drying. Nothing here was merely alive. Everything was being converted into survival.\par
Milk, hide, sinew, wax, honey, shell, fibre, medicine, fuel, cord, bandage, wick.\par
Balance was not abundance. It was refusal: refusal to waste, refusal to romanticise scarcity, refusal to let appetite pretend it had no consequence.\par
Faith made the refusal bearable. The attendants kept the group in rhythm: stand, silence, kneel, touch no channel, face the spring, lower eyes, answer. The words did not slow the work around them. They synchronized it. The prayer stones, the channels, the racks, the markers, the pens, the shade-lines: all of it seemed arranged so that devotion and logistics could occupy the same ground without confessing which one governed the other.\par
Sa-Kailan was allowed to stand after the first silence.\par
Not approach the spring. Not touch the channel. Only step to the third marker and look.\par
The reason was obvious.\par
Light candidates leaned toward water.\par
He stepped to the marker.\par
That was when he saw Va-Sheva.\par
She stood near the northern perimeter beyond the herb beds, where the Oasis began to turn back toward bare rock. Three Shield wardens were placed around her, though none had announced command. Authority did not always require speech. She wore dark Shield leather over her robe, a cloth wrapped around her throat against sun and grit, hair bound high, short blade at her hip, longer one across her back.\par
Inside the Hall, she had looked severe. Here she made sense.\par
She was looking away from him, toward the outer path.\par
After the reading, he saw her differently. Not less beautifully. That would have been easier. But Shield now seemed older than leather and blade. Shield had been born because beasts came from without and murder from within. Shield had failed once by not knowing the difference quickly enough. Va-Sheva’s stillness was an answer to that first failure.\par
One of the wardens spoke to her. She did not turn her head, only lifted two fingers. The warden moved left. Another shifted near the rocks. The perimeter changed shape around a thought.\par
Then she saw him.\par
Her expression altered by almost nothing. Recognition. Surprise, perhaps. Something close to amusement, contained before it could become warmth.\par
She crossed toward him while the Faith attendant moved away to assist the old women.\par
Sa-Kailan became suddenly aware of his robe, his bandaged thumb, the dust on his sandals, the fact that he had been looking too long at everything.\par
“Sa-Kailan,” she said.\par
The prefix in her voice wounded and delighted him at once.\par
“Va-Sheva.”\par
Her eyes went to his thumb.\par
“Light has already started cutting pieces off you.”\par
“Only the useless ones.”\par
“That cannot be true. You are still standing.”\par
He almost laughed. She did not. Her mouth moved enough to show she had intended the cruelty kindly.\par
“You were at the Hall,” he said, because he had to say something and chose badly.\par
“So were most of the people.”\par
“I meant—”\par
“I know what you meant.”\par
Wind moved between them, carrying goat, wet stone, smoke, and something dry from beyond the shade.\par
“You have grown,” she said.\par
The words should have pleased him. They did not. They belonged to someone who had last measured him as a child.\par
“So have you,” he said, then hated himself.\par
This time she did smile, and the smile was worse because it was real.\par
“I had four years’ advantage.”\par
“Five, almost.”\par
“You counted?”\par
He looked away toward the channels.\par
“Everyone knows.”\par
“No,” she said. “Boys know such things when they should be learning where to put their eyes.”\par
The rebuke was mild, but it landed where she meant it.\par
He felt himself become fifteen in front of her, intolerably and completely.\par
“I am learning,” he said.\par
“So I heard.”\par
“From whom?”\par
“Shield hears many things.”\par
“That is not an answer.”\par
“It is the only answer Shield enjoys giving.”\par
He looked at her again. The girl he had known remained somewhere in the face, but duty had been worked over it like leather. She was not pretending severity. Even now, while speaking, some part of her watched the perimeter.\par
“You are on duty.”\par
“Yes.”\par
“Then should you be speaking to me?”\par
“No.”\par
The answer startled him.\par
“But I am,” she said.\par
That should have made him happy. Instead it made him careful.\par
A gift from Va-Sheva was never simple because he always wanted it to mean more than she had given.\par
“Why?”\par
“Because you looked as if the sky had insulted you.”\par
“It is too large.”\par
“That is one of its lesser faults.”\par
“You do not like it?”\par
“I respect it.”\par
“That is not the same.”\par
“No.”\par
For a moment they stood side by side, not touching, both looking outward.\par
Sa-Kailan wanted to ask whether she had ever gone past the second ridge, whether the desert felt different without Faith nearby, whether Shield feared the Outsiders as scripture feared them or as wardens feared tracks in dust.\par
The questions were too hungry. She would hear the hunger before the words.\par
He chose a safer danger.\par
“The reading said the Outsiders came from the murderer.”\par
Her face changed. Not much. Enough.\par
“That is what the Book of Blood says.”\par
“You say it as if there is another way to say it.”\par
“I say it as one on duty during prayer access.”\par
The warning was cleanly given. He should have stopped. The reading had opened something in him, however, and the desert was still before him, and Va-Sheva was Shield.\par
“Do you believe it?”\par
Va-Sheva looked toward the outer path before answering.\par
“I believe the Mountain is hidden for a reason.”\par
“That is not an answer.”\par
“Do you believe the Book of the Beast?”\par
Sa-Kailan had expected refusal, or one of those Shield sentences shaped to close where he wanted opening. Instead she had placed another book between them.\par
Not Blood, with its brother and murderer. Not Genesis, with Mountain, Oasis, forest, and desert. The Beast belonged to older recitations, to gravures shown by low flame, to prayers spoken when gates were sealed. It belonged to the idea that the Mountain did not only hide from the Outsiders. It held something against them.\par
“You mean,” he said carefully, “that the five Keepers hold together the Beast of Hell, in case the Outsiders come back?”\par
Va-Sheva did not blink.\par
“That is what children say.”\par
“It is what Faith teaches children.”\par
“Faith teaches children the shape of things before they can bear the weight of them.”\par
Sa-Kailan looked toward the dark line of the outer path. Hell belonged below, beneath stone, beneath water, beneath red pigment in the carved warnings. Yet the Book of the Beast had always joined two directions that unsettled him: the thing beneath and the people beyond. One terror held in darkness. One terror moving in light.\par
“So there is another way to say that too.”\par
“There are always other ways to say what must not be mishandled.”\par
“That is not an answer.”\par
“No,” Va-Sheva said. “It is Shield trying to keep a boy from turning scripture into a toy.”\par
The word boy struck because her voice did not rise.\par
“I am Sa-Kailan now.”\par
“I know what you are called.”\par
“Then speak to me as if I can understand.”\par
Her gaze came to him fully then. Still, some part of her listened outward.\par
“You want understanding as if it were water owed to your measure,” she said. “You have been Candidate for days. You have stood outside once. You have heard one formal reading, and now you are ready to ask whether Blood is true and Beast is true, as if truth were waiting beside the channel for you to lean over it.”\par
The rebuke entered cleanly because it was not unfair.\par
“Then why ask me?”\par
“Because you asked whether I believed the Book of Blood.”\par
“And the answer is the Book of the Beast?”\par
“The answer is that belief is not the only duty scripture requires.”\par
Sa-Kailan said nothing.\par
Va-Sheva turned slightly, enough that they stood no longer face to face but beside one another, both looking toward the desert. The gesture made the conversation less private and more dangerous.\par
“The Book of Blood tells you why the Mountain closed,” she said. “The Book of the Beast tells you why it must not open carelessly.”\par
“Because of the Outsiders.”\par
“Because of what happens when the Outsiders find what should remain hidden.”\par
“The Beast.”\par
She looked at him with something almost like pity, controlled and unwelcome.\par
“Listen to yourself. You say Beast as if naming it gives you possession of the fear.”\par
He hated that because it was true.\par
The word had thrilled him since childhood. Beast. Hell. Chains. Five Keepers. The great last defence. The sacred terror held in reserve for the day when the children of the murderer returned to devour the Mountain’s shadow.\par
Children loved what frightened them when adults made the fear ceremonial enough.\par
“What does Shield say?” he asked.\par
“Shield says a thing does not need to be understood by everyone in order to be guarded by everyone.”\par
“That sounds like Ka-Dhavar.”\par
“It should.”\par
“And what do you say?”\par
She was silent long enough that he thought she would not answer. Then one of the wardens near the outer markers shifted, and her eyes moved to him before returning to Sa-Kailan.\par
“I say that a gate is not made holy because there is beauty behind it. Sometimes a gate is holy because there is something terrible behind it, and opening it wrongly would make the opener guilty of every death that followed.”\par
The words were not from a child’s lesson. He felt that at once.\par
“The Beast is real, then.”\par
“I did not say that.”\par
“You almost did.”\par
“No. You almost heard it.”\par
The frustration returned, but now fear entered it. Since Va-Raedin had shown him a lamp could make a false door more persuasive than a true one, nothing had been clean. The Oasis itself stood before him now as miracle and mechanism, scripture and channel, mercy and ration.\par
Va-Sheva seemed to read enough of that in his face to step back from what she had nearly given him.\par
“The Oasis is smaller than I remembered,” he said, because retreat had become necessary and he needed ground that did not look like surrender.\par
“Most holy things are.”\par
“And the desert?”\par
“Larger.”\par
He looked toward it.\par
The outer path ran from the Oasis along a fold of stone and vanished between two dark ridges. Beyond it, the desert opened in layers of heat and distance. On a rock far out, something flashed briefly, then disappeared.\par
He narrowed his eyes.\par
“What is that?”\par
Va-Sheva’s gaze moved instantly.\par
“Where?”\par
“There. Past the second ridge. The dark stone.”\par
She followed the line. For one moment she said nothing.\par
Then, very softly: “Do not point.”\par
His hand had begun to lift. He lowered it.\par
The softness left her face.\par
“Stay here.”\par
She turned from him completely.\par
What followed was too quiet to be dramatic. There was no shout. No drawn blade. No rush of bodies. Va-Sheva made three gestures in succession: two fingers low, palm closed, hand cut toward the outer path.\par
The Oasis rearranged itself.\par
One warden moved behind the goat pens and became invisible in shade. Another knelt near the dust at the outer marker. A third shifted between the prayer group and the spring. A hidden figure answered from the rock above the descent, where Sa-Kailan had seen only stone. Then another signal came from higher on the slope.\par
Shield had been around them the whole time: visible wardens, hidden bodies, lines of sight, markers arranged so disturbance could speak before danger had language.\par
The Faith attendant looked up, alarmed, but did not speak. One of the old women tried to rise and was pressed gently back by her companion. A goat stopped chewing.\par
The bees continued.\par
Va-Sheva crouched beside the kneeling warden and touched the ground. She lifted her fingers, rubbed something between them, then looked toward the outer path, the Balance terraces, the prayer group, and the rock where Sa-Kailan had seen the flash.\par
A Balance worker had gone still beside the snail trays, water ladle in hand.\par
Va-Sheva pointed to him, then to the shaded wall. He moved at once.\par
The warden near the stones gave a low call.\par
“Marker.”\par
The word was ordinary. Shield changed around it.\par
One outer marker had been turned. Barely. A flat side faced inward where the others faced skyward.\par
Va-Sheva did not touch it. She looked at the dust around it, the path beyond, the nearest channel, the position of the prayer group, and the rock beyond the ridge.\par
Then she spoke aloud.\par
“No one returns inside yet.”\par
The Faith attendant’s eyes widened.\par
“The access period—”\par
“Has changed.”\par
“By whose order?”\par
Va-Sheva looked at her.\par
The attendant lowered her head.\par
“Shield holds the threshold.”\par
“Good.”\par
Sa-Kailan heard the words enter the Oasis like a bar lowered across a gate. Faith governed prayer. Balance governed water and work. But here, at the edge of the world, Shield could interrupt sacred timing with two words and no explanation.\par
A young Balance boy near the herb beds began to cry. He was old enough to be ashamed of it and too young to stop.\par
Va-Sheva walked the marker line without stepping beyond it. At the turned stone she knelt, took a thin bone pin from her sleeve, and lifted a strand from the ground. It caught the light once, pale against her glove. She placed it in a covered dish handed by a warden.\par
“Second gate,” she said.\par
A hidden watcher answered from the slope.\par
Then she turned toward the workers.\par
“Count.”\par
An older Balance woman began at once.\par
“Goat pens: three visible, one shaded, all tied. Snail trays: two open, four covered, no tray missing. Herb knives: five issued, five in hand. Water ladles: eight.”\par
“Show eight.”\par
Eight ladles lifted. One child’s hand shook.\par
“Fibre hooks?”\par
“Three.”\par
“Show three.”\par
Three hooks lifted.\par
“Outside workers?”\par
“Eleven.”\par
“Name them.”\par
The Balance woman named them without slate. One by one, bodies answered.\par
When all eleven had answered, Va-Sheva nodded to the warden.\par
“Prayer group?”\par
The Faith attendant named seven.\par
The warden corrected her.\par
“Six.”\par
The attendant froze.\par
Sa-Kailan’s stomach tightened before his mind understood.\par
Then one of the old women said sharply, “Mara.”\par
The woman beside her looked around.\par
“She was at the spring rail.”\par
No one moved.\par
Va-Sheva did. She did not run. Running would have told everyone fear had outrun command. She crossed the prayer area at a fast walk, two wardens spreading behind her.\par
Sa-Kailan saw then that one of the older women from the access group was missing. Not the aunt, not the boy, not the grieving men. An old woman who had knelt nearest the spring basin and moved slowly because one hip did not bend properly.\par
The Faith attendant whispered, “God keep her in shadow.”\par
Va-Sheva heard.\par
“Silence.”\par
The word cut the air.\par
They found the woman by the spring rail. She had crouched there with one hand against stone, breathing hard, unable to rise. Whether she had wandered intentionally toward the water or simply lost balance, Sa-Kailan could not tell. In her other hand she held a small wet leaf.\par
A Balance worker made a sound of anger low in her throat.\par
Va-Sheva took the leaf gently, gave it to the worker, then helped the old woman stand. She did not humiliate her.\par
That surprised Sa-Kailan.\par
Va-Sheva only leaned close and said something that made the woman close her eyes and nod.\par
Then Va-Sheva signalled.\par
The warden near the channel examined the edge. The turned marker remained under guard. The rock beyond the second ridge flashed no more. The old woman was returned to Faith, who received her with a face drained of colour.\par
Va-Sheva came back to the inner shade. Her eyes passed over the group, counted, and settled briefly on Sa-Kailan.\par
“Seven bodies out,” she said. “Seven bodies held. Eleven Balance workers held. No one returns until the marker is reset and the channel is cleared.”\par
The Faith attendant said very quietly, “She is old.”\par
“Yes.”\par
“She meant no harm.”\par
“Meaning no harm is not the same as remaining harmless.”\par
The old woman began to weep silently, with such shame that Sa-Kailan had to look away.\par
The leaf had perhaps been for medicine, or memory, or some private superstition. The reason might have been tender. Shield treated it as breach until tenderness could be safely restored.\par
The containment ended as quietly as it had begun. The turned stone was lifted with bone pins and examined beneath a cloth. The channel was checked. A small clay seal was pressed beside the woman’s name. The prayer group was recounted. Balance workers returned to task with the rigid annoyance of people who had lost time but not permission to complain of it.\par
When Va-Sheva finally came back to Sa-Kailan, the official danger had passed, though nothing in her face softened.\par
“You saw the flash first.”\par
“I think so.”\par
“You did not point when told not to.”\par
“No.”\par
“Good.”\par
The word struck harder than it should have.\par
“Was it connected?” he asked.\par
“To the marker?”\par
“Yes.”\par
“Probably not.”\par
“Then why—”\par
She looked at him.\par
He stopped.\par
“Because two small wrongnesses at a threshold are not treated separately until Shield can afford the luxury,” she said. “A flash may be nothing. A turned marker may be nothing. An old woman out of place may be nothing. Three nothings close together become a shape.”\par
What is unseen can be known by what it disturbs. State the consequence of each choice. Curiosity is not free. Do not admire the flame. Ask where the air enters.\par
And now this.\par
Three nothings close together become a shape.\par
“You see like Light when you say that,” he said.\par
“No. Light sees what can be shown. Shield sees what must be stopped before it shows itself.”\par
It was not boastful. That made it more powerful.\par
He looked toward the desert again. The rock beyond the second ridge sat dark and ordinary. The flash might have been mineral, bird, old metal, heat, an eye.\par
“Have you gone beyond the second ridge?”\par
Va-Sheva did not answer immediately.\par
“That is not a prayer question.”\par
“No.”\par
“Then do not ask it during prayer access.”\par
He nodded.\par
But she had not said no.\par
The Faith attendant called the group back to the prayer stones for the return blessing. The access period had been shortened and extended both, made irregular by containment. Justice would dislike the irregularity. Balance would record lost task time. Faith would smooth fear with words. Shield would remember the turned marker longer than anyone remembered the prayer.\par
Va-Sheva stepped aside to let him pass.\par
“Sa-Kailan.”\par
He turned.\par
Wind lifted dust between them. With the Mountain shadow behind her and desert light along her cheek, she seemed less unreachable because she stood close enough to touch. Then the foolishness of that thought revealed itself.\par
She was not unreachable because of distance.\par
“You did well to notice,” she said.\par
The praise warmed him before he could defend against it.\par
Then she added, “But do not start believing the outside speaks to you.”\par
There it was. The correction inside the gift.\par
“I did not.”\par
“You are fifteen. Everything speaks to you.”\par
“I am not a child.”\par
“No,” she said. “Not entirely.”\par
The return blessing required them to wash dust from their fingers in a shallow bowl before re-entering the descent. Sa-Kailan watched the dirt darken the water. The Faith attendant spoke over each of them.\par
“What has touched the open returns under shadow.”\par
The group answered.\par
“Under shadow, we return.”\par
When his turn came, Sa-Kailan placed his fingers in the bowl. The water was warm now from sunlight and use. His reflected hand broke around the dust, the image bending, thickening, returning.\par
But Shield was counting. Faith was watching. Balance had already measured the bowl’s use. The outside did not lengthen itself for his curiosity.\par
They re-entered the Mountain in single file.\par
The first steps down felt wrong. The passage narrowed quickly. The ceiling returned like a hand over the skull. After the sky, the lamps seemed mean and intelligent. After the desert, the stone smelled of bodies, oil, and old water.\par
Yet relief came too, shameful and immediate.\par
The Mountain enclosed him, and some part of him was grateful to be enclosed.\par
Outside had not been freedom.\par
At the second threshold, he returned the clay counter. The Balance clerk checked the number against the slate. Shield marked seven bodies in. Faith took the dust bowl away. A Justice witness, seated almost invisibly near the wall, made a mark after the gate closed.\par
The cedar face of the inner door shut.\par
The iron bar fell.\par
The Mountain informed itself that they had returned.\par
Sa-Kailan stood in the passage a moment too long.\par
Behind him lay the Oasis: water in shadow, bees, goats, snails, herbs, channels, prayer stones, the Book of Blood spoken under open sky, and Va-Sheva’s hand cutting the air before explanation could catch up. Beyond it lay the desert, immense and bright and watched from places he had not known to look.\par
Before him lay the corridors of the Mountain, lit by measure, closed by rule, alive with permissions.\par
The Oasis was proof. He could not deny that. A hidden spring in the Mountain’s shadow. Water held against the open world. Food, fibre, medicine, wax, milk, seal, wick, hide, cord, prayer, count. It made The Covenant physical in a way no gravure could.\par
But the reading had made the desert proof too.\par
It was where the younger brother had fled. It was where holy speech had been stolen and broken. It was where the children of the murderer had wandered under the eye of Heaven, remembering the Mountain without finding it, twisting the Gospel into hunger.\par
The desert was not freedom in the Book of Blood. It was inheritance. It was curse.\par
And the Book of the Beast had added something worse, though Va-Sheva had barely opened the door to it.\par
If Blood explained why the Mountain had closed, Beast explained why the closing could not fail.\par
Five Keepers. Five charges.\par
Five hands kept apart because whatever slept beneath the Mountain was not merely a story for frightened children.\par
Or if it was a story, it was a story Shield treated like a gate.\par
Sa-Kailan touched the folded permission bark inside his robe, though it had already spent its authority.\par
The outside had not released him from the Mountain.\par
It had explained why the Mountain held so tightly.\par
\ornament
