\chapter{Mirrors}

\noindent Sa-Kailan arrived before first work with an empty stomach, a bandaged thumb, and the private conviction that he would not be made to look foolish twice.\par
The eastern practice halls lay beyond the public chambers of Light, past the polished corridors where visitors were permitted to admire lamps without understanding why they had been placed where they were. He had walked those corridors before, but never alone, never with the new name still raw around him, never with the cut in his thumb pulsing under linen each time his heart remembered it.\par
Yesterday he could have passed these thresholds only as Kailan: a boy under instruction, tolerated near the outer rooms because his teachers had vouched for him. This morning the first eastern gate did not ask for tolerance. It asked for record.\par
The stone changed under his sandals as he crossed from common passage into Light territory. It became smoother, colder, cut at shallow opposing angles so that a step did not merely sound; it returned a faint image from below. Even the darkness seemed tended here, assigned rather than left to gather where it pleased.\par
His mother had seen the bandage and asked whether the cut had closed cleanly. His father had asked whether Va-Raedin had named an hour. Neither had asked whether he was afraid. It would have been unkind to give fear that much room before first work.\par
At the first eastern gate, an attendant in pale grey examined his bandage, then his face, then a narrow tablet fixed beside the door. Her finger paused beneath a line so newly cut that pale dust still gathered inside the letters.\par
“Sa-Kailan of Light,” she said.\par
The name sounded more official in a stranger’s mouth. Less like honour. More like a key that had already turned in locks he had not seen.\par
The attendant opened the way without another word. Sa-Kailan passed into a corridor crossed by narrow shafts of light. They did not fall evenly. They arrived in bars, thin and pale, from slits too high to see, then vanished into black cuts along the opposite wall. Dust moved through them: visible, invisible, visible again, each mote existing only where the beam permitted it.\par
He slowed despite himself.\par
Faith reached its chambers by obedience. Balance by measure. Shield by permission. Justice by summons. Light, he had always thought, worked differently. Light showed. Light revealed. A man did not need to believe a lamp; he had only to open his eyes.\par
Then he remembered Va-Raedin’s voice in the Hall.\par
Do not admire the flame. Ask where the air enters.\par
Sa-Kailan looked again at the shafts crossing the passage. The obvious thing was their precision. The better question was where they began, who had narrowed them, why half of them were received by slots instead of stone, and what waited in the darkness that had been denied the beam.\par
He reached the practice hall before he had answered.\par
The Hall of Light was not one hall, although it had one name. From the north it appeared broad and shallow. From the eastern stair it narrowed into a blade of space, with mirrors running along one side and black apertures cut into the other. From the candidate entrance, where Sa-Kailan stood now, it appeared almost empty.\par
Then the room began to give itself away.\par
Mirrors stood everywhere, but none where a child would have placed them. Some were tall strips no wider than a hand. Others were round and small, set at knee height or above the shoulder. A few were not mirrors at all, but dark polished stone that returned shape without obedience. Lines had been cut into the floor: arcs, crosses, narrow triangles, circles left deliberately incomplete. Lamp shafts entered through high slits. The lamps themselves were hidden. At the far end, three doorways stood side by side. One was open and dark. One was closed by bronze. One appeared half-open, though Sa-Kailan distrusted it immediately.\par
Va-Raedin stood near the centre with his hands folded behind his back.\par
He had not changed after the Ascension, or had changed into something identical: pale grey robe, no ornament except the narrow silver thread at the cuff that marked him as Veil of Light. He looked rested. Sa-Kailan resented that. The ceremony had lasted half the day, and Va-Raedin had stood through it without shifting, leaning, or submitting to any of the small disgraces by which bodies confess themselves.\par
“You are early,” Va-Raedin said.\par
“You said before first work.”\par
“I did.”\par
The hall hummed faintly. Water, air, metal, hidden lamps: Sa-Kailan could not yet separate them.\par
Va-Raedin looked at the bandage.\par
“Does the thumb hurt?”\par
“No.”\par
“It should.”\par
“It does a little.”\par
“Then begin with truth where truth is inexpensive.”\par
Heat moved into Sa-Kailan’s face. He said nothing.\par
Va-Raedin turned toward the three doorways.\par
“Which door opens?”\par
Sa-Kailan almost answered quickly and disliked himself for it. The left doorway was open and dark. The middle was sealed by bronze. The right showed a line of brightness along one edge.\par
The right was obviously a lure, which meant the lure might be to distrust it too easily. He studied the floor. A thin strip of light lay at the base of the left doorway. The right doorway’s brightness touched nothing. The bronze door was too solid, too central, too theatrical.\par
“The left,” he said.\par
“Why?”\par
“The light at the base belongs to an opening. The right only reflects one. The middle is meant to make the eye feel opposed.”\par
“Open it.”\par
Sa-Kailan crossed the hall, careful not to step on the cut lines though he had not been told whether they mattered. He set his hand against the left door’s inner edge and pushed. Nothing moved. He pulled. Nothing. He searched for a latch, seam, hinge, weight, pressure. Smooth stone answered him everywhere.\par
Behind him, Va-Raedin did not speak.\par
Sa-Kailan hated the silence more than correction. He stepped back. The right door’s half-open edge was a polished strip catching an upper lamp. The middle door remained bronze, cold, closed, without handle. He pressed it once, because anger sometimes insists on humiliating itself physically. Nothing.\par
“You asked which door opens,” he said.\par
“I did.”\par
“Not which door appears open.”\par
“No.”\par
Sa-Kailan looked again. Doors, mirrors, floor. The nearest cut circle was not decorative. Its inner edge was worn where feet had stood often, but only on one side, as if candidates had learned to step into it from the same direction. He stood inside it.\par
The room changed.\par
Not by moving. By being seen from the point for which it had been arranged. The mirrors on the left wall no longer reflected the three doors. They caught instead a plain stone panel behind him, flush with the wall, unmarked by light. He turned, found the seam by touch, and pressed.\par
The panel opened inward.\par
Sa-Kailan tried not to smile. Va-Raedin saw him trying.\par
“You are pleased because you found the fourth door.”\par
“I found the door that opens.”\par
“You found it after failing.”\par
“I corrected the failure.”\par
“That is better than defending it. It is not the same as succeeding.”\par
Sa-Kailan let his hand fall from the opened panel. The chamber beyond smelled of cold metal and lamp oil.\par
Va-Raedin came beside him.\par
“What was the lesson?”\par
“That light deceives.”\par
“No. That is what boys say when a lamp embarrasses them.”\par
Sa-Kailan tightened his jaw.\par
“The lesson is that sight depends on position,” Va-Raedin said. “From the entrance, you saw three doors and chose among them. From the circle, you saw the room that had arranged the choice. The hidden door was not hidden because stone wished to be secret. It was hidden because you accepted the first place given to your eyes.”\par
Sa-Kailan looked back at the worn circle.\par
“A Candidate learns where light falls,” Va-Raedin said. “That is already more than most men ever learn. A Candidate who follows brightness serves whoever placed the lamp.”\par
“A Veil learns who placed it,” Sa-Kailan said.\par
“Eventually. If he stops being proud of finding doors built to be found.”\par
The correction landed cleanly because it was deserved. Sa-Kailan entered the dark chamber before he could be told, partly to escape his own expression.\par
The second room was narrower and colder. Lamps hidden behind the walls sent light through slits so thin they should not have lit anything, yet the chamber glimmered in fragments. A stone table stood at the centre. On it lay a mirror shard, a black measuring rod, a red glass bead fixed to a white tile, a bowl of water, two small bronze wedges, and a square of cloudy glass.\par
Va-Raedin closed the hidden door behind them. Darkness thickened, then organized itself around the slits.\par
“Move the red point onto the far wall.”\par
Sa-Kailan looked at the far wall. It was blank.\par
“May I touch the lamp?”\par
“No.”\par
“The table?”\par
“You may touch what is on the table.”\par
He began with the mirror shard. The nearest slit sent a fine blade of light above the table, too high to catch unless the mirror was lifted. He angled the shard until light jumped downward, bright and disobedient. It crossed the black rod and vanished.\par
He moved the rod, then realized the rod was not an obstacle but a measure. Its shadow showed the beam’s angle more precisely than his eye did. He set one bronze wedge beneath the mirror to hold it steady. The light struck the cloudy glass and widened, then struck the tile and briefly woke the red bead into a smeared glow. He adjusted the glass. The glow sharpened, slid up the wall, vanished.\par
For several minutes he worked in silence. The exercise was simple enough to understand and humiliating to perform. Each time the image steadied, the smallest movement of hand or wrist sent it running. His bandaged thumb made the mirror difficult to grip. He compensated with the wrong fingers, then overcorrected, lost the beam entirely, found it again, and heard Va-Raedin breathe once through his nose. Not a sigh. Worse.\par
At last Sa-Kailan set the mirror flat and used the water.\par
He tipped the bowl until its surface caught the beam. The light returned upward, unstable but bright. The red point appeared on the wall and trembled there, alive to every small movement in the bowl. He held his breath until the surface steadied.\par
Va-Raedin studied the wall.\par
“Acceptable.”\par
Sa-Kailan felt the word as insult and reward together.\par
“Why was the water there?” Va-Raedin asked.\par
“To reflect.”\par
“The mirror reflects.”\par
“To reflect differently.”\par
“Everything reflects differently. Be precise.”\par
Sa-Kailan looked into the bowl. The surface still carried tiny ripples from his hand. The red point on the wall trembled in sympathy.\par
“The mirror keeps its angle,” he said. “Water answers after the hand has stopped touching it.”\par
Va-Raedin was silent long enough for the answer to become visible to Sa-Kailan himself.\par
“Better.”\par
The word struck him harder than acceptable had.\par
“Water remembers disturbance,” Va-Raedin said. “Not as Faith uses the word remember. Not grief, not mercy, not witness. Motion. A careless hand leaves a pattern after the hand is gone. Light striking the pattern tells you something happened.”\par
Sa-Kailan watched the trembling point.\par
“So what is unseen can be known by what it disturbs.”\par
“Sometimes. The water tells you disturbance occurred. It does not always tell you what disturbed it. Young minds love to turn traces into stories.”\par
Sa-Kailan almost answered, then stopped. Va-Raedin allowed the silence. The red point slowly steadied.\par
He thought of Ka-Elun walking the Labyrinth without witness. Faith said God had guided her. Light would ask what marks were left by feet, breath, oil, scraped stone, fear, water, time. Neither answer cancelled the other. That unsettled him.\par
Va-Raedin removed the mirror shard from the table.\par
“Again. Without this.”\par
“That was the main tool.”\par
“It was the easiest tool.”\par
Sa-Kailan failed six times. Twice he spilled water and was made to wipe the table with his sleeve because, Va-Raedin said, Balance would not be summoned to clean Light’s impatience. Once he moved the red tile too far and lost the point entirely. Once he got a red stain onto the wall, wide and inverted, more wound than mark.\par
Frustration entered his body. It shortened his breath, tightened his shoulders, made his fingers too strong. Va-Raedin did not raise his voice. He corrected as if time were one more instrument in the room.\par
“Too much angle. Too much hand. You are moving the glass because you dislike moving yourself. Stop chasing the image. Trace the path. Source, aperture, surface, eye. Again.”\par
By the seventh attempt, Sa-Kailan stopped trying to force the red point into place. He moved around the table until slit, water, glass, tile, and wall made a relation he could feel before he could state. The point appeared low on the far wall, smaller than before, but stable.\par
He did not step back this time.\par
He waited.\par
“Acceptable,” Va-Raedin said.\par
The word stung less.\par
They moved through three more chambers before first work began elsewhere in the Mountain. Sa-Kailan heard the day waking beyond Light: carts rolling toward the food galleries, voices near the washing pools, the metal rhythm of a gate being tested by Shield. In the eastern halls, morning did not arrive; it was admitted. Lamps narrowed or widened. Mirrors turned by increments. A corridor he had thought dark became pale. A pale one became useless.\par
In the third chamber, Va-Raedin made him cross a floor divided by light and shadow without stepping on lit stone. Sa-Kailan completed the exercise quickly and was made to repeat it after Va-Raedin changed nothing except his starting position. From the new angle, every certainty failed.\par
“Shadow is not darkness,” Va-Raedin said.\par
“An absence of light.”\par
“Too broad.”\par
“An interruption.”\par
“Better. By what?”\par
Sa-Kailan looked again. A lamp, a slit, a pillar, a wall, his body, Va-Raedin’s body, the raised edge of a stone tile. Each interruption produced a different shape.\par
“By whatever enters the path.”\par
“Of light, of view, of expectation. If a man stands between you and a door, you name the door hidden. The door has not changed. The man has entered your seeing.”\par
“That seems obvious.”\par
“Most useful things do until someone uses them against you.”\par
Va-Raedin crossed to a blackened wall panel and placed his hand flat upon it.\par
“Where is the exit?”\par
Sa-Kailan looked around. The entrance behind them was one answer, and therefore probably not the answer. Va-Raedin’s hand drew the eye to the panel. It was too large, too flat, no seam. Unless soot concealed the seam.\par
Sa-Kailan walked toward it.\par
Va-Raedin removed his hand.\par
The panel became simply black stone.\par
Sa-Kailan stopped.\par
“Why did you come here?” Va-Raedin asked.\par
“You drew attention to it.”\par
“Yes.”\par
“You made yourself the lamp.”\par
“No. I made myself the obstruction.”\par
Sa-Kailan looked back across the chamber. When Va-Raedin had stood at the panel, his body had blocked a reflection from a low mirror behind him. Without him there, the reflection crossed the room and struck the left arch, revealing that the floor beyond it did not continue. The false corridor had declared itself only when Va-Raedin moved away.\par
“You hid the falsehood by standing in front of the light.”\par
“I prevented you from seeing the condition that made the falsehood visible. That is not the same thing.”\par
“The distinctions are convenient.”\par
“They are survival.”\par
“To whom?”\par
The question came sharper than Sa-Kailan intended.\par
Va-Raedin looked at him.\par
“To anyone who does not wish to be governed by the nearest brightness.”\par
Water began moving through stone, quietly at first, then with a deeper pulse. Va-Raedin crossed to the true exit, concealed behind a slit Sa-Kailan had dismissed as ventilation.\par
“You want Light to be clean,” he said.\par
Sa-Kailan did not answer.\par
“You think because Faith clothes uncertainty in prayer, Balance in measure, Shield in necessity, and Justice in record, Light should stand apart and reveal things as they are.”\par
“Should it not?”\par
“No.”\par
The bluntness made Sa-Kailan turn.\par
“Light is not truth,” Va-Raedin said. “It is a condition under which truth may become visible, falsehood may become persuasive, or danger may become manageable. A lamp in the wrong place can blind more thoroughly than darkness.”\par
“Then what is our charge?”\par
“To know the difference before others suffer for it.”\par
Sa-Kailan wanted to ask how anyone could know the difference if Light itself could be arranged into lies. He did not ask. Va-Raedin would have approved the silence, which made keeping it mildly intolerable.\par
They passed through the ventilation slit into a corridor no wider than Sa-Kailan’s shoulders. Dull metal plates lined both walls, each angled slightly differently. The corridor bent twice, though reflections made it appear straight. When he looked back, he could no longer see the chamber they had left.\par
He stopped.\par
Va-Raedin continued five steps, then paused.\par
“Do not trust backward sight merely because it comforts you.”\par
“It shows the hall.”\par
“Which hall?”\par
Sa-Kailan approached the reflection carefully. The longer he studied it, the more wrongness appeared. The third doorway was reversed. The eastern mirror strips were too high. The cut circle in the floor was missing. It was not the first practice hall, then, but a likeness of it assembled from enough correct pieces to soothe the eye.\par
“Who would build a false view behind a candidate?” Sa-Kailan asked.\par
“Someone who knows candidates look back when uncertain.”\par
The corridor opened into a chamber lit by a single lamp placed on the floor. Around it stood six mirrors on slender bronze stems. Each mirror faced the lamp at a different angle. Above the far wall was an inscription cut in plain letters:\par
WHERE LIGHT FALLS IS ONLY THE FIRST QUESTION.\par
Sa-Kailan read it twice.\par
Va-Raedin adjusted one mirror by less than the width of a fingernail. Three patches shifted. One vanished. Another appeared where the wall had been dark.\par
“Place them so the lamp finds the door.”\par
“There is no door.”\par
“No visible door.”\par
Sa-Kailan exhaled slowly.\par
He began with the brightest patch and wasted several minutes. It was bait. The mirror producing it caught not the lamp directly, but a reflection of a reflection: brightness doubled, direction corrupted. When he moved it, the arrangement collapsed and the chamber became nearly dark. He reset it badly. Va-Raedin did not help.\par
He tried again, this time tracing the order: lamp to mirror, mirror to wall, wall to absence, absence to edge. Pride objected to crouching before sense accepted that height was another form of ignorance.\par
From the floor he saw that one mirror did not reflect the lamp at all. It reflected a gap between two brighter paths. He shifted it toward the gap. A pale line sharpened on the far wall. Another mirror caught it and sent it downward. A seam appeared at ankle height.\par
Not a door for a man. A panel.\par
He stood too quickly and struck his shoulder against one of the bronze stems. The mirror rang. Three lights jumped and vanished.\par
Va-Raedin looked at the trembling mirrors.\par
Sa-Kailan closed his eyes.\par
“You may begin again,” Va-Raedin said.\par
The worst punishment was when the room merely revealed what haste had done. He began again.\par
By the time the hidden panel opened, his knees ached from crouching and the linen on his thumb had darkened with new blood.\par
The panel revealed no passage, only a recess containing a small stone token marked with the sign of Light: not a lamp, as children drew it, but a narrow aperture cut through a circle.\par
“Take it,” Va-Raedin said.\par
Sa-Kailan took the token. It was heavier than it looked.\par
“What is it?”\par
“Proof that you opened the room.”\par
“For whom?”\par
“For you, if you are vain. For me, if you are honest.”\par
Sa-Kailan almost smiled.\par
“I opened it.”\par
“You did.”\par
“You expected me to fail.”\par
“I expected you to fail several times. You did not disappoint me.”\par
He did smile then, despite himself. Va-Raedin saw it and did not punish it.\par
They left the mirror chamber by a door that had been visible the whole time but unopenable until the ankle panel was released. That irritated Sa-Kailan more than hiddenness. Hidden doors admitted the intelligence of the seeker. Visible doors that refused to open until an unseen condition had been satisfied felt like mockery.\par
The next passage overlooked a lower work hall through a row of narrow slits. Balance clerks moved below with baskets and sealed jars. A woman carried folded cloth marked with green thread. Two boys pushed a cart of lamp oil under the supervision of a Light attendant, who checked each wax seal before allowing the cart beneath the eastern stores.\par
Nothing here, Sa-Kailan realized, belonged to Light alone. Oil came by Balance. Doors by Shield. Witness by Justice. Songs by Faith. Light made things visible, but visibility itself depended on other offices keeping their own disciplines.\par
Farther down, Faith singers rehearsed near a side arch. No full song reached the passage through the stone, only a low pressure in the rail beneath Sa-Kailan’s hand, as if the wall had accepted the note and returned its body without its words. Their mouths opened together. Breath entered rhythm. Rhythm entered stone.\par
A thin mist rose behind them.\par
Sa-Kailan leaned closer. The mist did not come from the floor. It emerged from a line of holes above the arch, so fine he had first taken them for natural pores in the stone. A hidden lamp turned the vapour pale at the exact moment the singers lifted their faces.\par
Va-Raedin allowed him to look.\par
After a while he said, “What do you see?”\par
“A rehearsal.”\par
“What else?”\par
“Water in the air.”\par
“What else?”\par
“Light placed to make the water visible.”\par
“What else?”\par
Below, one singer raised both hands. The mist thickened with the movement, as if the prayer itself had called it forth. The stone under Sa-Kailan’s palm answered a breath later than the bodies below.\par
“Timing,” Sa-Kailan said.\par
“What else?”\par
He hesitated.\par
“Sound moving where sight cannot follow.”\par
Va-Raedin nodded once.\par
Faith’s miracles were not yet open to him, not fully. Balance’s measures were not his to read. But a possibility entered him there, small and sharp. A prayer could be sincere and placed with precision. A blessing could be engineered without becoming false. Water could be counted in one chamber and appear as mercy in another. Sound could enter stone and return as something the eye would call impossible if the ear had not been trained first.\par
If people were comforted, was the comfort a lie? If God provided through valves, did the valve disprove God, or serve Him?\par
He did not ask. Va-Raedin would approve too many silences today.\par
The singers finished. The mist thinned. The stone released the last of its low answering hum. It had been beautifully done. Sa-Kailan hated how beautiful it was.\par
They returned to the upper hall by a route that did not feel like return. The first chamber had changed in their absence, or perhaps he had. The three false doors seemed less convincing. The mirrors seemed less like objects than decisions. He noticed a hairline aperture above the bronze door and wondered how he had missed it. He noticed that the open dark doorway smelled faintly of oil and therefore could not lead far. He noticed, with annoyance, that Va-Raedin noticed him noticing.\par
“Again,” Va-Raedin said.\par
“The doors?”\par
“The first test.”\par
“I know the answer.”\par
“That will make the failure more instructive.”\par
Va-Raedin placed him at the original entrance. The room appeared as it had: three doorways, one open, one closed, one half-open with a line of brightness. The hidden panel behind him remained invisible because the candidate entrance concealed its seam.\par
Sa-Kailan did not go to the doors. He looked for the worn circle, found it, stepped into it, turned, and pressed the wall.\par
Nothing.\par
He searched for the seam. It was gone.\par
Va-Raedin waited near the centre of the room.\par
Sa-Kailan kept his face still with effort. He looked again.\par
Doors, mirrors, circle. The angles had changed by almost nothing. No, the circle had changed. Its wear was false, darkened with rubbed dust. The true wear lay two steps behind the first mark, faint and half-concealed by reflected shadow. He moved to it.\par
The room shifted. The hidden panel was no longer behind him but to his left, a narrow seam disguised by darkness thrown from a mirror. He touched it. It opened.\par
This time he smiled.\par
“Better,” Va-Raedin said.\par
Not acceptable.\par
Better.\par
Sa-Kailan stored the word away like contraband.\par
The practice continued until hunger became useful. It sharpened him for a while, then made him clumsy. Va-Raedin seemed to know the precise point at which fatigue would begin to disguise itself as insight. He changed exercises before Sa-Kailan could become pleased with competence. A mirror task became a shadow task. A shadow task became a route task. A route task became a question about who had placed the first false choice.\par
Several times Sa-Kailan answered quickly and was wrong. Once he answered slowly and was also wrong, which annoyed him more. But not always. He began to feel the structure beneath the humiliations. A beam was not a line but a narrowed field. A mirror was not a tool but an instruction. A shadow could hide or reveal depending on whether one asked what cast it. A false path did not need to fool the wise forever. It needed only to delay the proud at the right moment.\par
Near the end of the session, Va-Raedin led him to a narrow balcony overlooking one of the external light shafts.\par
Cold air rose through it, dry and mineral, with a distant smell beneath that made Sa-Kailan think of desert stone though he had seen the desert only from protected places. The shaft dropped into darkness below and rose into darkness above. A beam fell through it, pale and almost solid, until it struck a suspended plate of polished metal. From that plate it divided through apertures cut into the Mountain. He could not count them all.\par
Va-Raedin rested his hands on the railing.\par
“When you were a child, what did Faith tell you about the Mountain’s light?”\par
“That God remembered us beneath stone.”\par
“Did you believe it?”\par
“Yes.”\par
“And now?”\par
Sa-Kailan looked at the beam, the plate, the apertures, the carved throats waiting to accept and divide what little Heaven gave them.\par
“I think someone had to build this.”\par
“That is not an answer.”\par
“No,” Sa-Kailan said. “But it is the question I have.”\par
Va-Raedin turned his head slightly. Approval, perhaps. Or warning. With Va-Raedin the two often wore the same face.\par
“Good. But understand the danger. Most children stop at miracle. Most fools stop at mechanism. Light permits neither laziness. A miracle may have a mechanism. A mechanism may still serve a miracle. What matters is who knows, who does not, and what each ignorance protects.”\par
Sa-Kailan watched the beam strike the suspended plate.\par
Around the edge of the metal the light misbehaved. He leaned forward. For a moment, thin bands appeared: pale, dark, pale, dark, too delicate to name before they vanished when he shifted his head.\par
“What was that?” he asked.\par
“What?”\par
“The edge. The light changed.”\par
“Edges do that.”\par
“How?”\par
Va-Raedin looked at the plate, then back to Sa-Kailan.\par
“You are not ready for every how.”\par
The answer frustrated him. Beneath frustration, something else took root. Not knowledge. Not yet. A disturbance continuing after the hand had gone.\par
They returned through the eastern corridor. Work had begun fully in the rest of the Mountain. Carts, footsteps, voices, gate chains, water: always water. Sa-Kailan felt tired enough that his body seemed only partly his, but his mind would not quiet. The hidden panel. The trembling red point. The mist behind Faith. The false backward view. The bands at the metal edge.\par
At the entrance to the candidate passage, Va-Raedin stopped.\par
“You will come again tomorrow.”\par
“Yes, Veil.”\par
“Earlier.”\par
Sa-Kailan looked at him before he could stop himself.\par
Va-Raedin almost smiled, which was not the same as smiling.\par
“You are young enough to think exhaustion is an insult. It is not. It is a solvent. It removes the lies you tell yourself about your discipline.”\par
Sa-Kailan flexed his bandaged thumb. The cloth had dried stiff with blood.\par
“Will I repeat the rooms?”\par
“No.”\par
“Then how do I prepare?”\par
“By noticing what you failed to notice today.”\par
“That is not preparation.”\par
“It is the only preparation Light respects.”\par
Va-Raedin turned to leave, then paused.\par
“Sa-Kailan.”\par
The prefix still startled him.\par
“Yes?”\par
“A Candidate learns where light falls.”\par
“A Veil learns who placed it,” Sa-Kailan said.\par
“And a Keeper?”\par
“Learns what must remain dark.”\par
Va-Raedin held his gaze.\par
“Do not rush to the third sentence because the first wounds your pride.”\par
Then he was gone down the eastern passage, his robe catching one pale shaft and losing it at the turn.\par
Sa-Kailan looked back toward the corridor that led to the Hall of Ceremony.\par
He had learned to open a hidden panel and move a red point onto a wall. It should have felt small. It did not.\par
Light was not truth. Light could be placed, bent, interrupted, multiplied, withheld. A visible thing could be answer or bait. A shadow could conceal a door or reveal the thing standing before it. Water could remember disturbance. A crowd could be made to see mercy when mist, lamp, valve, breath, stone, and prayer agreed at the right moment. A boy could think he had solved a room and discover he had obeyed the room’s first lie.\par
He returned to the candidate quarters by a longer route than necessary. Twice he stopped to examine lamps he had passed a hundred times before. Once he crouched to find where a floor reflection began and discovered only a servant staring at him with polite concern.\par
Near the washing arch, he watched water run along a shallow groove and saw the lamp above it break into pieces on the surface.\par
The Mountain had not changed. He had.\par
By the time he reached his chamber, hunger had become a dull ache and his knees hurt from stone. He unwrapped the bandage from his thumb. The cut had reopened slightly during the exercises, and dried blood marked the cloth in a dark crescent.\par
What is unseen can be known by what it disturbs.\par
Va-Raedin had not said it exactly that way. Sa-Kailan had.\par
He sat beside the basin until the water stilled. Then he leaned over it and looked at his reflection. A boy with a new prefix and an old hunger. Sa-Kailan, Candidate of Light. Not Kailan in the storage corridors behind Sheva. Not the child looking up at gravures and believing stone because stone had been cut. Candidate now. Named. Corrected. Allowed to fail in rooms built by men who had hidden lessons inside humiliation.\par
He touched the water with one finger.\par
The reflection broke. His face vanished into rings of motion. The rings met the edge of the basin and returned inward, crossing one another, making patterns he could neither keep nor understand. Beside the basin lay the stone token from the mirror chamber.\par
He picked it up, turned it once between finger and thumb, and let it slip from his hand into the water.\par
The token struck the side of the basin, turned, and settled near the bottom with the sign of Light facing upward.\par
Sa-Kailan reached for it, then stopped.\par
The circle cut through the token did not look like itself. Its edge thickened when he moved his head left. It narrowed when he moved right. The aperture seemed to slide away from the stone that held it.\par
He had seen such things all his life: a spoon bent in a cup, a hand made strange in a washing pool, lamp-lines trembling through jars of oil or water. Everyone had seen them. Seeing was not the same as noticing.\par
He lowered his face nearer to the surface.\par
The lamp above the basin curved along the water’s edge and broke itself against the rounded wall. The light did not only return. It entered, turned, thickened, and came back altered. The basin was not a mirror. Not exactly. Nor was it a lens, though it borrowed something from lenses. Water did not merely remember disturbance. It changed the path of sight.\par
A sudden brightness went through him that had nothing to do with the lamp.\par
He looked toward the door, half-expecting Va-Raedin to stand there already knowing, already prepared to reduce discovery into correction. But the chamber was empty. No Veil watched him. No Keeper named the lesson. No doctrine told him where to place his eye.\par
Perhaps everyone in Light knew this. Perhaps Va-Raedin had known it for years. Perhaps the discovery was childish, and tomorrow he would be made to feel the smallness of it.\par
But perhaps not everyone had looked at water in this way.\par
Sa-Kailan moved the token with two fingers and watched the sign slide one way while the stone moved another. Excitement sharpened until it felt almost indecent. He forgot his hunger. He forgot the ache in his knees. He forgot, for a few breaths, Sheva, the Hall, the cut in his thumb, and the humiliations of the morning.\par
Light entered water and did not remain obedient.\par
That was the thing. That was the crack.\par
He withdrew his hand and waited for the surface to calm. Light did not simply reveal hidden things. It could be delayed, diverted, bent by what it crossed. Water was not only something seen by light. Water could change what seeing meant.\par
He did not yet know what that meant.\par
That made it better.\par
\ornament
