
\chapter{Festival}

\noindent Sa-Kailan could not stop arranging the five lessons in his mind.\par
Reflection. Focus. Shadow. Turning. Threshold return.\par
He carried them everywhere, like stones touched in darkness to remember where the wall begins.\par
Lamps still had to be cleaned. Brackets still drifted. Smoke still gathered where smoke was permitted to gather. 
Faith still filled the Hall at appointed hours, and Balance still counted oil whether or not a boy believed the oil concealed revelation.\par
The order felt like a path.\par
That was the danger.\par

For the first few days after Va-Raedin ended the exercise, Sa-Kailan tried to obey. He did not speak of salted water to another Candidate. He did not ask Balance for salt outside assignment. He did not attempt threshold return in basin, cup, washing pool, oil jar, channel, or any other vessel whose ordinary use seemed suddenly beneath its possible one.\par

Obedience did not stop observation.\par

Every wet surface became an invitation. Every lamp touching water became an insult. Every polished bracket, oil jar, washing bowl, clay vessel, glass bead, damp groove, and accidental spill appeared to him as part of a sentence he was not yet permitted to read. He watched light enter clear water and return bent. He watched it move through oil and thicken. He watched it break along curved jars in the kitchen stores. He watched the measured channels near Balance’s lower rooms, where water moved under stone lips and through cuts designed for flow, not sight.\par

Most of it taught him nothing.\par

That too was information.\par

The Mountain did not waste complexity. If a thing had been made difficult, it was difficult because someone had needed it difficult. Threshold return did not belong to ordinary bowls. Turning did not become useful because a boy leaned over a basin and wanted revelation. The rooms of Light had required fixed source, fixed mark, fixed angle, fixed substance. Precision was not atmosphere. It was the condition under which the secret could exist.\par

Weeks passed.\par
The lessons did not continue. The season did.\par
Instruction became ordinary again, which Sa-Kailan found almost insulting. 
Lamps were cleaned, mirrors adjusted, wicks trimmed, brackets inspected, old smoke scraped from high vents.\par
Twice he spent an entire morning learning how an unclean mirror could reveal a sealed door by scattering what it should have returned whole.\par

“Boredom is useful,” Va-Raedin said on the third such morning.\par

“To whom?”\par

“To anyone who mistakes amazement for discipline.”\par

Sa-Kailan accepted correction and hated accepting it.\par

When no lesson was given, he searched where searching could be disguised as service. He studied the Hall of Ceremony until the gravure above the dais blurred from looking. Adam, Eve, the five children, Mountain, Oasis, forest, rivers, desert Threshold. He watched how lamps struck the carved hand of God differently at morning and evening. He counted where shadows gathered and failed to move. He stood in the public tiers and tried to determine whether any line from the Light alcove might pass through Faith’s bronze rail and into the lower stone.\par

The geometry refused him.\par

He searched the public routes from the eastern halls to the ceremony floor. He watched maintenance crews change lamps before dawn and after evening work. He learned that the Mountain’s light was less mysterious when seen from ladders, and more mysterious afterward. A miracle could become a bracket. A bracket could become a hidden bracket. A hidden bracket could become a question. A question could become a wall.\par

He tried, three times, to follow Va-Raedin.\par

The first attempt failed because Va-Raedin stopped walking without turning around and said, “Whatever you think you are doing, you are making noise with your left sandal.”\par

The second failed because Sa-Kailan followed a reflected glimpse of pale robe through two corridors and discovered, after twenty breathless minutes, that he had been stalking a hanging cloth placed near a vent to dry.\par

The third failed worst of all. He waited near the old Light gallery where he had seen the five markings, believing Va-Raedin would return there after evening instruction. He remained still until his legs ached, proud of his patience. Then a junior attendant entered, handed him a folded note, and left without comment.\par

The note read:\par

A hidden watcher should not cast a visible shadow.\par

It was not signed.\par

After that, Sa-Kailan stopped trying to follow Va-Raedin.\par

Then Ka-Syphiron summoned him.\par

The summons came through Va-Raedin at the end of a morning exercise so plain that Sa-Kailan had nearly fallen asleep standing. The other Candidates were dismissed. Their footsteps withdrew into the eastern passage. Va-Raedin waited until the room had emptied, then said,\par

“Ka-Syphiron will receive you before the second meal.”\par

Sa-Kailan’s stomach tightened.\par

“For correction?”\par

“If correction is required, you will learn.”\par

“That is not an answer.”\par

“No. It is preparation.”\par

The Keeper of Light received him in a room with no view and almost no object: a low table, an unlit lamp, three polished stones, one chair for the Keeper, no chair for the Candidate. Ka-Syphiron stood when Sa-Kailan entered. That frightened him more than if the old man had remained seated.\par

Ka-Syphiron had the beauty of something preserved beyond kindness. Age had narrowed him without weakening the impression of design. His hair was white, his hands fine, his face calm enough to make emotion seem like a defect in others. He wore no ceremonial robe, only pale cloth with a thread of gold at the throat. Yet he looked more like Keeper there than he had in the Hall, because nothing around him competed for meaning.\par

“Sa-Kailan,” he said.\par

“Keeper.”\par

“Va-Raedin tells me you are quick.”\par

Sa-Kailan did not know whether to accept the word or defend against it.\par

“I try to learn.”\par

“Trying is common. Learning is less so. Understanding is rarer. Wisdom is rarer still.”\par

“Yes, Keeper.”\par

Ka-Syphiron gestured to a stone before him. Not to kneel. To stand in a place chosen by another.\par

Sa-Kailan stepped onto it.\par

“Tell me what a Veil of Light is.”\par

The answer came easily, and therefore badly.\par

“A Veil learns who placed the light.”\par

“That is doctrine. I asked what a Veil is.”\par

Sa-Kailan hesitated.\par

“The Keeper’s successor.”\par

“Sometimes.”\par

“The one permitted deeper Light knowledge.”\par

“Too vague.”\par

“The one who has crossed beyond Candidate instruction.”\par

“By what act?”\par

Sa-Kailan looked at the unlit lamp on the table. His mouth had gone dry.\par

“By finding the Veil’s Chamber.”\par

“Yes.”\par

The simplicity of the word entered him like cold water.\par

“The title is not given for cleverness,” Ka-Syphiron said. “Not for age, not for loyalty, not for hunger, not for the admiration of one’s teacher, not for the impatience of one’s own mind. A Candidate of Light becomes Veil of Light by finding the Veil’s Chamber. Until then, he is Candidate, whatever he believes he has understood.”\par

Sa-Kailan felt the words arrange themselves around him like bars.\par

“And if he never finds it?”\par

“Then he never becomes Veil.”\par

“Even if the Keeper dies?”\par

“The Mountain does not owe a man the office he failed to reach.”\par

Sa-Kailan thought of Va-Raedin. Had he found the chamber long ago? Of course he had. That was what being Veil meant. But had Ka-Syphiron guided him, or withheld the same mercy now being withheld from Sa-Kailan? Had Va-Raedin stood on this stone too, young and trying not to look too hungry?\par

“Does the Chamber teach the next path?” Sa-Kailan asked.\par

Ka-Syphiron’s face remained calm.\par

“You are not yet entitled to next paths.”\par

“I only meant—”\par

“I know what you meant. That is why I answered the sentence beneath the one you spoke.”\par

The old man crossed to the table and touched the unlit lamp.\par

“You have the flaw of many gifted Candidates. You think a locked door is an insult addressed personally to you. Sometimes a locked door is simply a mercy you have not earned the right to resent.”\par

Sa-Kailan lowered his eyes.\par

“Look at me.”\par

He looked.\par

“You are seen,” the Keeper said. “Not celebrated. Seen. There is a difference. Va-Raedin sees speed in you. I see appetite. Speed may serve Light. Appetite must be disciplined before it convinces itself that hunger is revelation.”\par

The younger brother of the Book of Blood came into Sa-Kailan’s mind without warning: small in body, cunning in heart, watching strength loved by others until envy entered him as a serpent entereth the grass. The old story had become too useful against him.\par

“You may go,” Ka-Syphiron said.\par

That was all. No clue. No riddle. No test. No mercy hidden in grammar. Only the rule, placed where he could no longer pretend not to know it.\par

A Candidate of Light became Veil by finding the Veil’s Chamber.\par

For several days after the interview, Sa-Kailan avoided the Hall of Ceremony because he did not want to admit to himself that he was going there to search rather than pray.\par

Then he went anyway.\par

The chamber received him indifferently. One restless Candidate of Light did not disturb stone laid down before his great-grandfathers were named.\par

He began with architecture and failed. Then he turned to scripture. Not because humility had opened him to God. Because scripture remained.\par

Faith had preserved the oldest words. Faith knew chambers Light did not enter, or claimed to. Faith’s rites opened doors no lamp could open.\par

He did not seek Ka-Elun. A Keeper was too high and too newly enthroned. Instead he requested audience with Va-Maera, Veil of Faith, through the public petition line beside the Hall of Ceremony. Va-Maera was older than Ka-Elun by several years, which made her position strange to anyone outside Faith and normal to everyone within it. Faith did not always move by age. It moved by devotion, trial, burden, and the kinds of obedience other Kahir found difficult to measure.\par

She received him in a narrow chamber outside the Hall of Prayers.\par

Not inside it.\par

That distinction was the first answer.\par

The chamber had no ornament except a covered bowl of water and a hanging cord of five knots. The walls were warm. No lamp was visible, yet the room was softly lit from somewhere above. Va-Maera stood throughout. She wore white, but a heavier white than the attendants, layered and severe, her hair fully covered, her face unpainted, her eyes dark and steady.\par

Sa-Kailan had expected devotion and found administration sharpened by holiness. He had expected softness and found a blade wrapped in linen.\par

“You request access to the Hall of Prayers,” she said.\par

“Yes, Veil.”\par

“Why?”\par

“To study the Covenant.”\par

“The Covenant is read publicly.”\par

“Not all of it.”\par

“No.”\par

“I am Candidate of Light.”\par

“I see your robe.”\par

“There may be passages concerning Light that are not read in public.”\par

“There are passages concerning many things that are not read in public.”\par

“I seek the Veil’s Chamber.”\par

The words sounded too naked once spoken.\par

Va-Maera did not react.\par

“Then seek it in Light.”\par

“I have. I am.”\par

“Then continue.”\par

He had prepared careful sentences. They had seemed subtle when rehearsed in corridors.\par

“Faith works closely with Light,” he said. “In ceremony. In mist. In timing. In the way certain signs are made visible. I have seen enough to know that what people call miracle may depend on arrangements not everyone understands.”\par

The air changed.\par

Not visibly. Not with any theatrical sign. But the room became suddenly smaller, and Sa-Kailan knew he had misjudged the distance between allusion and threat.\par

Va-Maera took one step toward him.\par

“Say that again,” she said.\par

He could not.\par

“Good,” she said. “Some doors remain closed because the fool outside them still has one living instinct.”\par

“I did not mean—”\par

“You meant enough.”\par

He swallowed.\par

“You come to Faith wearing Light’s new hunger, and you think because you have seen mist in a prayer hall, because a lamp and a valve once agreed before your clever eyes, that you now stand above the people who kneel. You think you have found a crack in Faith. You have found a crack in yourself.”\par

The words struck harder because she did not raise her voice.\par

“I was asking for access,” he said, though the sentence failed while he spoke it.\par

“No. You were suggesting that Faith should fear what you might know. Listen carefully, Sa-Kailan of Light. Such suggestions do not endanger Faith first. They endanger you first. If danger follows those words, it will find your door before it reaches mine.”\par

He felt cold beneath his robe.\par

Va-Maera came closer still.\par

“You have learned small tricks of perception and mistaken them for height. Mist is not Faith. Timing is not Faith. A hidden channel is not Faith. Do you think the sacred is diminished because it consents to pass through matter? Do you think God is disproved by a valve? You are not the first bright child to discover that blessing may have mechanism. You may be the first foolish enough to speak of it as leverage in a house of prayer.”\par

“I was not threatening.”\par

“No. You were too clumsy to threaten properly. That does not make you innocent.”\par

Sa-Kailan lowered his eyes. He could feel his pulse in his throat. Va-Maera’s voice softened further, which made it worse.\par

“We speak with archangels of obedience and demons of appetite. We preserve names that may not be spoken where children sleep. We know prayers for birth, madness, treason, plague, thirst, deformity, despair, and the moment before a condemned man understands that mercy will not interrupt justice. We bless marriages and executions. We comfort mothers whose children are given to roles that will consume them. We keep fasts that would break your pride before your body. We know punishments unknown to your little training rooms and consolations beyond your imagination. And you come to me with mist.”\par

He could not move.\par

“Look at me.”\par

He looked.\par

“You understand almost nothing,” she said. “That is not your sin. Your sin is that you have begun to suspect this and decided to compensate by becoming dangerous. You are making a chain of mistakes. If you continue, you will not become a god of hidden knowledge. You will become an example spoken of carefully to other gifted boys.”\par

The terror was not that she would call Shield, or Justice, or Ka-Elun. The terror was that she had made consequence feel larger than any of them. Sa-Kailan believed suddenly, completely, that Faith could destroy him without touching him. It could name his hunger incorrectly in the right chamber, place his words in the wrong sequence before Justice, turn his discoveries into contagion, his cleverness into evidence, his ambition into spiritual deformity.\par

Whether all of that was true did not matter.\par

It was true enough in the room.\par

He folded.\par

“I withdraw the request,” he said.\par

“Of course you do.”\par

“I apologise, Veil.”\par

“No. You submit. Apology is what people offer when they still wish to manage the injury.”\par

He could barely speak.\par

“I submit.”\par

Va-Maera stepped back.\par

“You may listen in the Hall of Ceremony like any child, elder, worker, mourner, or sinner. You may hear what is publicly read. You may pray where prayer is permitted. You may not enter the Hall of Prayers. You may not request private scripture. You may not speak again of Faith’s arrangements unless Faith first asks you what you think you have seen.”\par

“Yes, Veil.”\par

“And Sa-Kailan?”\par

“Yes?”\par

“If you truly seek Light, learn first that not every brightness belongs to you.”\par

She dismissed him with a movement of two fingers.\par

He left the chamber shaking and hated himself for the shaking only after he had gone three corridors and discovered that no one had followed.\par

At first shame accompanied him. He imagined Va-Sheva laughing, though he knew she would not. Worst of all, he imagined Ka-Syphiron looking at him with calm recognition, as if appetite had behaved exactly as expected.\par

But the Hall was public, and public devotion protected him from suspicion. The tribe approved of those who drew closer to God. An old woman touched his sleeve one morning and said his mother must be proud. A Faith attendant nodded to him when he arrived before a water thanksgiving. Children moved aside on the benches because a Candidate of Light sitting through recitation seemed serious rather than desperate.\par

No one asked whether he came to pray or to search.\par

Months passed that way.\par

He heard ration prayers, birth blessings, judgement remembrances, harvest petitions, mourning chants, water thanks, lamp rites, and fragments of the Covenant repeated until repetition became excavation. Lines that had once seemed ornamental began to catch on hidden hooks in his mind.\par

The Genesis chapter came often during those months, sometimes in fragments, sometimes nearly whole, and once, during the long preparation for the seasonal Festival, in full.\par

Ka-Elun herself read it that day from the central dais. Her Keeper’s veil fell cleanly around her face, covering the pilgrim scars that had once made Sa-Kailan stare. The Hall was full but not ceremonial-full; workers had been released in shifts, children brought by teaching groups, Balance clerks seated with tablets because even long scripture had to be scheduled against labour. Sa-Kailan sat on the eastern tier and listened with the attention of someone searching a wall for a seam.\par

Ka-Elun began.\par

\begin{scripture}
In the beginning God created the heaven and the earth. And the earth was without form, and void; and darkness was upon the face of the deep. And the Spirit of God moved upon the face of the waters. And God said, Let there be light: and there was light. And God saw it was good: and God divided the light from the darkness. And God called the light Day, and the darkness he called Night. And the evening and the morning were the first day.
\end{scripture}

\begin{scripture}
And God said, Let there be a division in the midst of the deep, and let it divide the heights from the depths. And God made the sky above, and the ground beneath, and the pit below the ground: and it was so. And God called the sky Heaven, and the ground Earth, and the pit He called Hell. And God set the Earth between Heaven and Hell, that the living should walk between the eye above and the fire beneath. And the evening and the morning were the second day.
\end{scripture}

\begin{scripture}
And God said, Let the waters under the heaven be gathered together unto one place, and let the dry land rise above them: and it was so. And God gathered the waters beneath the earth, and hid them in the hollow places of the deep; and the dry land appeared, and from the dry land He raised the Mountain. And God said, Let the waters bring forth a spring within the shadow of the Mountain, and let the spring be hidden from the eye of heaven: and it was so. And the spring came forth in the shadow of the Mountain, and there was grass around it, and trees for shade, and the green herb yielding seed, and the fruit tree yielding fruit after his kind whose seed is in itself, upon the Earth. And God called the spring in the shadow of the Mountain Oasis: and God saw that it was good. And God said, Let there be a forest round about the Mountain and the Oasis, and let rivers run through the forest, and let the trees be many, and let their branches give shade unto the living: and it was so. And the forest grew round about the Mountain, and the rivers ran through it, and the waters thereof gave life to every root and beast and creeping thing. And the evening and the morning were the third day.
\end{scripture}

\begin{scripture}
And God said, Let there be a desert round about the forest and the rivers, to divide the forest from the lands beyond, and to set thirst and wandering before the stranger: and it was so. And God called the desert the Threshold; and He set heat upon it by day, and cold upon it by night, and mirage upon its face, that no man should cross it lightly. And God said unto the living, The Mountain is refuge, and the Oasis is water, and the forest is plenty, and the rivers are life; but beyond the forest thou shalt not wander, for beyond the forest lieth thirst, and the eye of Heaven, and the forgetting of My shadow. And God saw that the hidden place was good. And the evening and the morning were the fourth day.
\end{scripture}

\begin{scripture}
And God said, Let us make man in our image, after our likeness: and let him walk upon the Earth, between the eye above and the fire beneath. And God formed Adam from the dust of the Mountain, and breathed into him the breath of life; and Adam became a living soul. And God said, It is not good that man should be alone in the shadow of the Mountain: and God caused a deep sleep to fall upon Adam, and he slept; and from his side God made Eve, and brought her unto the Oasis. And God blessed them, and God said unto them, Be fruitful, and multiply, and keep the Oasis, and dwell in the shadow of the Mountain; and eat of the fruit of the forest, and drink of the rivers thereof, but wander not beyond the forest. And Eve bare five children unto Adam; and the first opened his eyes unto the light, and the second drank from the waters, and the third laid his hand upon the stone, and the fourth took up the blade, and the fifth lifted his voice unto Heaven. And God saw Adam, and Eve, and their five children in the Oasis, beneath the shadow of the Mountain; and God saw that it was very good. And the evening and the morning were the fifth day.
\end{scripture}

The Hall answered not with words but with bowed heads.\par

Sa-Kailan remained still.\par

He had heard those lines before, scattered across childhood and ceremony, but never with that degree of attention and never with failure sitting so close beside him. In childhood, Genesis had been the first story, and first stories were too large to be questioned because they arrived before questions. Heaven, Earth, Hell. The waters beneath. The Mountain raised above them. The spring hidden from the eye of Heaven. The forest, the rivers, the desert Threshold. Adam, Eve, and the five children under the shadow of the Mountain.\par

Now every line seemed to contain a possible seam.\par

Light divided from darkness. Heaven divided from Earth, and Earth from Hell. Waters gathered beneath the ground. A spring hidden in shadow. Desert set as Threshold. The eye of Heaven. The forgetting of God’s shadow. Five children, but not yet five charges. No Blood, no Beast, no Cataclysm, no Covenant named openly in that chapter. Only beginnings. Only the world arranged before law had explained itself.\par

A later book could command.\par

A beginning could only imply.\par

He searched the reading for the Veil’s Chamber and found too much. The first opened his eyes unto the light. The second drank from the waters. The third laid his hand upon the stone. The fourth took up the blade. The fifth lifted his voice unto Heaven. Light, water, stone, blade, voice. The sequence was not enough because it was everywhere.\par

He needed an occasion.\par

That thought did not come to him immediately. It arrived slowly, through failure.\par

He became familiar with the Hall, and familiarity did not comfort him. He heard Genesis in pieces, the Book of Blood in fragments, the short water prayers of Balance, the threshold blessings of Shield, the correction hymns of Justice, and the lamp petitions of Light. He learned that scripture did not unlock itself by being heard often. It became more layered, not less.\par

No one troubled a young Candidate who sat through readings with lowered eyes. Faith approved of him. Balance approved of anyone who sat still without consuming resources. Justice saw nothing irregular in attendance. Shield did not care where a boy prayed so long as he remained where prayer permitted him. Even Va-Raedin, if he knew, gave no sign.\par

Perhaps that was the correction.\par

Perhaps the answer did not come because there was no answer to be taken.\par

The five lessons remained inside him like a sequence waiting for a location. Reflection, focus, shadow, turning, threshold return. He saw the first three everywhere. The Hall of Ceremony was full of them. The central gravure accepted certain lamps and rejected others. Bronze rails caught faces at the edge of sight. Shadows in the upper vaults concealed seams, vents, old repairs, and once, to his brief excitement, a narrow service niche that led only to a soot chamber.\par

But turning and threshold return did not belong to ordinary prayer. They required substance, angle, and precision. Water placed where light had reason to enter. A beam with enough discipline to be bent. A surface struck at the right condition.\par

The Hall had water in basins, water in bowls, water in ritual vessels, but none of it sat in the path of a serious beam.\par

The Mountain, he began to think, was withholding not the answer but the occasion.\par

Then came the Festival of Equal Light.\par

It arrived at the season when day and night were said to stand before Justice and neither could accuse the other of theft. Balance used it to review stores, seed, water, births, deaths, illness, labour loss, animal increase, and what could be spared for planting. Faith called it older than accounting: the day on which God’s first division of light and darkness was remembered without fear of either. Shield treated the gathering as a density problem. Too many bodies in one Hall, too many exits, too much attention turned toward one place. Justice opened tablets before the first lamp was lowered.\par

Light prepared the Hall in ways Sa-Kailan had never been permitted to observe so closely before.\par

Upper apertures were cleaned with white cord. Polished inserts in the central gravure were uncovered and wiped with cloths counted before use and after. The lamps were not raised but lowered. Bronze lips beneath the carved waters were checked by two attendants, then sealed again. Three old vents above the dais were opened under Justice witness and marked in the record immediately afterward.\par

The hidden people did not speak of the heavens as wanderers did.\par

But they knew seasons because seasons measured survival.\par

The Festival was not curiosity. It was order.\par

The Hall filled before the second morning silence. Sa-Kailan sat on the eastern tier, where he had sat for the Ascension, though he no longer felt like the boy who had stood there with a new bandage and a new name. Va-Raedin stood below with the attendants of Light. Ka-Syphiron sat in the eastern alcove, remote and pale, his face unreadable above folded hands. Ka-Elun stood beneath the central gravure. Balance had placed baskets of seed, preserved fruit, and sealed jars before the dais. Faith had laid the Genesis bark beside the festival prayers. Shield held the upper doors. Justice sat with tablets open.\par

The lamps were dimmed one by one.\par

A murmur passed through the Hall. Not fear. Anticipation disciplined by rite.\par

Ka-Elun lifted both hands.\par

\begin{scripture}
“God who divided the light from the darkness, measure us.” The Hall answered.
\end{scripture}

“Measure us in mercy.”\par

\begin{scripture}
“God who set Earth between Heaven and Hell, steady us.”
\end{scripture}

“Steady us in shadow.”\par

\begin{scripture}
“God who hid the spring beneath the Mountain’s hand, keep us.”
\end{scripture}

“Keep us from wandering.”\par

The incantation continued, not sung exactly, but carried in a rhythm that made breath part of architecture. Faith named the first day and the second, the waters beneath, the Mountain raised, the Oasis hidden, the forest of plenty, the desert Threshold, Adam and Eve, and the five children in the first shadow.\par

Sa-Kailan listened only partly.\par

Something in Light’s preparations had begun to disturb him. The dimmed lamps. The unsealed upper aperture. The polished fragments within the gravure. The old vents opened under witness. A ceremony about equal light held in a Hall designed to receive controlled darkness.\par

Then the sun entered.\par

At first it was only a pale line high above the central gravure, so thin Sa-Kailan almost missed it. It touched the carved division between Day and Night. The Hall did not move. Ka-Elun continued speaking. The line brightened. It slid, not downward exactly, but across the stone at an angle no lamp inside the Hall could have produced.\par

It passed over the carved waters beneath the Earth, struck the raised Mountain, and flashed.\par

Sa-Kailan stopped breathing.\par

The fragment threw the light downward through another cut, then toward the lower structure beneath the dais. Dust lit within it. The carved Oasis shone. The five children emerged from darkness as if newly made.\par

The people bowed lower.\par

Some wept.\par

Sa-Kailan did not bow.\par

The beam struck, turned, narrowed, and vanished into the Mountain.\par

Not faded.\par

Not scattered.\par

Vanished.\par

A path that did not exist yesterday. A path that would not exist tomorrow.\par

His mind arranged the five lessons so quickly it frightened him. Reflection. Focus. Shadow. Turning. Threshold. But not in training rooms. Not under ordinary lamps. A fixed beam. A rare day. A public rite. A Festival of Equal Light.\par

The angle could not be changed because Heaven itself had placed it.\par

The source could not be moved.\par

The path could not be found by searching on the wrong day, no matter how clever the searcher. Months of failure had not proven that nothing existed. They had proven only that he had been looking when the door was not there.\par

Unless one waited for Festival.\par

Unless one chased the beam now.\par

Ka-Elun’s voice continued below, solemn and beautiful, but Sa-Kailan barely heard it. Va-Maera’s warning returned: not every brightness belongs to you. Ka-Syphiron’s rule returned: the Chamber must be found. Va-Raedin’s prohibitions returned: no salt, no vessels, no threshold return outside instruction.\par

The beam trembled, still alive in the hidden structure.\par

Sa-Kailan rose before he had decided to rise.\par

No one near him noticed at first. The Hall was bowed in prayer. The lamps were dim. All eyes had been trained toward holiness, not toward one Candidate on the eastern tier moving too carefully between bodies. He stepped down once, then again, keeping to the shadow between two bench lines.\par

He did not know where the beam went.\par

But for the first time in months, the Mountain had shown him where to begin.\par

\ornament
