\typesetchapter{Chapter 1}{Ascension}

\noindent The Hall of Ceremony had been warmed for three days.\par

The Mountain had its own mercy, measured and rationed like everything else. Warmth came by permission. Light came by permission. Water came by permission. Even silence, on days like this, arrived through offices before it reached the people.\par

Kailan had been standing for longer than he wished to admit. His feet hurt inside the ceremonial sandals. The collar of his grey robe scratched at his throat. He had been told not to move. He had also been told that a body trained to stillness would eventually forget discomfort. His body had not been properly instructed.\par

“Beyond the forest thou shalt not wander,” the child beside him muttered.\par

Kailan did not turn. He knew better than to be seen turning during an Ascension. In the Hall, every small movement travelled. A cough rose from the lower galleries and returned as a softened echo from the carved vault. A sleeve brushed stone. A wooden bead clicked against a prayer cord. The old lamps hissed where thin oil met braided fibre, and from the high vents came the faint mineral smell of water moving somewhere behind the walls.\par

The Hall of Ceremony was the one place every child in the Mountain knew before he understood why it mattered. Its ceiling rose in five sloped vaults, each one cut toward a different Kahir alcove. Balance to the west, where a sheet of water fell behind a bronze grille and disappeared into measured channels below the floor. Shield to the north, where the arch was plain and unadorned, and the wardens stood facing the crowd as much as the dais. Justice to the south, where no flame was permitted and the walls were pale as bone. Light to the east, where mirrors caught the lamps and persuaded the stone to be less heavy than it was. Faith at the centre, beneath the highest carved panel, where God raised the Mountain and hid the spring from the eye of Heaven.\par

The gravure above the central dais was older than any name Kailan had ever heard spoken aloud. It was long and dark and so densely carved that from the tiers it looked almost like a bruise in the wall. Children were taught its shape before its meaning. The Mountain, the Oasis, the forest and rivers, the desert Threshold. Adam and Eve beneath the shadow. Five children. Five future charges. The figures were small, crowded together under the enormous hand of God, while the desert beyond them was empty except for heat-lines and beasts with ribs showing through their skin.\par

When Kailan had been younger, he had believed the gravure was simply history. Now, at fifteen, he had begun to suspect that history was never simple in a place where every wall had been carved by someone with permission to omit things.\par

A chime passed through the Hall.\par

It did not sound struck. It seemed to arrive already shaped, one clear note from somewhere above Faith, caught by the vaults and returned in five thinner echoes. Kailan had once asked why the Hall was built to make old voices larger and young voices smaller. His mother had told him that questions about Faith were safest when asked in Faith’s own rooms, which meant not at all.\par

The whispering stopped.\par

At the centre of the dais, three women of Faith entered barefoot, carrying shallow bowls of water. Their robes were white, but not the pure white of children’s ceremony. Faith-white was always a little stained by use: oil at the cuff, ash near the hem, water marks where hands had dried quickly after blessing. They walked in time, not quite music, not quite marching. Their feet touched the stone at intervals that made the Hall answer softly beneath them.\par

Behind them came the old Keeper of Faith.\par

A murmur passed through the Hall despite the discipline of the gathering. Kailan heard it, felt it more than heard it, a collective intake of pity and reverence. Ka-Mereth had not been seen in public for seven months. Some had said she no longer knew the order of the prayers. Some had said her voice had gone into God before the rest of her. Some had said she had already died and that Faith was waiting for the proper ceremony to announce what everyone feared.\par

She had not died.\par

Two attendants carried her in a low chair of dark wood. She had been small for as long as Kailan could remember, but now she looked as if the bones inside her had become too tired to hold their arrangement. Her face had narrowed to skin, eyes, and breath. The white veil that covered her hair had slipped to one side, exposing a scalp as fine and pale as the underside of a root. Her hands lay open on her lap. One trembled, the movement so slight that it might have been the flicker of lamplight.\par

The people lowered their heads. Kailan lowered his with them, but not before he saw the woman walking behind the chair.\par

Va-Elun of Faith. Not Va-Elun for much longer.\par

One of Faith’s two Veils, and the one who had walked the last path, was not old. That alone made the ceremony feel strange. She was perhaps thirty, perhaps less; with Faith, age was difficult to read because its women learned early how to empty their faces of appetite. She had the stillness of someone who had spent many years being watched while pretending not to know it. Her robe was white, but over it she wore a narrow mantle of undyed wool, rough against the ceremony’s polish. A pilgrim’s mantle. Its hem was torn in several places. One sleeve had been patched with darker cloth. At her throat, where the collar opened, Kailan saw two raw lines crossing the skin.\par

He stared at them until Va-Raedin touched two fingers to the back of his wrist.\par

Not hard. Not as reprimand. Enough.\par

Kailan lowered his eyes. Va-Raedin stood beside the Candidates of Light, as he always stood: without wasted movement, without visible impatience, without the small discomforts other men permitted themselves in public. He was the Veil of Light, and there were people in the Mountain who said he had never once been surprised by anything that chose to happen where lamps were lit.\par

Kailan did not believe that. He had seen Va-Raedin pause once before a mirrored door, barely a breath, barely a hesitation, but enough to prove that even he sometimes needed the world to explain itself.\par

At the foot of the dais, the bowls were placed on five low stones. Balance water, Faith water, Shield water, Light water, Justice water. A child might have asked how water differed by Kahir. An adult knew better. Water was never only water once hands had named it.\par

The Keeper of Faith was carried to the central seat. Faith said comfort weakened prayer. Balance said the old seat wasted less wood. Shield said nothing, which usually meant Shield thought both answers useful. Va-Elun knelt before her and the Hall waited.\par

For a while the old Keeper did not move. Her breath came through her mouth, dry and narrow. One attendant bent toward her, but the old woman lifted two fingers, and the attendant withdrew. When Ka-Mereth spoke, the voice did not seem strong enough to leave her body. Yet the Hall had been shaped for such things. The sound travelled up the cut stone, thinned but clear.\par

“What did God give the hidden people?”\par

“Stone, water, and shadow.”\par

The response came from every tier.\par

“What did God refuse the hidden people?”\par

“Sky, wandering, and the open world.”\par

“What is kept beneath the Mountain?”\par

“The Temple.”\par

“What is kept beneath the Temple?”\par

No one answered.\par

That was the old rhythm. Even the smallest children knew it. Some questions were asked so that silence could learn its place.\par

Ka-Mereth’s head bent forward, and for one terrible moment Kailan thought she had died between one breath and the next. Then she lifted her hand and touched Va-Elun’s forehead.\par

“You walked,” she said.\par

Va-Elun did not raise her face.\par

“I walked.”\par

“You entered without witness.”\par

“I entered without witness.”\par

“You took no lamp beyond the third descent.”\par

“I took no lamp beyond the third descent.”\par

“You drank no water after the fifth silence.”\par

“I drank no water after the fifth silence.”\par

“You passed the doors that opened.”\par

“I passed them.”\par

“You knelt where there was no altar.”\par

“I knelt.”\par

“And when the path returned upon itself?”\par

“I went on.”\par

The old Keeper’s hand trembled harder now. Her fingers slid from Va-Elun’s forehead to her cheek, almost like a mother touching a fevered child.\par

“What guided you?”\par

“God alone.”\par

“What did you find?”\par

“The Chamber of the Last Prayer.”\par

Kailan felt the words in his ribs.\par

The Chamber of the Last Prayer. He had heard the phrase before, of course. Everyone had. It appeared in the children’s version of the fifth text, when God commanded the people to keep the Mountain, the Oasis, and the Temple. It appeared in the burial rite for Keepers of Faith. It appeared in old women’s warnings when children lied about finishing prayers.\par

But to hear it said in the Hall, not as metaphor but as destination, changed it. Va-Elun had walked somewhere. Not simply endured sorrow, not simply prayed in darkness. She had entered the Labyrinth by the path reserved for Faith and had come out with the right to be more than witness.\par

Kailan glanced toward the Shield alcove.\par

He found Va-Sheva at once. He always found her quickly. She stood beside the northern arch with the other officers of Shield, four steps below the Keeper and two steps above the armed wardens. She wore dark green-black leather over a sleeveless robe, her hair bound high and tight, a short blade at her hip, a longer one across her back. Shield did not decorate itself. Its beauty came from function, which made looking at her feel dangerous.\par

She was nineteen, perhaps twenty. Old enough that the women beside her treated her as one of them. Old enough that the boys near Kailan watched her with the careful hopelessness reserved for things already claimed by adulthood. When she looked in his direction, her eyes passed over him, paused because recognition required it, and softened by a fraction. Not enough. Never enough.\par

To her, he knew, he was still partly the boy who had followed her through service tunnels when she was already training with wooden knives. He was still the child who had cut his hand on lamp glass and denied crying until she laughed. He was still becoming, while she had already been taken by duty and hardened into shape. Wanting her felt less like desire than trespass upon a rank he had not earned.\par

She looked away before he did. That was mercy, perhaps. Or indifference.\par

On the dais, the rite continued. The five Keepers, or those who served in their place where age or office required, came forward to witness Faith’s transfer.\par

Balance came first: Ka-Leth of Balance, broad, heavy, her hair grey at the temples, her hands stained faintly green from ledger dye and plant work. Two attendants followed her carrying a narrow measuring rod and a small copper vessel marked by rings. She poured a thread of water from her bowl into the central basin and waited until it reached the first line.\par

“Let prayer not empty the cistern,” she said.\par

Then Shield: Ka-Dhavar, tall, unsmiling, with scars crossing one hand. He did not pour. Shield placed a sealed iron ring beside the basin. As he did, the wardens at the northern arch changed stance together, one looking inward, one outward, one upward toward the gallery exits. No one in the Hall spoke of it. Everyone saw it.\par

“Let no threshold open because a holy tongue grows soft,” Ka-Dhavar said.\par

Justice came next, pale-robed, with no ornament but a strip of black cord at the wrist. Ka-Orun of Justice held a small tablet of polished stone and read the transfer into the record before it had fully happened.\par

That had always unsettled Kailan. Justice wrote events while they were still becoming them. The order mattered. A word placed before a breath, a witness named before a death, a silence recorded before anyone tried to fill it. Justice did not trust memory once desire had touched it.\par

“Let what is witnessed be kept,” Ka-Orun said.\par

“Let what is kept be answerable.”\par

Then Light came forward. Ka-Syphiron, Keeper of Light, was old but not diminished like Ka-Mereth. His white-gold robe caught the lamps around him and returned them with irritating grace. He was a beautiful old man in the way polished bone might be beautiful. Behind him stood Raedin, still as a blade laid flat.\par

Ka-Syphiron lifted a lamp and held it before Va-Elun’s bowed head.\par

“Let the path be seen only by the one appointed to walk it.”\par

Kailan leaned slightly forward before he could stop himself. The lamp was plain, no larger than a worker’s hand-lamp, but the flame inside it burned blue at the root. A trick of oil, perhaps, or air, or some secret of Light he had not yet been given. The flame bent toward Va-Elun though there was no visible draft.\par

Raedin’s voice came beside him, low enough that only the Light candidates heard.\par

“Do not admire the flame. Ask where the air enters.”\par

Kailan felt heat in his face.\par

“I was not admiring it.”\par

“No,” Va-Raedin said. “You were believing it.”\par

That stung because it was true, and because Va-Raedin had not bothered to make it sound cruel.\par

Faith’s attendants removed the old Keeper’s veil. Briefly Ka-Mereth’s bare head was visible to everyone. Thin hair, white scalp, the mortal indignity beneath sacred office. Then they placed the veil over Va-Elun. It was too large for her. It fell around her shoulders, covering the pilgrim’s mantle and scars. When she stood, she seemed at first to disappear inside it. Then she spoke. Her voice was stronger than the old Keeper’s but quieter than Kailan expected. She did not perform power. She let the Hall come toward her.\par

“The Mountain keeps what the open world destroys,” she said. “The Oasis remains because it is hidden. The Temple remains because hands are divided. The people remain because memory does not wander.”\par

The Hall answered in a single breath.\par

“Memory does not wander.”\par

She continued, and now the words were from The Books of The Covenant, though not one of the passages usually given to children.\par

\begin{scripture}
“And God said, Fear not the darkness which I have given thee; for the darkness shall be thy veil, and the stone shall be thy wall, and the deep waters shall be thy life.”
\end{scripture}

Kailan knew the line. Everyone knew the line. But when a new Keeper of Faith spoke it with the old Keeper still breathing behind her, the words changed their weight. Darkness was no longer an absence. It was an inheritance.\par

Va-Elun turned to Ka-Mereth. The old Keeper had begun to shake badly. One attendant wiped the corner of her mouth with white cloth.\par

“Mother of Faith,” Va-Elun said, though no blood joined them, “your charge is complete.”\par

Ka-Mereth’s lips moved. No sound came. The attendant bent close, listened, and spoke for her.\par

“She goes into prayer.”\par

The Hall lowered its heads. The old Keeper was carried away not through the central path but through the Faith arch, toward chambers Kailan had never seen. She would not return to public office. Perhaps she would live months. Perhaps years. Faith had rooms for those whose authority had outlived their strength: quiet cells, prayer benches, warm walls, attendants trained to wash bodies that had once commanded thousands. Children were told such retirement was holy.\par

Kailan thought it looked like being sealed slowly while still alive. He hated himself for thinking it.\par

Then the ceremony changed.\par

It did not end. The great Ascension had been completed, but the Kahir remained in place, and the Hall did not release its breath. The five bowls were still on the stones. The bronze grille behind Balance still whispered with water. From the Shield arch came the faint scrape of a boot being repositioned. In the Justice alcove, Ka-Orun turned one small tablet face-down and opened another.\par

Raedin placed his hand on Kailan’s shoulder.\par

“Now,” he said.\par

Kailan’s mouth went dry.\par

He had known. Of course he had known. His mother had washed his robe herself that morning though two servants from Light had come to do it. His father, who worked among the stone registers and rarely wasted emotion, had held Kailan’s wrist before leaving their chamber and said only:\par

“Stand straight when they speak you.”\par

His younger cousins had looked at him as if he had already become part of a story they were bored of hearing but proud to repeat. Knowing had not prepared him. The names of the young were not the ceremony’s heart. They were a side rite, attached to greater continuity, as young growth attached itself to old root. That made it worse. A private elevation might have belonged to him. This belonged to the Mountain.\par

Tavan stood before the Justice alcove without trembling. Kailan disliked him at once for that.\par

Ka-Orun of Justice read from the tablet.\par

“Tavan, child of the Mountain, called to the hearing of consequence, witnessed by the five, do you accept the discipline of Justice?”\par

“I accept.”\par

“Will you stand between anger and blood?”\par

“I will stand.”\par

“Will you record what others wish forgotten?”\par

“I will record.”\par

“Will you keep the order of events even when the order condemns the beloved?”\par

“I will keep it.”\par

“Will you hear before sentence?”\par

“I will hear.”\par

“Will you let the record change your judgment?”\par

Tavan paused.\par

It was a small pause, but Kailan noticed it and liked him slightly better.\par

“I will try.”\par

A faint movement went through the Justice alcove. Approval, perhaps. Justice distrusted certainty in the young.\par

Ka-Orun touched the black cord at his own wrist and then tied a narrower cord around Tavan’s.\par

“Then rise now as Sa-Tavan, Candidate of Justice.”\par

The Hall seemed larger when Kailan descended into it.\par

The faces blurred into tiers: workers with damp hair from the lower steam rooms, Balance clerks, Shield wardens, Faith singers, children held on hips, old men leaning on polished canes, Light attendants in pale robes. He thought suddenly of all the things they could see. Too thin, perhaps. Too young. Hair refusing to lie flat on the left side. Sandals dusty at the edge. Face too eager if he did not control it.\par

He passed below the Shield arch. Sheva was there. Close now. She did not smile. It would have been improper.\par

She would never desire a boy she had watched pass through awkwardness. The knowledge was not new, but standing beneath her in the Hall made it formal. She was a Veil now, or near enough in bearing that the distinction no longer mattered to him. A woman shaped by duty.\par

He was a Candidate about to be named.\par

He looked away before his face betrayed him. Va-Raedin led him to the eastern stone. Ka-Syphiron, Keeper of Light, watched him approach. The old man’s eyes were pale and unpleasantly clear.\par

“Kailan,” he said.\par

Kailan knelt. The stone was warm under his knees. Too warm. Someone had opened the vent beneath the eastern dais more than necessary. His palms began to sweat. Ka-Syphiron held the same blue-rooted lamp used during the Faith rite.\par

“Do you seek Light?”\par

“I seek Light.”\par

“Why?”\par

The question was not part of the children’s rehearsal.\par

Kailan felt Va-Raedin behind him. Felt the Hall listening. Felt Va-Sheva somewhere to his left, older and impossible. Felt his parents in the galleries. Felt Sa-Tavan of Justice wearing his new cord and probably standing correctly.\par

Because I am worthy, he thought.\par

No.\par

Because I want to know.\par

No. Too naked.\par

Because darkness is not obedience.\par

Absolutely not.\par

His mouth opened before he had fully chosen.\par

“Because what is hidden still has shape,” he said. “And if it has shape, it may be found.”\par

The Hall remained silent.\par

He knew at once that he had said something dangerous. Ka-Syphiron’s expression did not change, but Raedin’s hand, resting lightly behind Kailan’s shoulder, withdrew. Not in disapproval. In attention. The Keeper of Light lowered the lamp until the flame sat between Kailan’s eyes.\par

“A Candidate does not find all that has shape.”\par

“No, Keeper.”\par

“What does a Candidate learn?”\par

Kailan swallowed in relief. The rehearsed script was back in motion.\par

“Where light falls.”\par

“And what does a Veil learn?”\par

“Who placed it.”\par

“And what does a Keeper learn?”\par

Here he was meant to answer: what must remain dark. The words were familiar. Too familiar. They had been repeated in instruction halls, correction cells, whispered contests between boys who wanted to sound older than they were. Kailan almost said them automatically. But after the Faith ceremony, after Va-Elun’s pilgrimage, after the sight of Ka-Mereth carried away into sacred decline, the answer felt less like doctrine than warning.\par

“What must remain dark,” he said.\par

Ka-Syphiron studied him for another moment. Then he lifted his left hand. Raedin passed him a small blade.\par

It was not ornate. Light disliked unnecessary ornament where sharpness mattered. The handle was bone, the edge bright.\par

Kailan extended his hand. The cut was quick across the base of his thumb. Pain came first as surprise, then as heat. Blood gathered, dark in the lamplight. Ka-Syphiron turned Kailan’s hand and pressed the bleeding thumb against a thin plate of polished stone. The stone had been marked with five faint circles, but Kailan’s blood touched only the eastern one.\par

“Blood enters service,” the Keeper said. “Sight enters discipline. Pride enters correction.”\par

Kailan repeated, because repetition made fear useful.\par

“Blood enters service. Sight enters discipline. Pride enters correction.”\par

Ka-Syphiron gave the blade back to Raedin.\par

“Rise now as Sa-Kailan, Candidate of Light.”\par

Kailan rose.\par

The Hall did not cheer. Cheering belonged to stories of Outsiders and their kings. The Mountain acknowledged. That was different. The people struck two fingers against stone, once, twice, five times. The sound filled the hall like rain made hard. Kailan stood within it, his thumb throbbing, his robe suddenly too light for his body. He had thought he would feel transformed.\par

Raedin leaned close enough that only he could hear.\par

“You answered poorly,” he said.\par

Kailan’s stomach dropped.\par

Then Raedin added, “But not uninterestingly.”\par

That, somehow, was worse.\par

The final portion of the rite belonged to Faith again. Va-Elun, raised from Veil of Faith to Ka-Elun, stood before the five bowls and spoke a closing prayer. No one had the strength left for that. Heat had gathered beneath the vaults; a child had fallen asleep against his father’s neck; the old in the lower benches looked grey with fatigue. Even ceremony had to obey bodies eventually.\par

\begin{scripture}
“God who hid the waters beneath the Mountain,” Ka-Elun said, “keep us faithful in measure. God who heard the first blood cry from the dust, keep us faithful in consequence. God who set the desert before the stranger, keep us faithful at the threshold. God who gave the Gospel beneath stone, keep us faithful in memory. God who divided light from darkness, keep us faithful in sight.” She paused.
\end{scripture}

\begin{scripture}
“Let no one hand hold the whole.”
\end{scripture}

This time the Hall answered.\par

“Let no one hand hold the whole.”\par

Kailan said it with them. His thumb pulsed where blood had dried. Let no one hand hold the whole. He wondered whether Faith had found truth there, or only proof that suffering could become authority if witnessed afterward by the right people.\par

Then he looked toward Va-Sheva and wondered what kind of chamber Shield kept.\par

She was watching the western exit now, not him. Duty had reclaimed her. Two wardens had approached quietly, and whatever one of them said made her expression tighten. Not fear. Readiness. She touched the hilt at her side and nodded once. Sa-Kailan noticed because he was of Light now, or because he had always noticed her.\par

The ceremony dissolved slowly. Keepers withdrew through their arches. Families waited for their tiers to be released. Faith attendants lifted the bowls, careful again not to spill. Balance attendants checked the water levels in the basin before removing anything. Justice sealed one tablet before opening the next. Shield released the exits one gallery at a time.\par

The old Keeper’s chair remained empty on the dais for several minutes before someone came to take it away.\par

Va-Raedin bound Sa-Kailan’s thumb with a strip of pale cloth.\par

“You will report to the eastern practice hall before first work,” he said. “Do not eat heavily. Do not arrive proud.”\par

“I am not proud.”\par

Va-Raedin tied the cloth too tightly.\par

“You are fifteen. Pride is your natural condition.”\par

Kailan wanted to object, then saw no useful path through the objection and kept silent.\par

“Good,” Va-Raedin said. “You have learned one thing already.”\par

Across the Hall, Sheva moved with the Shield wardens toward the northern passage. Sa-Kailan’s gaze followed before he could discipline it. Briefly they looked at one another across the dispersing crowd. She smiled then. Not as a woman smiles at a man. He knew that, even in the foolish centre of himself that wanted to deny it. She smiled as one who remembered a boy and approved of his surviving a difficult morning.\par

It should have been enough.\par

It was not.\par

Then she was gone into the Shield passage, and the arch swallowed her without ceremony.\par

Sa-Kailan stood with his bandaged thumb at his side and the new name still settling around him. Sa-Kailan, Candidate of Light. Beneath the hurt of Va-Sheva’s smile and the sting of Va-Raedin’s correction, pride rose bright and undeniable. He had been chosen. Spoken. Cut. Seen. But the day had given him more than a name.\par

Faith had walked until the body nearly failed. Balance had measured water before it became blessing. Shield had watched the exits while others watched the altar. Justice had written the order of events before memory could improve it. Light, he thought, must do something cleaner.\par

But even as he thought it, he remembered Va-Raedin’s voice.\par

Do not admire the flame. Ask where the air enters.\par

The Hall emptied around him. Lamps dimmed one by one as attendants pinched the wicks. The gravure above the dais darkened, and the carved Mountain lost its edges. Briefly Sa-Kailan could no longer see the Oasis in the stone, only the shadow where it had been. He looked at that shadow until his eyes began to invent shapes inside it.\par

Then he turned toward the eastern passage and followed Va-Raedin into Light.\par

\ornament