
\typesetchapter{Chapter 8}{Chain}


The room was small and smelled of oil, dust, metal, old paper, and mistakes.
It was a workroom, low-ceilinged and warm, with walls blackened by lamp smoke and shelves crowded so heavily that they appeared to have survived by mutual pressure. 
There were lenses wrapped in cloth, brass tubes, old diagrams, trays of powdered glass, a cracked bowl, lengths of wire, mirror shards, sealed jars, folded papers, two blackened spoons, 
a small hand-lamp whose flame was turned downward into water, and an object that resembled the skeleton of a bird made by someone who had never seen a bird but had heard disturbing rumours.\par

Va-Raedin stood near the table. 
Ka-Syphiron sat behind it, not in a formal chair but on a low stool with one uneven leg. 
He was wearing no ceremonial outer robe. 
His sleeves were rolled to the wrist. 
A dark mark crossed one finger, either soot or ink or blood long dried. He looked less like one of the five Keepers of the Mountain than like a man who had been interrupted in the middle of ruining something interesting.
Yet when he looked at Sa-Kailan, the room changed.\par

“Close the door,” he said and Sa-Kailan obeyed and the latch fell into place with a small sound.\par

For several breaths nobody spoke.
The silence did not resemble the silences of training, where Va-Raedin let an error ripen until the boy who had made it could smell his own stupidity. 
This silence had witnesses inside it. 
It waited not for an answer, but for sequence.
Ka-Syphiron placed both hands flat on the table.\par

“Sa-Kailan of Light, yesterday, by witness of Va-Raedin, you found the Veil’s Chamber.”\par
“Yes, Keeper.”\par
“You entered by lesson, error, discipline, insight, and disobedience.”\par
“Yes.”\par
“You did not break the path behind you.”\par
“No.”\par
“Do you understand the interval in which you stand?”\par
“Not fully.”\par
“You are Veil in fact,” the old Keeper said. “You are not yet Veil in office. The chamber has recognized you before the Mountain has spoken you. 
That is an uncomfortable place. Many dangerous things begin in the interval between fact and ceremony. But Discovery is not possession" \par
"A Candidate" he continued "learns where light falls, and a Veil ?"\par
"A Veil learns who placed it"\par
"That is the public version, intelligible to all. A Veil learns who placed him. 
And tonight you should see that Va-Raedin placed you there as surely as any lamp: in every correction, every prohibition, every experiment he refused you and every error he let you survive." \par
“Tell me,” Ka-Syphiron continued. “By what condition does a Candidate of Light become Veil of Light?”\par
“By finding the Veil’s Chamber.”\par
“And did you find it?”\par
“Yes.”\par
“By what condition does a Veil of Light become Keeper of Light?”\par
Sa-Kailan looked at Va-Raedin but no rescue came.\par
“By finding the Keeper’s Chamber.”\par
“Incomplete. A Veil who finds the Keeper’s Chamber has reached the upper path. He has not completed the chain. Light does not pass upward because one man climbs. 
It passes when the lower chamber is opened again by one he trained. The Veil must find the Keeper’s Chamber, and a Candidate beneath him must find the Veil’s Chamber. Only then is the chain complete.”\par
Sa-Kailan understood slowly and turned toward Va-Raedin.\par
“You had found it.”\par
“Yes.”\par
“When?”\par
“Six years ago.”\par
The number entered him like cold water.\par
“Six years?”\par
“Yes.”\par
“And you waited.”\par
“I served.”\par
“For me?”\par
“For the chain.” The answer was exact enough to hurt.\par

Sa-Kailan looked down. 
He had spent one day waiting and found it almost intolerable. 
Va-Raedin had spent six years ready, not speaking readiness aloud, not claiming what the Mountain had not yet permitted him to hold. 
Six years correcting boys. Six years refusing impatience the dignity of calling itself destiny.
Siz years waiting for a candidate worthy of becomming Veil.\par


Ka-Syphiron watched both of them.\par
“Va-Raedin of Light found what he was required to find. 
He did not become Keeper because the lower proof had not been renewed. He trained, withheld, corrected. 
He endured the indignity of being ready before the Mountain was ready to receive him.”\par
"You see", he continued, "Faith not long ago had 2 Veils, but none had found the chamber of the last prayer until recently. 
When Va-Elun did, she immediately ascended to keeper because there was another Veil of Faith already in place."
Va-Raedin said nothing. The stillness on his face was not modesty. It was discipline so old it had become shape. 


Then the old Keeper stood.
That, more than the smile, startled Sa-Kailan. Ka-Syphiron did not rise ceremonially. 
He simply came around the table, the uneven stool complaining behind him, and placed one hand on Sa-Kailan’s shoulder.
The hand was warm, dry, and surprisingly heavy. Sa-Kailan froze.\par


“Congratulations, Sa-Kailan.”\par
The words were no longer part of a test. They were not sequence, not record, not condition. They were human.\par
“You have done a difficult thing,” Ka-Syphiron said, “without making it unbearable to praise you. That combination is rarer than genius.”\par
Sa-Kailan tried to answer but nothing came except the start of an incontrollable smile. Ka-Syphiron squeezed once, almost too briefly to be called an embrace.\par
“Do not rush to fill the silence. You have reached one of the few moments in life where silence improves your appearance.”\par
Va-Raedin made a sound that might, in another man, have been laughter. Ka-Syphiron turned toward him.\par
“And you. Six years.”\par
“There was work.”\par
“There is always work. That is how unremarkable men hide from glory.”\par


Ka-Syphiron placed his other hand on Va-Raedin’s shoulder.
Now the three of them stood linked beside the table: the Keeper who was about to be released, the Veil who had waited, the Candidate who had opened the lower door. 
Nothing in the Mountain’s visible order had yet changed. 
No robe had been exchanged, no office named aloud, no Hall warmed, no scripture read. 
Yet the chain had already formed in that cramped blackened room, among broken lenses and hidden bottles and a bird skeleton built by misunderstanding.\par
Sa-Kailan felt the weight of both men around him more than he understood the words that had made the moment possible.\par
“Congratulations, Va-Raedin.”\par
“Keeper.”\par
“No. Not Keeper. Not tonight. Tonight, we celebrate.”\par
Va-Raedin looked up again. For one absurd breath Sa-Kailan thought his teacher might refuse praise as an imprecision.\par
Instead Va-Raedin bowed his head.\par
“Thank you.”\par
The words were quiet enough that the room had to earn them.\par
Ka-Syphiron released them both and stepped back.\par
“There,” he said. “The chain has now been congratulated in all its necessary links. If Justice complains tomorrow that joy was not recorded, we shall say the failure was theirs.”\par
Sa-Kailan laughed once. It came out broken.\par
Va-Raedin looked at him.\par
“Congratulations,” he said.\par
The word had not been softened. Va-Raedin did not know how to soften things without deforming them. Yet the warmth in it was unmistakable because it had not been offered by accident.\par
“You opened the chamber,” Va-Raedin said. “You did not solve every lesson inside it. You did not understand half of what you saw. You probably misnamed the other half before sleeping.”\par
“Yes, Veil.”\par
“But you opened it. Bravo.”\par
For Va-Raedin, that was an embrace.\par
Sa-Kailan bowed his head the same way he saw Va-Raedin bow to Ka-Syphiron.\par
“Thank you.”\par
“For congratulating you?”\par
“For making me capable of receiving it.”\par
“Do not become poetic merely because you have been promoted in advance.”\par
Ka-Syphiron laughed once, bright and delighted.\par
“Oh, leave him one sentence, Va-Raedin. He earned at least one.”\par


Ka-Syphiron clapped his hands suddenly.\par
“Enough.”\par
The solemnity fled like a startled bird.\par
“Absolutely enough. If Light spends every important evening proving it knows how to stand still, then Shield has already won civilization.”\par
Ka-Syphiron crossed to a wall panel so blackened by smoke that it seemed merely dirty until his fingers found three hidden catches.
The panel opened with the reluctant sound of secrecy being asked to perform after years of disuse. Behind it was not a shelf but a hoard.
There were two dark bottles wrapped in oiled cloth, a smaller bottle stopped with wax, a stack of small white cups in a wooden tray, 
three parcels tied with string, two flat boxes, a bundle of dried fruit, a cloth of nuts, a round cheese, bread, strips of salted meat, 
small cakes pressed with seeds. \par

Ka-Syphiron pointed at Sa-Kailan with a soot-marked finger.\par
“Clear the table.” then turned to Va-Raedin "Ray, those, there". Ray, Sa-Kailan repeated in his head, did the partner
"Keeper" Va-Raedin responded
“Do not Keeper me. I am arranging a feast.”
“You cannot be serious.” Va-Raedin opposed
“I have never been more serious.”\par
“You are impossible.”\par
“An excellent quality in a Keeper approaching replacement.”\par

Sa-Kailan looked at the spread as it grew. 
Bread dark at the crust. 
Cheese wrapped in cloth. 
Dried figs, dates, apricots, nuts, salted meat, little seed cakes, preserved fruit, honey, the two bottles, the small wax-stopped bottle, and finally the tray of white cups.
It was absurd.
It was magnificent.
It was more food than Sa-Kailan had ever seen placed before ten people without a festival, a funeral, or an inventory mistake.\par


Ka-Syphiron uncorked the first bottle and placed a white cup in front of Sa-Kailan.
The smell of hydromel rose warm and sweet into the room. Va-Raedin’s expression tightened.\par
“He is fifteen.”\par
“An age at which a man may discover that honey, water, time, and disobedience can produce civilization.”\par
“He is fifteen,” Va-Raedin repeated, looking now not at the bottle but at the tray of cups.\par
Ka-Syphiron followed his gaze.\par
“Ah.”\par
“No.”\par
“Yes.”\par
“Absolutely not.”\par
“Absolutely yes. These are for your first light lesson as Veil, Sa-Kailan.”\par
Sa-Kailan looked from one man to the other.\par
“What is wrong with the cups?”\par
“We are not doing this.”\par
Ka-Syphiron leaned toward him.\par
“Ray, tonight you are being congratulated. Try not to make the room petition against joy.”\par
Sa-Kailan could not help himself.\par
“May I see?”\par


Both men paused to look at him. Va-Raedin’s face said no with the force of a sealed corridor.
Ka-Syphiron’s face brightened with exhilaration and impatience. He lifted one of the small cups and set it before Sa-Kailan.
It was not glass after all, but fine bone or something like it: white, thick-sided, faintly blue at the foot, worn by many hands. 
At the bottom, beneath the glaze, a cloudy glass waited like a blind eye.\par
Sa-Kailan took it in his hands before his courage could discover a more careful course. 
It was heavier than it looked. He tilted it toward the lamp, and the dull pale circle beneath the glaze stirred into a smear of colour, too faint to name, then vanished in the shine.\par


“A flaw,” he said. “Or a mark beneath the glaze perhaps.”\par
Ka-Syphiron looked at Va-Raedin.\par
“You trained this?”\par
“Under difficult conditions.”\par
“I see.”\par
Sa-Kailan turned the cup again, more carefully this time.\par
“Thick wall. Curved base. The glaze distorts whatever has been placed beneath it. Empty, it does not resolve.”\par
“Better.”\par
Sa-Kailan had learned to distrust that word.\par
Ka-Syphiron lifted the bottle.\par
“Look again.”\par
“I am looking.”\par
“And now?”\par


The hydromel poured. It climbed the white inner wall, trembled, and settled over the rounded base.\par
A figure appeared.
Sa-Kailan understood the optics a breath before he understood the subject.
Then he understood both.
The smear beneath the glaze gathered itself through liquid and curvature into a naked woman reclining at the bottom of the cup, one arm behind her head, the other arranged with a confidence that no woman of the Mountain would ever have permitted to be drawn, seen, owned, hidden in a wall, or used as instructional material before a newly recognized Veil.\par
“Oh,” Sa-Kailan said.\par
"Oh !" Sa-Kailan added.
His face burned so quickly that he felt betrayed by his own blood. 
He looked up, then down again, then too far to the left, as if there were a correct place in the room for a boy to put his eyes after finding a naked woman at the bottom of a lesson.
There was not.\par
Ka-Syphiron struck the table with his palm.\par
“There.”\par

Va-Raedin laughed. Not approved. Not breathed through his nose. Laughed. 
That made everything much better and much worse all at once.
Sa-Kailan stared at the cup again, horrified by the fact that he wanted to stare at the cup again, and that the room had broken open around him.
He accepted for a while to be the target of a good joke, then pulled himself together. \par

"Inverse refraction, instead of scrambling a source, we are rebuilding it" \par
“Spectacularl is it not ?” Ka-Syphiron answered. \par
“And old.” Va-Raedin added. \par
“Older than several current moral prohibitions and at least two failed reforms of candidate discipline.” Ka-Syphiron raised his glass.
"To Light,  and all its mysteries." \par

Sa-Kailan lifted the cup. Empty light returned as the liquid shifted, and the woman blurred back toward colour and porcelain. He lowered it. She reappeared. He tilted it left and she lengthened; right, and she dissolved into a pale curve; steady, and there she was again, waiting in the bottom of the cup as if she had always been there and the world had merely lacked the proper drink.\par
“Who made this?”\par
“An important question,” Ka-Syphiron said. “Made the cup? Made the image? Made the principle? Made the hiding place? Light is full of disputed authorship.”\par
“Who kept it?”\par
Ka-Syphiron raised one finger.\par
“That is the better question.”\par
Sa-Kailan looked at Va-Raedin.\par
“Do not look at me,” Va-Raedin said. “I inherited the scandal.”\par
“And preserved it,” Ka-Syphiron said.\par
“For study.”\par
“For continuity.”\par
“For study.”\par
Ka-Syphiron filled the other cups before the distinction could be defended any further.\par
“To Light,” he said, “which has, across generations, bravely preserved even the least necessary evidence.”\par
“To Light,” Sa-Kailan said, still too red in the face to make the words dignified.\par
Va-Raedin looked at his cup, looked at Ka-Syphiron, and surrendered with the expression of a man watching civilization make a written mistake.\par
“To Light.”\par
They drank.\par


The hydromel was sweet first, then hot, then sharper than Sa-Kailan expected. 
It entered his empty places too quickly. He tried not to cough and failed. 
That produced another round of laughter, and this time even Va-Raedin did not retreat from it quickly enough.
Only then did the feast truly begin.\par
Ka-Syphiron broke the bread with both hands and gave the first piece to Va-Raedin.\par
“To six years.”\par
Va-Raedin closed his eyes.\par
“No.”\par
“To six years,” Sa-Kailan said, and by the time he realized he had joined the Keeper against his teacher, it was too late.\par
Va-Raedin looked at him.\par
“You are learning dangerous habits early.”\par
“Good,” Ka-Syphiron said. “A Veil of Light must learn to disappoint his teacher without becoming useless.”\par

They ate. That alone felt improper. 
Not the food, but the abundance of it. 
Sa-Kailan took bread, then cheese, then one dried fig because restraint still seemed, to some foolish part of him, like something he ought to display. 
Ka-Syphiron saw the gesture and placed three more figs on his plate.\par

“This is madness,” \par
Ka-Syphiron’s face lit.\par
“Yes.”\par
Va-Raedin said, “That is not a recognized category of feast.”\par
“It should be,” Ka-Syphiron said. 
“A crazy feast is one at which Balance has not been invited, Justice has not been informed, Shield has not inspected the door, Faith has not decided whether the songs are appropriate, and Light has hidden the cups.”\par
“Then it is not a feast. It is institutional collapse.”\par
“Only if discovered.”\par


Ka-Syphiron poured again, too soon. Va-Raedin corrected the arithmetic. Ka-Syphiron ignored the correction on theological grounds. 
Sa-Kailan ate more than he intended, drank less than Ka-Syphiron wanted, and discovered that laughter could be as disorienting as fear when it arrived in a room where one had expected only judgment.
The evening did not settle after that. It loosened.
Bread broke. 
Cheese softened near the lamp. 
Nuts moved from one side of the table to the other. 
The hidden woman appeared, vanished, reappeared, and became less shocking only because everything else had become impossible as well. 
Ka-Syphiron told three versions of the same story about a Light attendant who had once tried to classify an obscene cup as a devotional vessel. 
Va-Raedin corrected two of them and was accused of vandalizing joy. 
Sa-Kailan laughed until his stomach hurt and then felt guilty for laughing, which only made Ka-Syphiron laugh harder.\par

“To Va-Raedin,” Ka-Syphiron said suddenly.\par
Sa-Kailan raised his cup at once.\par
Ka-Syphiron looked at the man beside him and softened, almost despite himself.\par
“Yes, Ray,” he said. “Tonight you receive praise.”\par
Va-Raedin held his gaze for a moment, then lifted his cup.\par
“Yes,” he said.\par
“To readiness,” Ka-Syphiron said. “Properly humiliated by time.”\par
They drank.\par

Ka-Syphiron reached for the cheese and, with a knife far too ceremonial for the task, cut three uneven portions.\par
“You understand this begins now?”\par
“The office?” Sa-Kailan asked.\par
“The partnership.”\par
Va-Raedin’s expression became carefully blank.\par
“Do not inflate him.”\par
“I am not inflating him. I am informing him where the air will be required.”\par
Ka-Syphiron pointed around the room, then beyond it, as if the walls were no more than an inconvenience of architecture.\par
“The eastern halls do not administer themselves. Lamps do not refill because Light has a beautiful opinion of illumination. 
Candidate sequences do not arrange themselves into wisdom. 
Mirror stores, lens stores, failed apparatus, dangerous apparatus, apparatus whose danger is hidden by its success, cross-Kahir requests, ceremonial obligations, inspection reports, corridor permissions, old prohibitions no one understands and newer permissions no one deserves — all of that will begin to touch you.”\par
Sa-Kailan swallowed.\par
“Me?”\par
“Not as child. Not as ornament. Not as little genius sent to fetch praise from rooms.”\par
Va-Raedin said, “He will remain junior.”\par
“Obviously,” Ka-Syphiron said. “I am describing partnership, not equality. Equality is what young men call partnership when they wish to skip the unpleasant portion.”\par
Sa-Kailan sat straighter despite himself.\par
Ka-Syphiron saw it and smiled.\par
“You will still be corrected.”\par
“Yes, Keeper.”\par
“Relentlessly.”\par
“Yes.”\par
“Sometimes correctly.”\par
Va-Raedin said, “Always correctly.”\par
“There,” Ka-Syphiron said. “You see the scale of the burden.”\par


Sa-Kailan laughed again, and this time Va-Raedin did not immediately punish it with dignity.
The grimness did not vanish.
It moved beneath the table.
That was mercy enough.
After the second Ka-Syphiron opened the second bottle, which Va-Raedin had not approved but had also not prevented, Sa-Kailan found the courage to ask what had been troubling him since the previous day.\par
“Keeper.”\par
Ka-Syphiron narrowed his eyes.\par
“That tone has architecture.”\par
"I have a question."
"Worse."
“It concerns Faith.”\par
“Catastrophic. Continue.”\par
Sa-Kailan looked at the cup, where the woman blurred beneath the hydromel.\par
“Va-Maera, Veil of Faith, said she knew about, what she called 'our little training rooms'.”\par
Ka-Syphiron’s expression did not change.\par
“In the passages. In the remarks. The way they speak of tests, of chambers. Va-Maera frightened me. Not by threatening me. 
By knowing where to press. By speaking as if despair had architecture. As if a prayer could be a tool. As if a voice could enter rooms where hands cannot.”\par
Va-Raedin’s cup paused before it reached the table.\par
“And there are other things,” Sa-Kailan said. “She spoke of archangels and demons. Prayers for plague or despair. 
Words they use as if entire rooms sit behind them. We speak as if Light is the discipline of seeing what others merely believe. 
But Faith has chambers too. Mechanisms too. Hidden paths. Perhaps hidden sciences. 
I do not know how much of what they say is symbol and how much is knowledge.”\par
Ka-Syphiron set down his cup.\par
The room became attentive, but not cold.\par
“Excellent,” he said.\par
Sa-Kailan had expected correction.\par
“Excellent?”\par
“Jealousy would have disappointed me. Fear would have bored me. Unsettlement is a promising beginning.”\par
He turned Sa-Kailan’s cup between thumb and forefinger.\par
“It is one thing to observe refraction. It is another thing entirely to understand why light consents to appear bent.”\par
“Consents?”\par
“All verbs are temporary errors until a better mechanism is found.”\par
Va-Raedin murmured, “He used that defence on Ka-Xian for eleven years.”\par
“And I remain correct.”\par
Ka-Syphiron looked back to Sa-Kailan.\par
“You are pleased to have reached the Veil’s Chamber.”\par
“Yes.”\par
“You should be. If you pretended otherwise, I would distrust either your honesty or your pulse. But look ahead of yourself.”\par
He tapped the cup once.\par
“Decades.”\par
Tap.\par
“Decades of study await you. You have seen patterns in the Veil’s Chamber and mistaken the wound for the body. 
There are lenses you have not learned to name, colours you have not learned to separate, shadows you have not yet earned the right to misread. 
There are paths in Light that Faith can observe only from the outside, if that. There are matters of beyond refraction beyond interference, image, edge, aperture, and delay that no prayer singer in the Mountain has any possible comprehension of.”\par
He smiled slightly.\par
“And there are things only Va-Raedin and I understand.”\par
Va-Raedin’s eyebrow rose.\par
“Some things,” Ka-Syphiron amended. “There are things only I understand because Va-Raedin is stubborn. 
There are things only Va-Raedin understands because he conceals competence under manners. 
There are things neither of us understands, which is why existence remains bearable.”\par
Sa-Kailan looked at the cup.\par
“So Faith has seen effects.”\par
“Of course.”\par
“But not the principles.”\par
“Some principles, perhaps. Never make contempt do the work of thought. Faith is not stupid. Faith is old, as are we.”\par
Ka-Syphiron finished the last of his cup, stood, and wiped his mouth with the back of his hand in a manner that would have caused a formal attendant to faint.\par
“But tonight you are in Light. Which means when anxiety appears, we answer it by fetching objects that should probably have been burned by every moral authority in the Mountain.”\par
Va-Raedin said, “No.”\par
Ka-Syphiron was already moving.\par
“Ray, your objections have been received and placed beneath the nuts.”\par


He disappeared through a side curtain Sa-Kailan had not noticed. The room quieted behind him. 
It happened so quickly that the absence felt physical. 
For several breaths the sounds of the old gallery entered: a faint hiss from a lamp, stone cooling somewhere in the wall, distant water striking a channel below. 
The mountain was asleep.
Sa-Kailan looked at Va-Raedin.
Without Ka-Syphiron’s brilliance filling the room, the Veil seemed more tired than before, but less distant. 
The hydromel had not made him loose. Nothing could. But it had moved something in him one finger’s width from its usual place. \par


“Six years,” Sa-Kailan said.\par
Va-Raedin did not pretend not to know what he meant.\par
“Yes.”\par
“Why did you never tell me?”\par
“Because you would have made it part of your hunger.”\par
Sa-Kailan wanted to object but couldn't.
Va-Raedin took a piece of bread, broke it, and did not eat. \par


“Light is not inherited by announcement. 
If I had told you, you would have tried to become the boy who completed my chain. 
That boy would not have found the chamber. 
He would have searched for praise in the shape of a door.”\par
“I thought you were withholding because you did not think me worthy.”\par
“I was withholding because worthiness is often damaged by being named too early.”\par
That was not comfort.\par
It was more valuable than comfort.\par
“I hated you sometimes,” Sa-Kailan said.\par
“I know.”\par
“I thought you enjoyed correction.”\par
“I do.”\par
Sa-Kailan looked up.\par
Va-Raedin’s face remained perfectly still.\par
Then the corner of his mouth moved.\par
“Not all of it.”\par
Sa-Kailan laughed once, quietly.\par
The silence after that was different.\par
“You did well,” Va-Raedin said.\par
The words were so plain that Sa-Kailan almost missed them.\par
Then they entered him.\par
“I did not understand everything.”\par
“No.”\par
“I still do not.”\par
“Good.”\par
“Good?”\par
“A Veil who thinks the chamber has been solved because it has been opened is unfit to stand near it.”\par
Sa-Kailan nodded.\par
Va-Raedin finally ate the piece of bread.\par
“You are no longer only my student,” he said.\par
The words altered the room more than the formal recognition had.\par
“Not yet my equal. Do not look pleased.”\par
“I was not.”\par
“You were preparing to be.”\par
Sa-Kailan smiled despite himself.\par
“But no longer only my student. That will be difficult for both of us.”\par


Va-Raedin looked toward the side curtain through which Ka-Syphiron had gone.\par
“You will see me speak with others now. Not as your examiner. 
Not only as the man who tells you no. 
You will see me uncertain in places where I cannot afford to appear uncertain. 
You will see old mistakes preserved because no one has yet found a safe way to correct them. 
You will see offices negotiate. You will see Light ask for things and be refused. 
You will see me refuse things I privately admire. It will be unpleasant.”\par
“For me?”\par
“For both of us.”\par


Sa-Kailan thought of Va-Maera. Of Faith. Of chambers hidden inside words.\par
“And when I ask badly?”\par
“I will correct you.”\par
“And when I see something you missed?”\par
Va-Raedin looked at him for a long moment.\par
“Then I hope you will have the discipline to say it before pride makes you polite.”\par
Sa-Kailan felt the sentence enter somewhere deep.\par
Then Va-Raedin’s expression shifted, very slightly.\par
“Though if you ever again say ‘acceptable’ to a servant of Light with that much satisfaction, I may deny this entire conversation occurred.”\par
Sa-Kailan stared at him.\par
“I said that?”\par
“You did.”\par
“To whom?”\par
“To me.”\par
“No.”\par
“Yes.”\par
“I said acceptable to you?”\par
“With the face of a boy who believed he had invented judgment.”\par
Sa-Kailan covered his mouth.\par
The laugh escaped anyway.\par
“No, I did not.”\par
“You did.”\par
“That is terrible.”\par
“It was acceptable.”\par
That destroyed them both.\par
Not loudly. Not enough for dignity to be completely abandoned. But Sa-Kailan bent forward over the table, laughing in horrified disbelief, and Va-Raedin’s mouth lost control in a way that would have terrified the eastern practice halls.\par
“Why did you not strike me?” Sa-Kailan asked.\par
“Because Ka-Dareth would have approved.”\par
“The stool man?”\par
“Among other errors.”\par
Va-Raedin recovered enough to sit back.\par
“Ka-Dareth formalized the present candidate method. Before him, Light chose too many children who wished to be right more than they wished to see. They could memorize. They could repeat. They could polish an answer until no one could bear to touch it. Then they entered a real room and defended their first mistake to the point of spiritual injury.”\par
Sa-Kailan listened.\par
“Ka-Dareth insisted candidates should be selected through recoverable failure. Not humiliation for its own sake. Not cruelty. Trial, error, return. A child who can be wrong without beginning a war against truth may be taught. A child whose ego stands between the eye and the object must be sent elsewhere before he damages a lens or himself.”\par
“So acceptable was his word?”\par
“One of them.”\par
“And I used it on you.”\par
“You did.”\par
“After one chamber.”\par
“Before office.”\par
“Terrible.”\par
“Historic.”\par
Sa-Kailan laughed again.\par
Va-Raedin allowed himself a smaller answer.\par
“Ka-Dareth also invented, perhaps, the Upfall Lamp.”\par
“Perhaps?”\par
“The record is unclear.”\par
“How can the record be unclear in Light?”\par
“Because Light records what was seen, and six people saw six different explanations.”\par
“What is the Upfall Lamp?”\par
Va-Raedin glanced toward the curtain again, as if measuring how much time propriety had left.\par
“A falling thread of water. A lamp broken into pulses by a turning shutter. If the drops fall at one rhythm and the light strikes them at another, the eye accuses them of slowing. Stopping. Sometimes climbing.”\par
“Drops climb?”\par
“No.”\par
“But the eye sees them climb.”\par
“Sometimes.”\par
“That is impossible.”\par
“You have had a busy evening abusing that word.”\par
A crash sounded behind the curtain.\par
Then Ka-Syphiron’s voice.\par
“That is slander. He had a basin as well.”\par
Va-Raedin closed his eyes.\par
“He heard us.”\par
“I am a Keeper of Light,” Ka-Syphiron called. “I hear all conversations that mention contraband, my predecessors, or basins.”\par
He reappeared carrying two objects under one arm and a third clutched between his fingers like evidence recovered from a crime. He looked triumphant.\par
“Also, the record is not unclear. The record is cowardly.”\par
Va-Raedin said, “The record says the authorship is disputed.”\par
“The record says several men with insufficient courage survived Ka-Dareth and wished to attach their names to his better sins.”\par
Ka-Syphiron placed the third object on the table first. It was a small brass shutter, circular, slotted, and blackened near one edge.\par
“The Upfall Lamp was his. The Watchman’s Wheel may have been his. The Returning Rain was unquestionably his, unless one accepts Ka-Meryn’s absurd claim that he had independently discovered how to make dripping water embarrass theology.”\par
Sa-Kailan looked at the shutter.\par
“Did the drops really climb?”\par
Ka-Syphiron leaned close to him.\par
“Of course not.”\par
Then he smiled.\par
“And yes.”\par
Va-Raedin murmured, “This is why Faith dislikes us.”\par
“Faith dislikes us because we keep finding crude ways to demonstrate that mystery has handles.”\par
“And because of the cups.”\par
“The cups are a separate diplomatic burden.”\par
Ka-Syphiron tapped the brass shutter.\par
“Ka-Dareth once carried a leaking goatskin into a disciplinary hearing and made the water climb for six witnesses. Three demanded an inquiry. One began praying. One asked whether the basin could be returned. Ka-Dareth said the Mountain had spent too many generations teaching children to fear error and not enough teaching them to make errors perform useful labour.”\par
“And then?” Sa-Kailan asked.\par
“Then he was forbidden to bring water to hearings.”\par
Va-Raedin said, “For seven years.”\par
“For seven productive years.”\par
Ka-Syphiron pushed the shutter aside and placed the other two objects on the table with theatrical care.\par
The first was a small book bound in worn brown hide. The second was a black cylinder mounted on a wooden base. Narrow vertical slits had been cut around its wall at regular intervals. Inside, Sa-Kailan could see a band of drawings fixed to the inner surface: the same woman repeated again and again, each version changed by a fraction.\par
“Now,” Ka-Syphiron said. “Motion.”\par
He opened the small book.\par
On the first page, a drawn woman stood in a long robe, hair pinned at the back of her head. On the second page, one hand had moved to her collar. On the third, the collar had loosened. On the fourth, the robe had slipped from one shoulder. On the fifth, Sa-Kailan looked up too quickly.\par
Ka-Syphiron’s face was a portrait of innocence.\par
“Is there a problem?”\par
“No.”\par
“Then observe.”\par
He held the pages beneath his thumb and released them quickly.\par
The woman undressed.\par
Not crudely. Not fully enough to be called more than suggestion. But unmistakably, gracefully, scandalously, with all the technical care of a serious Light study and none of the moral seriousness by which the Mountain survived itself.\par
Sa-Kailan stared.\par
Va-Raedin said, “This is historically important.”\par
Sa-Kailan looked at him.\par
Va-Raedin’s face remained solemn for almost one full breath.\par
Then he failed.\par
Ka-Syphiron laughed so hard he had to put the book down.\par
“This,” the Keeper said, tapping the cover, “was invented many, many links above us. The first version was almost certainly not this subject. The second version, alas, discovered civilization.”\par
“Why is it kept?” Sa-Kailan asked.\par
“For study,” Va-Raedin said.\par
“For morale,” Ka-Syphiron said.\par
“For evidence,” Va-Raedin corrected.\par
“For continuity,” Ka-Syphiron replied.\par
“For shame,” Sa-Kailan said.\par
Both men looked at him.\par
Then Ka-Syphiron pointed.\par
“There. He is learning.”\par
He turned the black cylinder on its base.\par
“And this one is from five links above us.”\par
He adjusted a lamp behind it and placed one hand on the handle.\par


“Look through the slits.”\par
Sa-Kailan bent toward the cylinder.\par
At first he saw only interruptions: darkness, image, darkness, image. Ka-Syphiron turned the handle. The drum began to spin. The woman inside shattered into repeated glimpses, each broken by blackness.\par
Then the drum found its rhythm.\par
The woman moved.\par
She lifted her hands to her collar. Cloth loosened. A shoulder appeared. The garment slipped. The motion continued, graceful and scandalous, until the sequence reversed itself and she dressed again, condemned by Light to undress and redress forever inside a spinning cylinder.\par
Sa-Kailan stepped back.\par
“This is obscene.”\par
“Yes,” Ka-Syphiron said. “And mechanically excellent.”\par
Va-Raedin said, “The second quality is the official reason for its preservation.”\par
“The first,” Ka-Syphiron said, “is why nobody misplaced it.”\par
For a moment no one spoke.\par
Va-Raedin looked at Sa-Kailan.\par
“Is he disappointed it is not Va-Sheva?”\par
Sa-Kailan turned violently red.\par
Ka-Syphiron leaned back as if struck.\par
“Oh,” he said. “That was cleanly done.”\par
Sa-Kailan opened his mouth, found no safe word in any Kahir, and closed it again.\par
“He looks as though he is going to be sick,” Ka-Syphiron said, delighted.\par
“Men are all sick in that regard,” Va-Raedin said.\par

The silence that followed was inescapable as if destiny had given no way out.\par
Ka-Syphiron lifted one finger.\par
“Speaking of archangels.”\par
Va-Raedin laughed once, softly.\par
“Ha. Speaking of cleanly done.”\par
“Archangels?” Sa-Kailan said at once, grateful for any path that led away from Va-Sheva.\par
“Yes,” Ka-Syphiron said. “On rare occasions, they come into this world to suspend time.”\par
He tapped the spinning cylinder.\par
“This one was a most enjoyable visit.”\par
Sa-Kailan looked again through the slits.\par

The woman moved because she did not move. Stillness had been arranged into temptation. Drawings had become time. Propriety had been cut into a circle and made reversible by a handle.\par
The inappropriateness was not separate from the device. It announced the room more clearly than any inscription could have done.\par
No Faith chamber would have preserved this. Shield would have locked it away with the person who built it. Balance would have inventoried it under a category too dry to admit what it contained. Justice would have asked who knew, when they knew, and why no one recorded the first offence. But Light had kept it, repaired it, improved it, and on the eve of martyrdom produced it as instruction.\par
Sa-Kailan understood then that the room behind the old gallery was not merely private.\par
It was civilization with its formal face removed.\par
“Light turns stillness into motion,” Ka-Syphiron said. “Or rather, Light studies the conditions under which the eye accuses stillness of having moved.”\par
“It is deception,” Sa-Kailan said.\par
“It is participation.”\par
Va-Raedin added, “Sometimes those are the same accusation made by different offices.”\par
Ka-Syphiron pointed at him.\par
“You see? He is almost ready to be old.”\par
The drum slowed.\par
The woman became jerks, then drawings, then stopped with one hand at her collar, eternally about to begin again.\par
“When our sacred role is filled,” Ka-Syphiron said, “we have time to be irresponsible in ways that later become respectable. A civilisation that never plays only repeats its oldest fears with better posture.”\par
Va-Raedin said, “That sentence will not be used tomorrow.”\par
“No. Tomorrow belongs to people with cleaner sleeves.”\par
Sa-Kailan sat again, still half-smiling, still flushed, still trying to fit embarrassment, awe, and happiness into one body.\par
“I understand almost nothing of Faith,” he said.\par
The admission came out more quietly than he intended.\par
Ka-Syphiron did not mock him.\par
“No,” he said. “You do not.”\par
That was almost kind.\par
Sa-Kailan looked at the cup again.\par
“The prayer for despair. Archangels. Their chambers. Their walking rhythms. The way their voices change stone. I can observe pieces, but the whole remains closed.”\par
“Good,” Ka-Syphiron said.\par
“You keep saying that.”\par
“Because ignorance that knows its own outline is one of the few useful kinds.”\par
He took the cup in which the hidden woman had blurred again beneath the hydromel.\par
“This is Light. A shape is concealed by the conditions of sight. Change the medium, and the eye receives what was already there. Now imagine the same principle, but remove the eye.”\par
He tapped the table once.\par
Sound answered from the walls more richly than the tap deserved.\par
“Echo.”\par
Another tap, slower.\par
“Architecture.”\par
A third.\par
“Rhythm.”\par
A fourth, with his knuckle rather than fingertip.\par
“Music.”\par
The table, the shelves, the small room, even the bottles seemed to answer each tap differently.\par
“Faith listens to stone the way we look through glass. A corridor lengthens a note. A vault divides it. A chamber stores part of it and returns it late. A stair eats one tone and carries another. A rhythm may become an instruction if the wall has been shaped to understand it.”\par
Sa-Kailan listened.\par
“A prayer,” Ka-Syphiron said, “may be words addressed to God. I do not presume to settle that matter tonight. But it may also be a shaped sound moving through stone. A woman sings here. The Mountain answers elsewhere. An echo returns altered. Faith hears the alteration and reverses it, as you reversed the image under liquid. Or they sing something that sounds like an echo in one room, and somewhere else in the Mountain it becomes a song.”\par
Sa-Kailan felt the old discomfort return, but sharpened now by wonder.\par
“So their prayers are mechanisms.”\par
“No.”\par
Ka-Syphiron’s answer was immediate.\par
Sa-Kailan stopped.\par
“That is the foolish version of the clever answer. Do better.”\par
Sa-Kailan thought.\par
“Their mechanisms pray.”\par
Ka-Syphiron’s eyes brightened.\par
“Closer.”\par
Va-Raedin said nothing, but he was watching Sa-Kailan with the same attention he had used in the training rooms when an answer came near enough to be worth silence.\par
“Faith has had centuries. Centuries of bodies in halls, voices under stone, grief given rhythm, doctrine given architecture. Do not reduce them because you have learned a lens. A lens is also ridiculous until it shows you a moon.”\par
Sa-Kailan looked at the blackened walls.\par
“Do you believe in the Archangels?”\par
The question landed more heavily than he intended.\par
Ka-Syphiron leaned back.\par
“Believe is a word Faith polished until everyone else became afraid to touch it.”\par
“Then do you think they exist?”\par
“I do not build rooms for things I cannot see.”\par
“That is not an answer.”\par
“It is an excellent answer. You merely wanted a different one.”\par
Va-Raedin reached for the hydromel.\par
Sa-Kailan turned the thought before speaking it. The drink made courage easier and precision harder. That seemed dangerous.\par
“So you do not believe in the Beast?”\par
The room stilled.\par
Not much.\par
Enough.\par
Ka-Syphiron looked at him for a long moment.\par
Then the old Keeper smiled.\par
“What makes you think I have not seen it?”\par
Sa-Kailan’s skin tightened.\par
Ka-Syphiron lifted his cup.\par
“Remember, Sa-Kailan, I am one of five who keep it beneath the Mountain.”\par
The words had laughter in them.\par
They also had no laughter in them at all.\par
Sa-Kailan looked to Va-Raedin.\par
Va-Raedin, traitor that he was, drank.\par
Ka-Syphiron’s smile widened.\par
“No office talk tonight.”\par
“But—”\par
“No. Especially not if the question begins with but. But is the hinge on which young men swing themselves into locked rooms.”\par
He pointed at Sa-Kailan.\par
“Do I lie? Do I tease? Do I conceal? Do I instruct by making a boy chase his own certainty until it falls into a hole? You will have time to determine that for yourself.”\par
The Beast remained in the room after the joke ended.\par
Ka-Syphiron did not let it remain long.\par
He turned to Va-Raedin with sudden brightness.\par
“And you, Ray. When you were newly Veil, do you remember what Ka-Xian showed you?”\par
Va-Raedin’s expression altered.\par
For a moment he looked younger. Not young, exactly. Va-Raedin might have been born already withholding judgment. But younger.\par
“Yes.”\par
“What did he show you?”\par
“A wheel turning inside glass.”\par
“By?”\par
“Steam.”\par
Sa-Kailan leaned forward.\par
“Steam?”\par
Va-Raedin nodded.\par
“A sealed vessel. A narrow tube. A wheel so light it looked useless until vapour touched it. I thought the lesson concerned motion.”\par
“It did,” Ka-Syphiron said.\par
“It concerned pressure.”\par
“That too.”\par
“And restraint.”\par
“Naturally.”\par
“And then he made me spend an hour discussing yeast.”\par
Sa-Kailan blinked.\par
“Yeast?”\par
Va-Raedin looked almost embarrassed.\par
“Bread. Fermentation. Warmth. Air. Sugars. The transformation of one substance by a living process too small to see.”\par
Ka-Syphiron looked delighted.\par
“He was insufferable about dough.”\par
“He was correct.”\par
“The worst kind of insufferable.”\par
Sa-Kailan looked between them.\par
“Balance?”\par
“Balance,” Va-Raedin said.\par
Ka-Syphiron’s amusement softened.\par
“Most people take them for counters. Ledgers, jars, grain, water, birth allotments, death registers, goat inventories, the moral tyranny of proper storage.”\par
“They are that,” Va-Raedin said.\par
“Oh, entirely. Let us not rescue them from justice.”\par
Then Ka-Syphiron’s voice changed.\par
“But at their heart? Chemistry. Transformation. Patience applied to matter. Balance knows what heat does to milk, what time does to fruit, what sealed darkness does to grain, what pressure does to vapour, what rot can become before it becomes poison. They understand the world as substances altering under condition.”\par
“Like Light,” Sa-Kailan said.\par
“In their own unbearable way, yes.”\par
Va-Raedin placed his cup down.\par
“Their knowledge is endless. They hide it under necessity.”\par
“Just as we hide ours under illumination,” Ka-Syphiron said.\par
“And Faith under prayer,” Sa-Kailan said.\par
The old Keeper smiled at him.\par
“There is hope.”\par
Then the smile thinned.\par
“There is no harm in being interested in a little cross-Kahir knowledge. No harm in admiration. No harm in analogy. No harm in understanding that the Mountain is larger than the room one office permits you to enter.”\par
The hydromel sat warm in Sa-Kailan’s stomach.\par
“But?”\par
“But,” Ka-Syphiron said, “be very, very careful.”\par
No joke followed.\par
That made the warning larger.\par
“Knowledge divided is not only jealousy. It is structure. It is sometimes cruelty. It is sometimes protection. It is always older than your impatience. If you look across a wall, know first whether the wall keeps you out, keeps something in, or prevents two truths from producing a third neither office can survive.”\par
Sa-Kailan did not fully understand.\par
He understood enough to feel the shape of danger.\par
Ka-Syphiron let the silence hold.\par
Then, exactly when it had become almost too grave, he reached beneath the table and produced a wooden box.\par
“Speaking of dangerous cross-Kahir knowledge,” he said, “I have a Justice game.”\par
Va-Raedin laughed in disbelief.\par
“You cannot be serious.”\par
“I am almost never serious in ways people recognise at the proper time.”\par
Ka-Syphiron opened the box.\par
Cards lay inside. Not the rough counting tablets used by children, but thin painted slips marked in suits and numbers. Six stacked packs, worn at the edges, each one slightly different, as if copied over generations by men who believed duplication was a moral aspiration rather than a physical possibility.\par
Sa-Kailan touched one.\par
“Is this permitted?”\par
“By whom?”\par
“Justice.”\par
“Certainly not.”\par
Va-Raedin said, “That is not true.”\par
Ka-Syphiron considered.\par
“No. It is worse than not true. It is incomplete.”\par
He dealt the first cards face up and began explaining.\par
The game was simple enough to learn and difficult enough to distrust immediately. Each hand sought twenty-one without exceeding it. Numbered cards kept their count. Court cards counted as ten. Aces shifted, which Sa-Kailan found morally suspicious. A natural twenty-one paid more. Certain hands could be doubled if their first count permitted it. Equal results did not rescue the player; the dealer won what would elsewhere have been called a tie.\par
“Dealer wins pushes?” Sa-Kailan said.\par
“Justice dislikes ambiguity.”\par
“Justice invented ambiguity.”\par
“Which is why they know how to charge for it.”\par
They played.\par
At first Sa-Kailan thought the lesson lay in courage. Then in restraint. Then in memory. Then in the humiliating fact that Va-Raedin seemed to lose more slowly by doing almost nothing interesting.\par
Ka-Syphiron dealt with maddening calm.\par
Sa-Kailan won several hands early and felt pleased. The pleasure was punished almost immediately. He doubled once because the number looked strong and lost twice as much. He refused a hit because the number felt fragile and watched Ka-Syphiron turn a weak hand into twenty. He split badly. Then he split well and lost both hands, which seemed to him an argument against the universe.\par
Va-Raedin lasted longer.\par
Not because fortune loved him. Fortune did not appear to know he was present. He folded himself around the rules and bled slowly.\par
Sa-Kailan ruined before half the shoe had gone.\par
“That seems unfair,” he said.\par
“It is a game,” Va-Raedin said.\par
“That does not answer me.”\par
“It answers you more often than games deserve.”\par
Eventually Va-Raedin lost as well.\par
Ka-Syphiron gathered the cards.\par
Sa-Kailan stared at the remaining shoe.\par
“The Mountain wins.”\par
Ka-Syphiron looked up.\par
“Does she?”\par
“She did.”\par
“This time.”\par
“And most times?”\par
“Yes.”\par
“Then why play?”\par
“That is the correct question asked too early,” Ka-Syphiron said.\par
He squared the cards with long fingers.\par
““A Veil of Justice, properly trained, could double his token count during a shoe some of the time,” he said. \par
“A Keeper could play an infinite number of shoes, never ruin, and always accrue.”\par
Sa-Kailan looked at Va-Raedin.\par
Va-Raedin gave him nothing.\par
“That is impossible,” Sa-Kailan said.\par
“It is a useful word, impossible. It saves lazy men the trouble of distinguishing between what cannot happen and what cannot happen to them.”\par
“How?”\par
“First, you must play every hand perfectly. Not bravely. Not cleverly. Perfectly. A hand has a correct action under condition. The condition changes. The action changes. That alone takes time to grasp.”\par
Sa-Kailan looked at the cards.\par
“But that is not enough.”\par
“No. Perfect play delays inevitable ruin. It does not abolish it.”\par
“Then how do they do it?”\par
Ka-Syphiron smiled.\par
There was something in the smile that made Sa-Kailan regret the question.\par
The old Keeper shuffled once, slowly.\par
Sa-Kailan watched the cards vanish into one another.\par
“So how do they do it?” he asked again.\par
“Aha.”\par
Ka-Syphiron stopped shuffling.\par
“You asked twice.”\par
Va-Raedin’s eyes moved toward him.\par
“Do not forget that,” Ka-Syphiron said.\par
Sa-Kailan felt the words mark him.\par
The Keeper placed the shoe between them.\par
“You must not become a Justice man. You have candidates to train, chambers to understand, an office to enter without allowing it to enter your bloodstream too quickly. You will have more than enough mysteries without stealing theirs.”\par
He rested one hand on the cards, but did not yet touch the discard pile.\par
“And you must not do the common young thing of stealing one idea from another Kahir and mistaking it for the whole doctrine. One Justice sentence without the others is a trap. One rule without its counter-rule is worse than ignorance, because it gives error the posture of knowledge. They teach in layers for a reason. Lesson one may be true, and lesson three may tell you why lesson one kills men when applied alone.”\par
Sa-Kailan looked at the cards again.\par
“You mean I may misunderstand the clue.”\par
“I mean you will almost certainly misunderstand it. The danger is not misunderstanding. The danger is enjoying the misunderstanding because it feels complete.”\par
Va-Raedin said, “That is not only a Justice warning.”\par
“No. But Justice suffers most visibly when boys collect fragments and call them law.”\par
He tapped the exposed discard pile.\par
“But I will give you a clue.”\par
Tap.\par
“The revealed cards are the past.”\par
He touched the shoe.\par
“The remainder is the future.”\par
Sa-Kailan looked from one to the other.\par
“If the record of the past is kept perfectly, it alters the decision required for the future. Not by prophecy. Not by hope. By condition. What has already appeared changes what remains possible. Justice knows this so deeply they have made morality from it.”\par
Va-Raedin said quietly, “Generous tonight, Keeper.”\par
Ka-Syphiron did not look at him.\par
“Tonight is unusually shaped.”\par
The cards sat between them like a small, silent tribunal.\par
Sa-Kailan tried to see it. Past in the discard. Future in the shoe. The present as decision between them. Every revealed card narrowing the unseen. Every record a pressure placed upon what had not yet happened.\par
He almost understood.\par
Not enough.\par
Perhaps that was the point.\par
Ka-Syphiron closed the box.\par
“And there. You have now drunk illicit hydromel, discovered an indecent woman at the bottom of a cup, watched another undress by circular argument, accused Faith of trespassing, mistrusted Archangels, interrogated the Beast, admired Balance against your own instincts, lost money you did not have to Justice, and been given a clue too large for your current office.”\par
“I have no office yet.”\par
“Exactly. An excellent time to overburden you.”\par
Va-Raedin looked at the unlit lamp.\par
The evening changed.\par
No object moved. No lamp dimmed. Yet the laughter withdrew to the edges of the room, not banished, not denied, but completed.\par
Ka-Syphiron placed one hand on the white cloth still lying near Sa-Kailan.\par
“Tomorrow,” he said, “the Mountain will speak what is already true.”\par
Sa-Kailan did not answer.\par
The old Keeper’s eyes were bright, steady, unreadable.\par
“You will stand where you are called. You will listen when Justice names conditions. You will not reach for understanding faster than ceremony permits. Let the Hall be larger than your mind for one morning. It will do you good.”\par
“Yes, Keeper.”\par
“And when you are given cloth, lamp, word, silence, or wound, you will receive each as condition, not ornament.”\par
“Yes.”\par
Ka-Syphiron nodded.\par
Then the old mischief returned, but more gently now.\par
“And if you must be terrified, try not to do it with your shoulders. It makes the robe hang badly.”\par
Sa-Kailan laughed once because he could not help it.\par
The old man looked pleased with that too.\par
Then he said, “All right.”\par
The words closed something.\par

“Sa-Kailan, I have a few things to tell Va-Raedin.”\par
There it was.\par
The door inside the door.\par
Sa-Kailan had been brought near. Fed. Praised. Warned. Taught beyond his level. He had been allowed to laugh in the room where tomorrow had already begun. But proximity was not possession. Veil in fact was not yet Veil in office. Some words still passed above him.\par
He rose.\par
For a moment he wanted to say something worthy of the evening. Nothing came that would not diminish it.\par
So he bowed.\par
Not as he had bowed on entering.\par
Lower.\par
When he straightened, Va-Raedin was watching him. The warmth was gone from his face, but not from the room. It had simply taken the form Va-Raedin could bear.\par
“Sleep,” Va-Raedin said.\par
Sa-Kailan almost smiled.\par
“That is an instruction?”\par
“It is a hope. Less reliable.”\par
Ka-Syphiron laughed softly.\par
“Go before he becomes sentimental.”\par
Sa-Kailan turned toward the door.\par
“Wait,” Va-Raedin said.\par
The word stopped him more completely than a command would have done.\par
Va-Raedin reached into the inner fold of his robe and removed a small object wrapped in grey cloth. He held it for a moment before offering it, as if even now he disliked the disorder of gifts.\par
Sa-Kailan took it carefully.\par
Inside was a triangular piece of clear glass, no longer than his thumb, polished on three faces and chipped at one corner. It was old enough to have lost perfection and fine enough to make the loss beautiful. When he turned it toward the lamp, the flame broke faintly along one edge into a thin wound of colour.\par
“A prism,” Sa-Kailan said.\par
“Yes.”\par
“For tomorrow?”\par
“No.”\par
“For the chamber?”\par
“No.”\par
Va-Raedin’s eyes held his.\par
“For after.”\par
Sa-Kailan closed his fingers around it.\par
Va-Raedin said, “When a beam seems whole, do not believe it too quickly.”\par
Ka-Syphiron looked away, but not fast enough to hide that the line pleased him.\par
“I do regret one thing,” the old Keeper said.\par
Sa-Kailan looked at him.\par
“Only one?” Va-Raedin asked.\par
“Tonight, yes. Tomorrow I may be more ambitious.”\par
The smile faded almost at once.\par
“I regret not finding the secret of the memory of light.”\par
Sa-Kailan looked down at the prism in his hand.\par
“The memory of light?”\par
“A poor name for an unpierced mystery.”\par
“What is it?”\par
“Something my predecessor observed. Something neither he, nor I, nor Va-Raedin has opened.”\par
Va-Raedin said, “It may not be a door.”\par
“Everything interesting denies being a door at first.”\par
Ka-Syphiron reached for one of the lamps and turned it a little, so that the prism in Sa-Kailan’s hand caught a thinner edge of flame.\par
“You know how a song may seem one thing until Faith separates the voices inside it,” he said. “One voice, another above it, another below it, all travelling together, so that the ear receives one sound and thinks it has heard simply.”\par
Sa-Kailan nodded slowly.\par
“Light may do the same,” Ka-Syphiron said. “Some lights are a sum. They look single because the eye is merciful, but break them correctly and many lights are travelling together. Others are almost a single thing. A gas may take one light from the world, and when struck correctly give back just that one.”\par
Sa-Kailan frowned.\par
“As if it remembers it?”\par
“As if it is permitted nothing else.”\par
The words stayed in the air longer than the lamp-glow did.\par
Then Ka-Syphiron smiled again, lightly, almost cruelly.\par
“There. A useless gift. The best kind for a young man with too many useful hungers.”\par
Sa-Kailan bowed, smaller this time, because the object in his hand made any larger movement feel false.\par
“Thank you,” he said.\par
Va-Raedin gave the smallest possible nod.\par
“Now sleep.”\par

Sa-Kailan opened the door.\par
The old gallery waited outside, dim and smoke-blackened, full of failed lamps and preserved errors.\par
Behind him, Ka-Syphiron and Va-Raedin did not speak until the door had nearly closed.\par
He heard only one fragment.\par
Ka-Syphiron’s voice, lower than before.\par
“Here, for tomorrow and for ...”\par
Then the latch caught, and the rest belonged to those not yet dead and those not yet raised.\par
Sa-Kailan stood for a moment with his hand still on the door.\par
He carried three mysteries away with him.\par
Faith, which might turn echo into prayer and prayer into signal.\par
Justice, which might read the future by refusing to forget the past.\par
And the thing Ka-Syphiron still had to say to Va-Raedin on the last night before Light changed hands.\par
The corridor had not altered.\par
Everything in it had.\par
The public ceremony began before the Hall filled.\par
That was how it felt later, when he tried to remember sequence. One moment he stood in the smoke-blackened old gallery with his hand still near the closed door, carrying three mysteries away from the room behind it. The next, attendants of Light were dressing him in pale robes he had never worn, plain but heavier than Candidate grey. The folded white cloth was taken from the table where Ka-Syphiron had laid it and bound around his wrist, not tightly, not like a wound, but like a visible condition.\par
He was led by Va-Raedin through corridors that had emptied ahead of them. Shield had already closed side routes. Balance had already marked warmth and lamp oil for the Hall. Justice had already opened record.\par
The Mountain had begun to change around the fact.\par

\ornament