Book 1 - Genesis
In the beginning God created the heaven and the earth. And the earth was without form, and void; and darkness was upon the face of the deep. And the Spirit of God moved upon the face of the waters. And God said, Let there be light: and there was light. And God saw it was good: and God divided the light from the darkness. And God called the light Day, and the darkness he called Night. And the evening and the morning were the first day.
And God said, Let there be a division in the midst of the deep, and let it divide the heights from the depths. And God made the sky above, and the ground beneath, and the pit below the ground: and it was so. And God called the sky Heaven, and the ground Earth, and the pit He called Hell. And God set the Earth between Heaven and Hell, that the living should walk between the eye above and the fire beneath. And the evening and the morning were the second day.
And God said, Let the waters under the heaven be gathered together unto one place, and let the dry land rise above them: and it was so. And God gathered the waters beneath the earth, and hid them in the hollow places of the deep; and the dry land appeared, and from the dry land He raised the Mountain. And God said, Let the waters bring forth a spring within the shadow of the Mountain, and let the spring be hidden from the eye of heaven: and it was so. And the spring came forth in the shadow of the Mountain, and there was grass around it, and trees for shade, and the green herb yielding seed, and the fruit tree yielding fruit after his kind whose seed is in itself, upon the Earth. And God called the spring in the shadow of the Mountain Oasis: and God saw that it was good. And God said, Let there be a forest round about the Mountain and the Oasis, and let rivers run through the forest, and let the trees be many, and let their branches give shade unto the living: and it was so. And the forest grew round about the Mountain, and the rivers ran through it, and the waters thereof gave life to every root and beast and creeping thing. And the evening and the morning were the third day.
And God said, Let there be a desert round about the forest and the rivers, to divide the forest from the lands beyond, and to set thirst and wandering before the stranger: and it was so. And God called the desert the Threshold; and He set heat upon it by day, and cold upon it by night, and mirage upon its face, that no man should cross it lightly. And God said unto the living, The Mountain is refuge, and the Oasis is water, and the forest is plenty, and the rivers are life; but beyond the forest thou shalt not wander, for beyond the forest lieth thirst, and the eye of Heaven, and the forgetting of My shadow. And God saw that the hidden place was good. And the evening and the morning were the fourth day.
And God said, Let us make man in our image, after our likeness: and let him walk upon the Earth, between the eye above and the fire beneath. And God formed Adam from the dust of the Mountain, and breathed into him the breath of life; and Adam became a living soul. And God said, It is not good that man should be alone in the shadow of the Mountain: and God caused a deep sleep to fall upon Adam, and he slept; and from his side God made Eve, and brought her unto the Oasis. And God blessed them, and God said unto them, Be fruitful, and multiply, and keep the Oasis, and dwell in the shadow of the Mountain; and eat of the fruit of the forest, and drink of the rivers thereof, but wander not beyond the forest. And Eve bare five children unto Adam; and the first opened his eyes unto the light, and the second drank from the waters, and the third laid his hand upon the stone, and the fourth took up the blade, and the fifth lifted his voice unto Heaven. And God saw Adam, and Eve, and their five children in the Oasis, beneath the shadow of the Mountain; and God saw that it was very good. And the evening and the morning were the fifth day.
Book 2 - Blood
And after the fifth day, God spoke unto Adam and unto Eve, and unto the five children whom He had set in the Oasis beneath the shadow of the Mountain; and God said, Let there be Keepers among you, that the Mountain be not without order, nor the Oasis without care, nor the people without peace. And God called the woman of water, and said unto her, Keep thou the Balance, and gather the fruits, and count the waters, and keep the resources of survival; for what is wasted in peace shall be missing in hunger. And God called the woman of holy speech, and said unto her, Keep thou the Faith, and teach My Gospel unto the children, and keep prayer, breath, rhythm, and memory beneath the Mountain; for I shall provide.
And God called the man of strength, and said unto him, Keep thou the Shield, and stand before the beasts of the Earth, and before the threshold of men; suffer not the children to be devoured, nor danger to enter unnamed. And God called the woman of brightness, and said unto her, Keep thou the Light, and illuminate the Mountain, and teach the eye where to stand, that the people stumble not in darkness. And God called the man of peace, and said unto him, Keep thou Justice, and stand between brother and brother, and keep the record of sequence, that blood answer not blood without measure. And the Keepers received their charges, and each kept what had been given, and the people multiplied in the Oasis.
And the Keepers had sons and daughters, and their children had children after them; and the Oasis was filled with voices, and the Mountain gave shadow, and the waters were counted, and the Gospel was taught, and the beasts were driven back, and the lamps were lit, and peace was kept among the people. And the Keeper of Shield had two sons; and the elder was great in strength, and broad in the shoulder, and mighty before the beasts; and the younger was small in his body, and cunning in his heart. And the younger looked upon his brother, and saw that the people loved his strength, and envy entered him as a serpent entereth the grass.
And the younger spoke softly unto his brother, and smiled with his mouth while murder grew within him; and he rose against his brother in secret, and slew him where no beast had entered. And the first blood of man cried from the dust. And fear came upon the people, for Shield had defended the Oasis from the beasts without, but had not seen the beast within. And the younger fled from the Oasis, and took with him the wife of his brother, a daughter of Faith, and went out from the shadow of the Mountain into the open land.
And God saw that blood had entered the world of men, and that holy speech had been carried away by the murderer; and God cursed them, saying, Thou shalt wander under the eye of Heaven, and the Earth shall not hide thee. Thy children shall thirst for what they cannot keep, and conquer what they cannot love. Thou shalt remember the Mountain, but thou shalt not find it; and thou shalt hear the Gospel, but twist it into hunger. And from the murderer and the stolen daughter came the Outsiders.
Book 3 - Beast
And the curse went forth upon the Outsiders, and they wandered under the eye of Heaven; and the Earth hid them not, neither did the waters keep them, nor the Mountain give them shade. And they multiplied in the open land, and their hunger multiplied with them; and brother rose against brother, and city rose against city, and king rose against king. And they made war without measure, and famine without pity, and prisons without mercy; and they bound the weak, and burned the innocent, and betrayed the hand that fed them.
And God looked upon the Outsiders, and saw the works of their hands; and behold, their cruelty was greater than the curse that had been laid upon them. For they tortured for pleasure, and conquered for pride, and lied with holy words, and slew their children for the hunger of their fathers. And God was wroth, and God said, The evil of the Outsiders is deep, and their hunger hath no end; and if they find the Mountain, they shall not spare the Oasis, nor the Temple, nor the children beneath My shadow.
And God gave unto the Oasis a great defence, and set power in the keepers thereof. Unto the Keeper of Light He gave the Dance of Light, that the hidden ways might be revealed, and the blind stone awakened, and the signs within the holy word made visible to the appointed eye. Unto the Keeper of Balance He gave the Movement of the Waters, that the deep might rise and fall by measure, and that water might become breath, and breath become force. Unto the Keeper of Justice He gave the Key, the Number, and the Binding of the Chains, that no motion begin before its hour, nor any burden move without order and witness. Unto the Keeper of Shield He gave the Sword and the Seal, that the threshold might be defended, and the Mountain closed against the wrath it had awakened. And unto the Keeper of Faith He gave the Prayer and the Listening of Stone, that the holy word might be spoken in rhythm, and the trembling of the Earth understood.
And God said unto the Keepers, Keep these secrets apart, and let no one hand hold the whole; for the power that defendeth the Oasis may also devour the Earth. And He taught each Keeper his portion, and sealed the rest from him. And the Light knew not the full prayer, and the Faith knew not the hidden sight, and the Balance knew not the final number, and the Justice knew not the deep waters, and the Shield knew not the word that opened what it sealed; and each was strong in his charge, and weak without the others.
And God said, If the Outsiders come again, and if they find the Mountain, and if they lay their hands upon the Temple, then shall the Keepers stand as one and unleash the Beast upon the Earth. And the poisoned waters shall rise, and the volcanoes shall erupt, and the ground shall break beneath the feet of the enemy, and the sea shall climb over their cities; and the Outsiders shall taste the cruelty they gave upon the world. But if the Keepers fail, or if one hand seize the whole, then the Beast shall know neither faithful nor stranger, and Hell shall not distinguish between the guilty and the guarded.
Book 4 - Cataclysm
And the Outsiders multiplied in the open land, and their kings became many, and their cruelty became without number; and they remembered the Mountain in dreams, and hated the shade they could not find. And they gathered armies, and forged blades, and raised engines of war, and came with banners and drums and false prayers upon their tongues. And they said, Let us find the Mountain, and take the Oasis, and drink the waters thereof, and break the Temple, and make ourselves gods upon the Earth.
And the Outsiders came unto the desert, and the desert rose against them with thirst and wandering; yet they turned not back. And they drove the weak before them, and cast down the weary, and slew the beasts that bore them, and drank what was forbidden, and ate what was unclean; and the sun burned them by day, and the cold struck them by night, and still they came. And the desert saw their hearts, and the Mountain heard their footsteps, and the Oasis trembled in the shadow.
And the Keepers stood before the Temple, and each brought forth the secret that had been given unto him. And the Keeper of Shield sealed the threshold, and closed the Mountain against the wrath to come. And the Keeper of Justice set the Key, and counted the Number, and took witness of the five, that no hand should move before the others. And the Keeper of Balance moved the deep waters beneath the Earth, and loosed them by measure into the chambers below. And the Keeper of Faith spoke the Prayer in its appointed rhythm, and listened for the answer of stone. And the Keeper of Light danced the Dance of Light, and the hidden signs were revealed, and the way was made visible. And the five stood apart as one, and no one hand held the whole.
And God remembered His word, and Hell answered. And the deep waters fell into the fire beneath, and the breath of Hell rose in the hidden chambers, and the works of the Mountain began to move. And the Beast beneath the Mountain struck the roots of the Earth with his fists, and the ground broke beneath the feet of the Outsiders. And the prayer travelled through stone, and the weak places of the Earth answered; and the poisoned waters rose, and the volcanoes erupted, and fire came upon the left hand of the army, and the sea rose upon the right hand thereof; and the mountains shook, and the sky was made black with ash, and the Outsiders were burned, and drowned, and crushed, and scattered; and they tasted the cruelty which they had given unto the world.
And when the wrath had passed, the armies of the Outsiders were no more; but the forest was wounded, and the forest was broken, and the open waters were spoiled, and the land around the Mountain became desert. And the people wept for the world that had been saved and lost in the same hour. And God said unto them, Dwell no more in the open, but enter the Mountain, and keep the Temple, and remember the price of being seen. And the Temple was sealed, and the people went beneath the stone.
Book 5 - Promise
And the people dwelt beneath the stone, and darkness was above them, and the Mountain covered them; and they remembered the forest, and the rivers, and the open waters that were spoiled. And they wept for the world that had been saved and lost in the same hour. And God heard their mourning beneath the Mountain.
And God said unto them, Fear not the darkness which I have given thee; for the darkness shall be thy veil, and the stone shall be thy wall, and the deep waters shall be thy life. Thou shalt dwell within the Mountain, and thy children shall be born beneath the Mountain, and the eye of Heaven shall not find thee.
And God said, Keep thou the Mountain, and keep thou the Oasis, and keep thou the Temple. Keep the secrets apart, and let no one hand hold the whole. Let no sign be carried from its chamber, nor proof borne beyond its seal; for a secret held in place may preserve the people, but a secret made to wander shall become fire among them. Let the Keeper of Balance count the waters, and the Keeper of Faith teach the Gospel, and the Keeper of Shield defend the threshold, and the Keeper of Light illuminate the passages, and the Keeper of Justice keep peace between brother and brother.
And God said, If thou keepest My word, and if thy children keep it after thee, then I shall remember the forest and the rivers that were lost. And the day shall come when the desert shall withdraw, and the waters shall run openly again, and the trees shall rise where sand and ash had lain.
And in that day the Mountain shall open without fear, and the Beast shall be chained forever in Hell, and the children of the hidden people shall walk beneath Heaven and not be consumed. The people shall be reborn, and the Oasis shall no longer hide in shadow; and the people shall know neither thirst, nor wandering, nor the fear of being seen. And God saw the promise, and behold, it was very good.
