\typeprologue

Clémentine,\par

I cannot ask for your forgiveness without first setting down the chain of events that made history unfold as it did. 
The difficulty is knowing where such a chain begins.\par

The beginning, what a strange word. 
I could begin with myself: a young scientist studying seismic resonance and chain reactions. 
I could begin with the men who abducted me, and later broke me in rooms no court will ever inspect. 
I could begin with the oasis and the mountain, found exactly as they had been described to me.\par

\begin{scripture}
In the beginning God created the heaven and the earth.
\end{scripture}

Starting there would be at least one chapter too far back, and, somehow, not far enough.\par

In 1998, while working for an organization unapologetically called "The Outsiders", I met a young man in Switzerland who said his first name had once been Kailan.\par

What follows is my retranscription of his account. 
I have ordered it as faithfully as I can. 
Some scenes may carry the shape I gave them, some silences may have been filled from what I later learned, and some details may bear the mark of my own memory of the Mountain. 
I may have misunderstood. I may have mistranslated. 
I may have placed emphasis where another witness would not have placed it. 
But the substance is true. The people were real. The rooms were real.
The rites, the wounds, the names, the laws, the fear, the mercy, and the catastrophe happened as written.\par

This is my retranscription of his account, and the first link in my story.\par

Paul Morgan.\par

\ornament